---
title: "Dynamic Structural Equation Models"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{ag_dsem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```


A dynamic structural equation model (dsem) specifies how a set of time-series affect each other using an arrow lag notation, and computes
a joint density over all time-series. Within the context of a stock assessment, the time-series we are primarily focused on are deviation process errors (e.g., recruitment), and how these process errors might be explained by environmental covariates. In `SPoRC`, various population processes can be linked to a dsem model, thereby replacing the process error penalty a given process originally had (e.g., iid recruitment deviations) with a dsem density (which can express iid recruitment deviations and more). In the following, we demonstrate how a dsem can be specified (Thorson et al. 2024) using `SPoRC`, and the various processes to which a dsem can be linked. Specifically, we first demonstrate the dsem model and the mathematical underpinnings of it, fit a dsem within the context of a stock assessment model, conduct a self test and cross test, and illustrate dsem forecasting capabilities. 

```{r setup, warning = FALSE, message = FALSE}
library(SPoRC)
library(RTMB)
```

## What a dsem is

### Arrow and Lag notation

In general, a dsem can be specified using arrow and lag notation (and see Thorson et al. 2024 for further details). Here, we specify one arrow per line, which follows the nomenclature of `from -> to, lag, name, start`.  The time-series `from`, `lag` years earlier, affects the time-series `to`. The coefficient is the parameter `name`, or is fixed at `start` when the name is set to `NA`. This formulation therefore specifies a path coefficient in a dsem model (i.e., analogous to a regression coefficient in a linear model). An sd line can then be specified using a double-headed arrow notation `a <-> a, 0, name, start`, while a covariance between two series' innovations is `a <-> b, 0, name, start`. A name used on two arrows is one shared parameter (i.e., the same name in two lines). An example is provided below where we specify that an environmental covariate in year `x` has an impact on recruitment in year `x`, while the environmental covariate itself follows an AR(1) like process $x_t = \rho x_{t-1} + \varepsilon_t$. Process error for the environmental covariate and recruitment deviations is then specified as well.

```{r arrows}
arrows <- c("env -> rec, 0, b_env, 0.5",     # this year's covariate affects this year's recruitment
            "env -> env, 1, rho, 0.5",       # the covariate follows itself from last year
            "env <-> env, 0, sd_env, 0.5",   # covariate innovation sd
            "rec <-> rec, 0, sd_rec, 0.5")   # recruitment innovation sd

m <- read_dsem_arrows(arrows, variables = c("env", "rec"))
m$arrows[, c("type", "from", "to", "lag", "name")]
```

`read_dsem_arrows` is not used directly by the user but can be a useful way to understand the various processes specified within the dsem as well as where things fit within matrices used when specifying a dsem. 

### Going from arrows to a density function 

Building off of this simplified example, we hope to provide intuition for how a dsem likelihood is computed. In this case, we continue using the `env` time-series and `rec` time-series, while specifying three years and the four arrows written as above. Here, we write the values as a
matrix / grid, years in the rows and series in the column, and number the cells series by series:

```
          env     rec
year 1    1       4
year 2    2       5
year 3    3       6
```

Note that each cell has an innovation (i.e., the part that the arrows themselves do not explain). For example, the recruitment innovation in year 2 is cell 5 minus `b_env` times cell 2, because `env -> rec, 0`:

$$
\begin{aligned}
\varepsilon^{rec}_2 &= rec_2 - b_{env}\,env_2 \\
rec_2 &= b_{env}\,env_2 + \varepsilon^{rec}_2
\end{aligned}
$$

The covariate innovation in year 2 is cell 2 minus `rho` times cell 1, given that `env -> env, 1`:

$$
\begin{aligned}
\varepsilon^{env}_2 &= env_2 - \rho\,env_1 \\
env_2 &= \rho\,env_1 + \varepsilon^{env}_2
\end{aligned}
$$

Note that year 1 has nothing before it so the covariate innovation is just cell 1. We can then collect all those subtractions into a matrix called `B` with one row per cell: e.g., row 5 has `b_env` in column 2, row 2 has `rho` in column 1. Then the vector of innovations can be expressed as:

$$\varepsilon = (I - B)(x - \mu)$$

where

- $x$ is the grid stacked into one vector of length 6, cell order as above
- $\mu$ is each cell's mean, a constant per series (zero for a deviation)
- $B$ is the 6 by 6 matrix of path coefficients, row $k$ holding the
  coefficients on the cells that point into cell $k$
- $I$ is the identity, so $I - B$ has ones on the diagonal and minus each
  coefficient off it
- $\varepsilon$ is the vector of innovations, one per cell

Written out in full for this 3-year, 2-series example, $B$ is (rows are the cell being predicted, columns are the cell doing the predicting):

$$
B =
\begin{pmatrix}
0 & 0 & 0 & 0 & 0 & 0 \\
\rho & 0 & 0 & 0 & 0 & 0 \\
0 & \rho & 0 & 0 & 0 & 0 \\
b_{env} & 0 & 0 & 0 & 0 & 0 \\
0 & b_{env} & 0 & 0 & 0 & 0 \\
0 & 0 & b_{env} & 0 & 0 & 0
\end{pmatrix}
\begin{matrix}
\leftarrow env_1 \\ \leftarrow env_2 \\ \leftarrow env_3 \\ \leftarrow rec_1 \\ \leftarrow rec_2 \\ \leftarrow rec_3
\end{matrix}
$$

Rows 2-3 hold `env -> env, 1, rho` (one step off the diagonal within the `env` block), rows 4-6 hold `env -> rec, 0, b_env` (the diagonal of the lower-left block, since the lag is 0), and row 1 is empty because year 1 has no previous values. Given that, it follows that:

$$
I - B =
\begin{pmatrix}
1 & 0 & 0 & 0 & 0 & 0 \\
-\rho & 1 & 0 & 0 & 0 & 0 \\
0 & -\rho & 1 & 0 & 0 & 0 \\
-b_{env} & 0 & 0 & 1 & 0 & 0 \\
0 & -b_{env} & 0 & 0 & 1 & 0 \\
0 & 0 & -b_{env} & 0 & 0 & 1
\end{pmatrix}
\begin{matrix}
\leftarrow env_1 \\ \leftarrow env_2 \\ \leftarrow env_3 \\ \leftarrow rec_1 \\ \leftarrow rec_2 \\ \leftarrow rec_3
\end{matrix}
$$


Writing the centered values as $\tilde{x} = x - \mu$, and taking $\mu = 0$ as it is for a deviation series, multiplying through recovers every innovation at once:

$$
\varepsilon = (I - B)\,x =
\begin{pmatrix}
1 & 0 & 0 & 0 & 0 & 0 \\
-\rho & 1 & 0 & 0 & 0 & 0 \\
0 & -\rho & 1 & 0 & 0 & 0 \\
-b_{env} & 0 & 0 & 1 & 0 & 0 \\
0 & -b_{env} & 0 & 0 & 1 & 0 \\
0 & 0 & -b_{env} & 0 & 0 & 1
\end{pmatrix}
\begin{pmatrix}
env_1 \\ env_2 \\ env_3 \\ rec_1 \\ rec_2 \\ rec_3
\end{pmatrix}
=
\begin{pmatrix}
env_1 \\
env_2 - \rho\,env_1 \\
env_3 - \rho\,env_2 \\
rec_1 - b_{env}\,env_1 \\
rec_2 - b_{env}\,env_2 \\
rec_3 - b_{env}\,env_3
\end{pmatrix}
=
\begin{pmatrix}
\varepsilon^{env}_1 \\ \varepsilon^{env}_2 \\ \varepsilon^{env}_3 \\ \varepsilon^{rec}_1 \\ \varepsilon^{rec}_2 \\ \varepsilon^{rec}_3
\end{pmatrix}
\begin{matrix}
\leftarrow env_1 \\ \leftarrow env_2 \\ \leftarrow env_3 \\ \leftarrow rec_1 \\ \leftarrow rec_2 \\ \leftarrow rec_3
\end{matrix}
$$

Next, the two `<->` lines then fill a diagonal matrix of innovation standard deviations:

$$
\Gamma = \mathrm{diag}(sd_{env},\, sd_{env},\, sd_{env},\, sd_{rec},\, sd_{rec},\, sd_{rec}), \qquad \varepsilon \sim \mathrm{MVN}(0,\, \Gamma\Gamma^\top)
$$

The two `<->` lines give the variance of each innovation: $\mathrm{Var}(\varepsilon^{env}_t) = sd_{env}^2$ and $\mathrm{Var}(\varepsilon^{rec}_t) = sd_{rec}^2$, with all innovations assumed to be independent. From these, we can work out the variance of the time-series themselves, one cell at a time, using the same equations as above.

In the case of the covariate, year 1 is only its own innovation, so:

$$
\mathrm{Var}(env_1) = sd_{env}^2
$$

Year 2 is $env_2 = \rho\, env_1 + \varepsilon^{env}_2$. The two terms are independent, so their variances add:

$$
\mathrm{Var}(env_2) = \rho^2\,\mathrm{Var}(env_1) + sd_{env}^2 = (1 + \rho^2)\, sd_{env}^2
$$

Lastly, year 3 works the same way:

$$
\mathrm{Var}(env_3) = \rho^2\,\mathrm{Var}(env_2) + sd_{env}^2 = (1 + \rho^2 + \rho^4)\, sd_{env}^2
$$

In words, each year keeps a shrunken copy of last year's variability and adds a new shock. For $|\rho| < 1$, this levels off at $sd_{env}^2 / (1 - \rho^2)$ (i.e., it is stationary).

For recruitment defined as $rec_t = b_{env}\, env_t + \varepsilon^{rec}_t$, the variance can be written as:

$$
\mathrm{Var}(rec_t) = b_{env}^2\,\mathrm{Var}(env_t) + sd_{rec}^2
$$

Recruitment variability is its own shock plus whatever the covariate passes through $b_{env}$. When $b_{env} = 0$, this is just $sd_{rec}^2$ every year, which is the usual iid recruitment deviation.

In addition to the variances, the covariances can also be written out as:

$$
\mathrm{Cov}(env_1, env_2) = \rho\, sd_{env}^2, \qquad
\mathrm{Cov}(env_t, rec_t) = b_{env}\,\mathrm{Var}(env_t), \qquad
\mathrm{Cov}(rec_1, rec_2) = b_{env}^2\,\rho\, sd_{env}^2
$$

Note that recruitment deviations become autocorrelated even though there is no `rec -> rec` arrow, because they inherit the covariate's autocorrelation through $b_{env}$. Putting this together in a compact matrix form, this can be expressed as:

$$
\Sigma = \mathrm{Var}(x) = (I - B)^{-1}\,\Gamma\Gamma^\top\,(I - B)^{-\top}
$$

Here, $(I - B)^{-1}$ substitutes the equations into each other, as we did by hand above. Each row writes a cell as a sum of the innovations that flow into it:

$$
(I - B)^{-1} =
\begin{pmatrix}
1 & 0 & 0 & 0 & 0 & 0 \\
\rho & 1 & 0 & 0 & 0 & 0 \\
\rho^2 & \rho & 1 & 0 & 0 & 0 \\
b_{env} & 0 & 0 & 1 & 0 & 0 \\
b_{env}\rho & b_{env} & 0 & 0 & 1 & 0 \\
b_{env}\rho^2 & b_{env}\rho & b_{env} & 0 & 0 & 1
\end{pmatrix}
\begin{matrix}
\leftarrow env_1 \\ \leftarrow env_2 \\ \leftarrow env_3 \\ \leftarrow rec_1 \\ \leftarrow rec_2 \\ \leftarrow rec_3
\end{matrix}
$$

For example, row 3 says $env_3 = \rho^2\varepsilon^{env}_1 + \rho\,\varepsilon^{env}_2 + \varepsilon^{env}_3$. The innovations themselves are independent, with variance:

$$
\Gamma\Gamma^\top =
\begin{pmatrix}
sd_{env}^2 & 0 & 0 & 0 & 0 & 0 \\
0 & sd_{env}^2 & 0 & 0 & 0 & 0 \\
0 & 0 & sd_{env}^2 & 0 & 0 & 0 \\
0 & 0 & 0 & sd_{rec}^2 & 0 & 0 \\
0 & 0 & 0 & 0 & sd_{rec}^2 & 0 \\
0 & 0 & 0 & 0 & 0 & sd_{rec}^2
\end{pmatrix}
$$

Multiplying these out, and writing $v_1 = sd_{env}^2$, $v_2 = (1+\rho^2)\,sd_{env}^2$, and $v_3 = (1+\rho^2+\rho^4)\,sd_{env}^2$ for the covariate variances derived above, the full variance is:

$$
\Sigma =
\begin{pmatrix}
v_1 & \rho v_1 & \rho^2 v_1 & b_{env} v_1 & b_{env}\rho v_1 & b_{env}\rho^2 v_1 \\
\rho v_1 & v_2 & \rho v_2 & b_{env}\rho v_1 & b_{env} v_2 & b_{env}\rho v_2 \\
\rho^2 v_1 & \rho v_2 & v_3 & b_{env}\rho^2 v_1 & b_{env}\rho v_2 & b_{env} v_3 \\
b_{env} v_1 & b_{env}\rho v_1 & b_{env}\rho^2 v_1 & b_{env}^2 v_1 + sd_{rec}^2 & b_{env}^2\rho v_1 & b_{env}^2\rho^2 v_1 \\
b_{env}\rho v_1 & b_{env} v_2 & b_{env}\rho v_2 & b_{env}^2\rho v_1 & b_{env}^2 v_2 + sd_{rec}^2 & b_{env}^2\rho v_2 \\
b_{env}\rho^2 v_1 & b_{env}\rho v_2 & b_{env} v_3 & b_{env}^2\rho^2 v_1 & b_{env}^2\rho v_2 & b_{env}^2 v_3 + sd_{rec}^2
\end{pmatrix}
\begin{matrix}
\leftarrow env_1 \\ \leftarrow env_2 \\ \leftarrow env_3 \\ \leftarrow rec_1 \\ \leftarrow rec_2 \\ \leftarrow rec_3
\end{matrix}
$$

The structure of the covariance can then be seen as:

- The top-left block is the covariate's AR(1) variance.
- The off-diagonal blocks are that same block scaled by $b_{env}$.
- The bottom-right block is that same block scaled by $b_{env}^2$, plus recruitment's own $sd_{rec}^2$ on the diagonal.

Every entry of $\Sigma$ is non-zero, because every cell shares at least one innovation with every other cell (only in one direction for later years).

In practice, the density is evaluated using the precision $Q = \Sigma^{-1} = (I - B)^\top(\Gamma\Gamma^\top)^{-1}(I - B)$ rather than $\Sigma$. Written out, $Q$ is sparse:

$$
Q =
\begin{pmatrix}
\frac{1+\rho^2}{sd_{env}^2} + \frac{b_{env}^2}{sd_{rec}^2} & -\frac{\rho}{sd_{env}^2} & 0 & -\frac{b_{env}}{sd_{rec}^2} & 0 & 0 \\
-\frac{\rho}{sd_{env}^2} & \frac{1+\rho^2}{sd_{env}^2} + \frac{b_{env}^2}{sd_{rec}^2} & -\frac{\rho}{sd_{env}^2} & 0 & -\frac{b_{env}}{sd_{rec}^2} & 0 \\
0 & -\frac{\rho}{sd_{env}^2} & \frac{1}{sd_{env}^2} + \frac{b_{env}^2}{sd_{rec}^2} & 0 & 0 & -\frac{b_{env}}{sd_{rec}^2} \\
-\frac{b_{env}}{sd_{rec}^2} & 0 & 0 & \frac{1}{sd_{rec}^2} & 0 & 0 \\
0 & -\frac{b_{env}}{sd_{rec}^2} & 0 & 0 & \frac{1}{sd_{rec}^2} & 0 \\
0 & 0 & -\frac{b_{env}}{sd_{rec}^2} & 0 & 0 & \frac{1}{sd_{rec}^2}
\end{pmatrix}
\begin{matrix}
\leftarrow env_1 \\ \leftarrow env_2 \\ \leftarrow env_3 \\ \leftarrow rec_1 \\ \leftarrow rec_2 \\ \leftarrow rec_3
\end{matrix}
$$

The non-zero entries in $Q$ correspond exactly to the arrows. There is one entry for each `env -> env` link between adjacent years and one for each `env -> rec` link in the same year. A zero means two cells are independent given all the others (conditional independence, which induces sparsity). For example, $rec_1$ and $rec_2$ are unrelated once $env_1$ and $env_2$ are known. $Q$ is built directly from the sparse $I - B$ without inverting anything, which is why dsem (and RTMB) can evaluate the density efficiently even when there are many years and series. An example of this is shown below.

```{r density}
cells <- SPoRC:::get_dsem_cells(m, n_grid_yrs = 3)   # internal helpers, so the package prefix
parts <- SPoRC:::get_dsem_matrices(dsem_beta = c(b_env = 0.5, rho = 0.6), ln_dsem_sd = log(c(sd_env = 1, sd_rec = 0.8)),
                                   dsem_model = m, dsem_cells = cells)
as.matrix(parts$IminusB)
parts$sd_cell
```

### Conditional and Marginal Variance in dsem

A dsem time-series has two ways to represent spread, and the sd line (e.g., `rec <-> rec, 0, sd_rec, 0.5`) can describe either one:

- Conditional (the default): the sd line is the size of each year's new shock, $\varepsilon_t$. It is "conditional" because it is how much the series would still vary if we already knew last year's value and any covariates pointing in. It is analogous to the residual sd in a regression. The series itself ends up more variable than this, because the arrows pass variability forward from the past.
- Marginal: the sd line is the size of the series itself, $x_t$, i.e., how much the series itself varies overall, including what goes in through the arrows. It is analogous to the raw sd of $y$ in a regression. The model then works out how big each year's shock must be to hit that target.

The two are equal only when no arrows point into the series. The `dsem_variance` setting in the setup functions (introduced later) chooses which one the sd line fixes.

#### Example: an AR(1) series

Let's use an AR1 series to motivate the difference between conditional and marginal variance, where $x_t = \rho\, x_{t-1} + \varepsilon_t$ with sd line $s$. Substituting back, year 3 is a sum of past innovations, with older ones shrunk by $\rho$:

$$
x_3 = \varepsilon_3 + \rho\,\varepsilon_2 + \rho^2\varepsilon_1
\quad\Rightarrow\quad
\mathrm{var}(x_3) = \mathrm{var}(\varepsilon_3) + \rho^2\,\mathrm{var}(\varepsilon_2) + \rho^4\,\mathrm{var}(\varepsilon_1)
$$

The two parameterizations fix opposite sides of this equation:

- `"conditional"` (the default) sets $\mathrm{var}(\varepsilon_t) = s^2$ every year. The series' spread then grows and settles at $s^2 / (1 - \rho^2)$ (i.e., the stationary variance of $x_t$).
- `"diagonal"` / `"marginal"` set $\mathrm{var}(x_t) = s^2$ every year and then solve backwards for the innovation (i.e., figures out what the innovations should be to satisfy the condition). 

With $\rho = 0.5$ and $s = 1$:

| Year | Conditional $\mathrm{sd}(\varepsilon_t)$ | Conditional $\mathrm{var}(x_t)$ | Marginal $\mathrm{sd}(\varepsilon_t)$ | Marginal $\mathrm{var}(x_t)$ |
|---|---|---|---|---|
| 1 | 1 | 1.00 | 1 | 1 |
| 2 | 1 | 1.25 | 0.87 | 1 |
| 3 | 1 | 1.31 | 0.87 | 1 |
| $\infty$ | 1 | 1.33 | 0.87 | 1 |

In short, under conditional $s$ is the size of each year's new shock. Under marginal, $s$ is the size of the swings you actually see in the series.

Note that a random walk ($\rho = 1$) cannot use the marginal form. The past alone already fills the target ($\mathrm{var}(\varepsilon_2) = s^2 - s^2 = 0$), which leaves no room for new noise, so the setup refuses that pairing (i.e., because the random walk series is non-stationary).

#### Why it matters for recruitment

The choice matters most for recruitment. Take `env -> rec, 0, b_env`, so each year's recruitment deviation is the covariate's effect plus a recruitment shock:

$$
rec_t = b_{env}\, env_t + \varepsilon^{rec}_t
$$

The two pieces are independent, so their variances add:

$$
\mathrm{var}(rec_t) = b_{env}^2\,\mathrm{var}(env_t) + \mathrm{var}(\varepsilon^{rec}_t)
$$

To make this concrete, suppose $b_{env} = 0.5$, $\mathrm{var}(env_t) = 1$, and the sd line is `sd_rec` $= 0.6$. The covariate's contribution is then $0.5^2 \times 1 = 0.25$, and the two settings split things differently:

| | Covariate part | Shock part, $\mathrm{var}(\varepsilon^{rec}_t)$ | Total, $\mathrm{var}(rec_t)$ | Total sd |
|---|---|---|---|---|
| No covariate ($b_{env} = 0$) | 0 | 0.36 | 0.36 | 0.60 |
| Conditional | 0.25 | $0.6^2 = 0.36$ | $0.25 + 0.36 = 0.61$ | 0.78 |
| Marginal | 0.25 | $0.36 - 0.25 = 0.11$ | $0.6^2 = 0.36$ | 0.60 |

- Thus, under conditional, `sd_rec` fixes the shock, so the covariate adds variability on top of it. Recruitment deviations become more variable (sd 0.78) than in a model without the covariate.
- By contrast, under marginal, `sd_rec` fixes the total, so the covariate takes a share of it. Recruitment deviations keep the same spread (sd 0.60) as an iid model, and the covariate explains $0.25 / 0.36 \approx 70\%$ of it, leaving a smaller shock (sd 0.33).

In other words, marginal keeps `sd_rec` comparable to a standard assessment with iid recruitment deviations. Conditional keeps it comparable to `dsem`, where `sd_rec` is the variability left over after the covariate.

#### Diagonal versus marginal

In general, `"diagonal"` and `"marginal"` both read the sd line as the spread of the series, i.e., $\mathrm{var}(x_t) = s^2$. They differ only in how they handle covariance between series (`a <-> b`). Without covariance specified, the two give identical results.

Consider the recruitment example again where, $rec_t = b_{env}\, env_t + \varepsilon^{rec}_t$, and add `env <-> rec`, so the environmental and recruitment innovations are themselves correlated (correlation $r$). The two pieces of $rec_t$ are then no longer independent, and their variances no longer simply add. Therefore, a cross term appears:

$$
\mathrm{var}(rec_t) = \underbrace{b_{env}^2\,\mathrm{var}(env_t)}_{\text{covariate}} + \underbrace{\mathrm{var}(\varepsilon^{rec}_t)}_{\text{shock}} + \underbrace{2\, b_{env}\, r\, \mathrm{sd}(\varepsilon^{env}_t)\, \mathrm{sd}(\varepsilon^{rec}_t)}_{\text{cross term}}
$$

- In this context, `"diagonal"` solves for the shock as if the cross term were not there, i.e., $\mathrm{var}(\varepsilon^{rec}_t) = s^2 - b_{env}^2\,\mathrm{var}(env_t)$. The cross term is then added on top, so the series misses the target.
- `"marginal"` includes the cross term when solving. Because the unknown shock sd appears in the cross term as well, there is no simple closed form, so the model finds it with a short fixed-point iteration (guess, update, repeat until it converges to some solution).

Using the same numbers as above ($b_{env} = 0.5$, $\mathrm{var}(env_t) = 1$, target $s^2 = 0.36$) plus $r = 0.3$:

| | Shock variance | Covariate part | Cross term | Total $\mathrm{var}(rec_t)$ |
|---|---|---|---|---|
| Diagonal | $0.36 - 0.25 = 0.11$ | 0.25 | $2(0.5)(0.3)(1)(0.33) = 0.10$ | 0.46 (misses 0.36) |
| Marginal | 0.046 | 0.25 | $2(0.5)(0.3)(1)(0.21) = 0.064$ | 0.36 (hits target) |

In practice, use `"diagonal"` when there are no covariance lines, since it is simpler and gives the same answer. However, we recommend specifying `"marginal"` when there are covariance lines and you need the series' spread to equal the sd line exactly.

### Series observed with error

In the above, we assumed that the grid holds a series $x_t$ observed without error, and the arrows specify how it may move from year to year. However, this is rarely the case, where there is often observation error. In such a case, this can be specified as:

$$y_t \sim f(x_t)$$

where $y_t$ is what you measured in year $t$ and $f$ is whatever density suits the measurement, which is set by the `dsem_family` argument (discussed further below). The mean of $f$ is $x_t$ through a link function, so that $x_t$ sits on a scale where a normal makes sense: counts take a log and $x_t$ is the log mean count, while proportions take a logit link instead.

| `dsem_family` | link | mean of $y_t$ | extra spread |
|---|---|---|---|
| `fixed` | identity | $x_t$ itself | none, and no observation density |
| `normal`, `gaussian` | identity | $x_t$ | estimated sd |
| `gaussian_fixed_sd` | identity | $x_t$ | one supplied per observation |
| `lognormal` | log | median $e^{x_t}$ | estimated sd, log scale |
| `poisson` | log | $e^{x_t}$ | none, set by the mean |
| `bernoulli`, `binomial` | logit | $1/(1+e^{-x_t})$ | none, set by the mean |
| `gamma` | log | $e^{x_t}$ | estimated coefficient of variation |
| `tweedie` | log | $e^{x_t}$ | estimated sd and power |

Because $y_t$ is an observation of the time-series and not the true time-series itself, $x_t$ (the true underlying state that generated the time-series) can be estimated in every year, whether or not that year has an observation. Thus, $x_t$ becomes the random effect for that time-series, and the years you did not observe are interpolated by the arrow-lag notation specified. In the case where a year with no observation contributes no $y_t \sim f(x_t)$ term, the arrows decide what that value should be (i.e., its conditional mean given the other years). For instance, under an AR1 it comes back as $\rho(x_{t-1} + x_{t+1})/(1+\rho^2)$ with standard deviation $\sigma/\sqrt{1+\rho^2}$, where $\sigma$ is the innovation sd. Under independent deviations it returns the series mean at the full $\sigma$. Supplying known values is the one case with no observation density at all, and there the series is the data and nothing is estimated for it (i.e., `dsem_family = fixed`).

## Recruitment bias correction behavior under dsem
Note that when using a dsem, the behavior of a recruitment bias correction changes. Firstly, under a dsem, `SPoRC` enforces the condition that no recruitment bias ramp can be turned on. Next, `bias_correct_pe` is a switch that determines whether a lognormal bias correction using the marginal variance of the dsem linked series is applied. Moreover, under a dsem, the variances used for bias correction are determined by the arrows specified. 

Two implications follow for the initial age deviations. Because the arrows take over the recruitment
deviations' density, `ln_sigmaR` is not used by anything, and the initial ages are instead held at the recruitment
series' marginal variance under the arrows, taken at the last year the link covers, provided that variance
settles. For a stationary series it stops changing well before the end, so the choice of year is immaterial.
Two kinds of series never settle, and each falls back a step: a random walk, written as a lagged self path of
one, would otherwise take a value that grows with the length of the model, so it reads its own `<->` sd line
instead, which is what a walk fitted outside a dsem also does; and a series whose sd is moderated by another
has no single sd line to fall back to, so it reads `ln_sigmaR` as supplied, which is the one case where that
argument is still live under a dsem. The operating model applies the same three rules, so a simulation draws
the initial deviations at whatever the estimation model penalizes them at. `Setup_Sim_DSEM` writes the same value into the operating model's
`ln_sigmaR`, so a simulation draws the initial deviations at the spread the estimation model penalizes them
at. This only matters where those deviations are penalized at all: under `equil_init_age_strc = "equil"`
none are estimated, and under `"stoch_all_no_pen"` they are estimated without a penalty, so the marginal
variance goes unread in both cases.

That marginal variance is taken conditional on whatever the model treats as known, which is every year a
covariate is observed together with the years before a linked series begins. A covariate observed in every
year therefore contributes nothing to it, and the marginal collapses to the innovation standard deviation
named on that series' own `<->` line; only missing years or a latent series widen it. This is worth keeping
in mind when the value that comes back is smaller than the spread of the simulated series would suggest.

Finally, a series whose variance is itself moderated by another series has no single variance to correct on,
and the initial ages, born before the grid begins, have no defensible one at all. `SPoRC` refuses that
combination, so a moderated recruitment variance requires `bias_correct_pe = "none"`.

## Using dsem inside an assessment model

Having described the underpinnings of a dsem model, we next seek to integrate such a model class within the context of a stock assessment model. The following example does so using recruitment deviations as the series we attempt to model using a dsem. We use a contrived example of the packaged GOA dusky rockfish fit (`dusky_rtmb_model`) and demonstrate various ways in which dsem can be used to specify recruitment deviations, ranging from penalized likelihood with iid deviations, random effects with iid deviations, and environmental linkages to recruitment. In the following example, we demonstrate a penalized likelihood approach, which yields the same model results as the GOA dusky vignette.

```{r dusky_base, message = FALSE, warning = FALSE}
# setup base config
data("dusky_rtmb_model")
base <- list(data = dusky_rtmb_model$data, par = dusky_rtmb_model$parameters, map = dusky_rtmb_model$mapping, verbose = FALSE, store_config = FALSE)
# the penalized fit below reproduces the packaged one, which doesnt do bias correction. the random effect fits take the full lognormal correction, so they read a base that asks for it
re_base <- base
re_base$data$bias_correct_pe <- 1

# setup recruitment for dsem
pen_base <- Setup_Mod_Rec(
  input_list = base,
  sigmaR_switch = 1,
  ln_sigmaR = array(-0.1068576, dim = c(2, base$data$n_pop, base$data$n_regions)),
  rec_model = "mean_rec",
  init_age_strc = 1,
  ln_global_R0 = log(2.7),
  t_spawn = base$data$t_spawn,
  RecDevs_model = "dsem" # set recdev model as dsem
)

# setup dsem - penalized likelihood w/ sigmaR fixed
pen_recdev <- Setup_Mod_DSEM(
  pen_base,
  dsem_arrows = "
  rec <-> rec, 0, NA, 0.8986536
  ",
  dsem_data = NULL
)

# fit model
pen_recdev_mod <- fit_model(
  pen_recdev$data,
  pen_recdev$par,
  pen_recdev$map,
  NULL, silent = TRUE
)

# get sd report
pen_recdev_mod$sd_rep <- sdreport(pen_recdev_mod)
pen_recdev_mod$sd_rep

# check if fit with dsem is the same with previous penalized fit (should be)
pen_recdev_mod$rep$jnLL # dsem
dusky_rtmb_model$rep$jnLL # original

# show plot
plot(pen_recdev_mod$rep$dsem_x_grid, ylab = 'RecDev', xlab = 'Year')
lines(as.vector(dusky_rtmb_model$rep$ln_RecDevs))
```

Having shown the equivalence above, we next estimate recruitment deviations as random effects using the dsem interface. Note that using the dsem interface here to specify iid recruitment is also equivalent to simply estimating `sigmaR` and treating recruitment deviations as random effects (without the dsem interface):

```{r dusky_dsem_comparison, message = FALSE, warning = FALSE}
# random effect deviations take the full lognormal correction, so no ramp
dsem_base <- Setup_Mod_Rec(
  input_list = re_base,
  do_rec_bias_ramp = 0,
  sigmaR_switch = 1,
  ln_sigmaR = array(-0.1068576, dim = c(2, re_base$data$n_pop, re_base$data$n_regions)),
  rec_model = "mean_rec",
  init_age_strc = 1,
  ln_global_R0 = log(2.7),
  t_spawn = re_base$data$t_spawn,
  RecDevs_model = "dsem"
)

# setup dsem - iid state space w/ sigmaR estimate
re_dsem_recdev <- Setup_Mod_DSEM(
  dsem_base,
  dsem_arrows = "
  rec <-> rec, 0, sd_rec, 0.8986536
  ",
  dsem_data = NULL
)

# fit model
re_dsem_recdev_mod <- fit_model(
  re_dsem_recdev$data,
  re_dsem_recdev$par,
  re_dsem_recdev$map,
  "ln_RecDevs", silent = TRUE
)

# get sd report
re_dsem_recdev_mod$sd_rep <- sdreport(re_dsem_recdev_mod)
re_dsem_recdev_mod$sd_rep

# setup w/o dsem. one sigmaR across the early and late period
nodsem_base <- Setup_Mod_Rec(
  input_list = re_base,
  do_rec_bias_ramp = 0,
  sigmaR_switch = 1,
  ln_sigmaR = array(-0.1068576, dim = c(2, re_base$data$n_pop, re_base$data$n_regions)),
  rec_model = "mean_rec",
  init_age_strc = 1,
  ln_global_R0 = log(2.7),
  t_spawn = re_base$data$t_spawn,
  RecDevs_model = "iid",
  sigmaR_spec = "est_shared_all"
)

# fit model w/o dsem
re_nodsem_recdev_mod <- fit_model(
  nodsem_base$data,
  nodsem_base$par,
  nodsem_base$map,
  "ln_RecDevs", silent = TRUE
)

# get sd report
re_nodsem_recdev_mod$sd_rep <- sdreport(re_nodsem_recdev_mod)
re_nodsem_recdev_mod$sd_rep

# should give equivalent results
re_dsem_recdev_mod$rep$jnLL
re_nodsem_recdev_mod$rep$jnLL

# show plot
plot(re_dsem_recdev_mod$rep$dsem_x_grid, ylab = 'RecDev', xlab = 'Year')
lines(as.vector(re_nodsem_recdev_mod$rep$ln_RecDevs))
```

Next, we can specify an AR(1) process on recruitment deviations for Dusky Rockfish. This is done using arrow-lag notation, and is shown in the following:

```{r dusky_dsem_ar1_comparison, message = FALSE, warning = FALSE}
# random effect deviations take the full lognormal correction, so no ramp
dsem_base <- Setup_Mod_Rec(
  input_list = re_base,
  do_rec_bias_ramp = 0,
  sigmaR_switch = 1,
  ln_sigmaR = array(-0.1068576, dim = c(2, re_base$data$n_pop, re_base$data$n_regions)),
  rec_model = "mean_rec",
  init_age_strc = 1,
  ln_global_R0 = log(2.7),
  t_spawn = re_base$data$t_spawn,
  RecDevs_model = "dsem"
)

# setup dsem - ar1 state space w/ sigmaR estimate
re_ar1_dsem_recdev <- Setup_Mod_DSEM(
  dsem_base,
  dsem_arrows = "
  rec -> rec, 1, ar1_rho, 0.5
  rec <-> rec, 0, sd_rec, 0.8986536
  ",
  dsem_data = NULL
)

# fit model
re_ar1_dsem_recdev_mod <- fit_model(
  re_ar1_dsem_recdev$data,
  re_ar1_dsem_recdev$par,
  re_ar1_dsem_recdev$map,
  "ln_RecDevs", silent = TRUE
)

# get sd report
re_ar1_dsem_recdev_mod$sd_rep <- sdreport(re_ar1_dsem_recdev_mod)
re_ar1_dsem_recdev_mod$sd_rep

# compare with iid version
plot(re_ar1_dsem_recdev_mod$rep$dsem_x_grid, ylab = 'RecDev', xlab = 'Year', col = 'blue', ylim = c(-2, 3), type = 'l')
lines(re_dsem_recdev_mod$rep$dsem_x_grid, col = 'red')
```

As shown in the plot above, the AR(1) process is a lot smoother relative to an `iid` assumption, as would be expected.

### Linking covariates to recruitment

A covariate can appear correlated with recruitment only because it drives something else
that matters. For instance, a large-scale climate index `x` might affect local temperature `y`, which
then affects recruitment. Regressed on its own `x` looks to be useful, but once `y` is in it has
nothing left to add. Defining a dsem writes both paths at once and allows us to better understand causal drivers,
which is the case Champagnat et al. (2026) make for ecosystem and climate-linked assessments generally:

$$
\begin{aligned}
y_t &= b_{xy}\, x_t + \varepsilon^{y}_t \\
rec_t &= b_{x,rec}\, x_t + b_{y,rec}\, y_t + \varepsilon^{rec}_t
\end{aligned}
$$

where $x_t$ and $y_t$ are the two covariates in year $t$, $rec_t$ is the recruitment
deviation, each $b$ is the path coefficient named on its arrow, and each $\varepsilon$ is
that series' innovation. The direct effect of `x` is $b_{x,rec}$ and its indirect effect is
$b_{xy} \times b_{y,rec}$. Under full mediation $b_{x,rec} = 0$ by the way we have defined this model, so the $b_{x,rec}\,x_t$ term drops out and recruitment depends on `x` only through `y`.

We use the packaged dusky recruitment deviations and build two contrived simulated covariates. `y` is
those deviations plus noise, so it relates to recruitment directly. `x` is `y` plus noise,
so it relates to recruitment only through `y`. Both are standardized.

```{r mediation_data}
set.seed(777)
rec_dev <- as.vector(dusky_rtmb_model$rep$ln_RecDevs[1,1,])
y <- as.vector(scale(rec_dev + rnorm(length(rec_dev), 0, 0.9)))
x <- as.vector(scale(y + rnorm(length(y), 0, 0.5)))
cov_df <- data.frame(year = base$data$years, x = x, y = y)

# x correlates with recruitment, but adds nothing once y is in the regression
c(cor_xy = cor(x, y),
  cor_x_rec = cor(x, rec_dev),
  cor_y_rec = cor(y, rec_dev),
  partial_x = coef(lm(rec_dev ~ x + y))["x"])
```

`x` correlates with recruitment at 0.58, which is what regressing one covariate at a time
would pick up, and its coefficient in a regression holding both is -0.11.

Both fits below reuse `dsem_base`, the input list from the comparison above, which already
hands the recruitment deviations to the dsem. Both covariates are standardized so their
means are fixed at zero, and both are observed in every year, so the recruitment deviations
are the only random effects.

```{r mediation_fits, message = FALSE, warning = FALSE}
fit_dsem_cov <- function(arrows) {
  il <- Setup_Mod_DSEM(
    dsem_base,
    dsem_arrows = arrows,
    dsem_data = cov_df,
    dsem_mu_spec = c(x = 0, y = 0)
  )
  mod <- fit_model(il$data, il$par, il$map, "ln_RecDevs", silent = TRUE)
  mod$sd_rep <- sdreport(mod)
  mod$arrow_names <- c(il$data$dsem_model$beta_names, il$data$dsem_model$ln_sd_names)
  mod
}

# x regressed on its own, with the mediator left out of recruitment
x_only <- fit_dsem_cov("
  x <-> x, 0, sd_x, 1
  x -> y, 0, b_xy, 0.9
  y <-> y, 0, sd_y, 0.5
  x -> rec, 0, b_x_rec, 0.3
  rec <-> rec, 0, sd_rec, 0.8986536
")

# both paths in, so the fit can presuambly identify a direct effect of x if one is there
both <- fit_dsem_cov("
  x <-> x, 0, sd_x, 1
  x -> y, 0, b_xy, 0.9
  y <-> y, 0, sd_y, 0.5
  x -> rec, 0, b_x_rec, 0
  y -> rec, 0, b_y_rec, 0.3
  rec <-> rec, 0, sd_rec, 0.8986536
")

# arrow estimates, named
lapply(list(x_only = x_only, both = both), function(m) {
  tab <- summary(m$sd_rep, "fixed")
  tab <- tab[rownames(tab) %in% c("dsem_beta", "ln_dsem_sd"), , drop = FALSE]
  rownames(tab) <- m$arrow_names
  round(tab, 4)
})
```

Regressed on its own, `x` gets a direct effect of 0.664 with a standard error of 0.158.
Put `y` in beside it and the same coefficient is -0.030 against a standard error of 0.408,
while `y` takes 0.718. The path from `x` to `y` comes back at 0.919 in both fits, since it is just
the path for the two covariates alone. Holding both paths in one fit is what separates those effects,
and regressing one covariate at a time cannot, demonstrating the utility of a dsem approach.

## Self-testing
Given the models above, one can use simulation self-testing to understand whether or not the coefficients obtained were spurious due to short time-series length, and whether or not results obtained from the dsem integrated approach are robust. In this case, the `simulation_self_test` function conducts a conditional simulation where the random effects / states are held at the estimated values and considered the truth, but new observational datasets are generated and fitted to. 

```{r cond}
n_cores <- min(10, future::availableCores())
self_test_cond <- simulation_self_test(
  both$data,
  both$parameters,
  both$mapping,
  random = 'ln_RecDevs',
  rep = both$rep,
  sd_rep = both$sd_rep,
  n_sims = 10,
  newton_loops = 2,
  n_cores = n_cores,
  what = c("SSB", "Rec", "dsem_x_grid"),
  what_par = c("dsem_beta", "ln_dsem_sd"),
  do_par = TRUE,
  sim_type = 'conditional'
)
```

```{r, fig.width = 11, fig.height = 4.5}
par(mfrow = c(1,3), mar = c(5, 5.5, 1, 1), mgp = c(3.2, 0.9, 0),
    cex.lab = 1.5, cex.axis = 1.25, las = 1)
plot(apply((self_test_cond$SSB - self_test_cond$truth$SSB) / self_test_cond$truth$SSB, 3, median),
     ylim = c(-0.5, 0.5), xlab = 'Year', ylab = 'Relative Error in SSB',
     type = 'l', lwd = 2)
abline(h = 0, lty = 1)

plot(apply((self_test_cond$Rec - self_test_cond$truth$Rec) / self_test_cond$truth$Rec, 3, median),
     ylim = c(-0.5, 0.5), xlab = 'Year', ylab = 'Relative Error in Recruitment',
     type = 'l', lwd = 2)
abline(h = 0, lty = 1)

# dsem stuff
beta_mat <- (self_test_cond$dsem_beta - self_test_cond$truth$dsem_beta) / self_test_cond$truth$dsem_beta
sd_mat <- (exp(self_test_cond$ln_dsem_sd) - exp(self_test_cond$truth$ln_dsem_sd)) / exp(self_test_cond$truth$ln_dsem_sd)
dsem_par_mat <- rbind(beta_mat, sd_mat)
rownames(dsem_par_mat) <- c("b_xy", "b_x", "b_y", "sd_x", "sd_y", "sd_rec")
stripchart(as.data.frame(t(dsem_par_mat)),
           vertical = TRUE, method = "jitter", pch = 16, cex = 1.1, xaxt = "n", xlab = "",
           ylab = "Relative error", ylim = c(-3, 3))
axis(1, at = 1:nrow(dsem_par_mat), labels = rownames(dsem_par_mat), las = 2)
abline(h = 0, lty = 2)
```

A conditional simulation keeps the states fixed, so variation across replicates comes only from
observation error. Each coefficient is therefore tested only at the fitted state. Inspecting the figure above,
we notice that some biases arise. We hypothesize that this is likely due to imperfect data availiability
and quantity, rather than model misspecification or coding errors (though like an identifiability issue), and we
verify this by setting `perfect_data = TRUE`. Setting `perfect_data = TRUE` shrinks every
observation standard deviation to 0.001 and every input sample size to a million, while leaving
process error alone. With virtually zero observation error, a correctly specified model returns the operating model to several decimals. Anything that still fails to come back is a property of the model or of the simulation rather than sampling
noise, and anything that increases in precision was only ever noise.

```{r cond_perfect}
self_test_cond_perfect <- simulation_self_test(
  both$data,
  both$parameters,
  both$mapping,
  random = 'ln_RecDevs',
  rep = both$rep,
  sd_rep = both$sd_rep,
  n_sims = 10,
  newton_loops = 2,
  n_cores = n_cores,
  what = c("SSB", "Rec", "dsem_x_grid"),
  what_par = c("dsem_beta", "ln_dsem_sd"),
  do_par = TRUE,
  perfect_data = TRUE,
  sim_type = 'conditional'
)
```

```{r, fig.width = 11, fig.height = 4.5}
par(mfrow = c(1,3), mar = c(5, 5.5, 1, 1), mgp = c(3.2, 0.9, 0),
    cex.lab = 1.5, cex.axis = 1.25, las = 1)
plot(apply((self_test_cond_perfect$SSB - self_test_cond_perfect$truth$SSB) / self_test_cond_perfect$truth$SSB, 3, median),
     ylim = c(-0.5, 0.5), xlab = 'Year', ylab = 'Relative Error in SSB',
     type = 'l', lwd = 2)
abline(h = 0, lty = 1)

plot(apply((self_test_cond_perfect$Rec - self_test_cond_perfect$truth$Rec) / self_test_cond_perfect$truth$Rec, 3, median),
     ylim = c(-0.5, 0.5), xlab = 'Year', ylab = 'Relative Error in Recruitment',
     type = 'l', lwd = 2)
abline(h = 0, lty = 1)

# dsem stuff
beta_mat <- (self_test_cond_perfect$dsem_beta - self_test_cond_perfect$truth$dsem_beta) / self_test_cond_perfect$truth$dsem_beta
sd_mat <- (exp(self_test_cond_perfect$ln_dsem_sd) - exp(self_test_cond_perfect$truth$ln_dsem_sd)) / exp(self_test_cond_perfect$truth$ln_dsem_sd)
dsem_par_mat <- rbind(beta_mat, sd_mat)
rownames(dsem_par_mat) <- c("b_xy", "b_x", "b_y", "sd_x", "sd_y", "sd_rec")
stripchart(as.data.frame(t(dsem_par_mat)),
           vertical = TRUE, method = "jitter", pch = 16, cex = 1.1, xaxt = "n", xlab = "",
           ylab = "Relative error", ylim = c(-3, 3))
axis(1, at = 1:nrow(dsem_par_mat), labels = rownames(dsem_par_mat), las = 2)
abline(h = 0, lty = 2)
```

The standard deviations behave differently from the path coefficients under this comparison. A path
coefficient is identified by the precision of the data, so it tightens sharply. A standard deviation
is identified by the number of deviations it is estimated from, and perfect data adds no years to
the series, which is why its variability barely moves. Residual spread on `sd_rec` is a signal about the length of the time-series rather than an error in the model and coding. Following verificaiton of a conditional simulation, we next turn to a joint simulation approach using the same function `simulation_self_test`. However, this approach further requires a user to provide a joint precision matrix, from which the function can sample new parameter vectors from.

The conditional and joint self tests address different sources of uncertainty. A conditional simulation treats the fitted parameters as fixed and asks whether the model can recover those effects from data with the same observation error. A joint simulation also allows the parameters to vary according to their estimated uncertainty, asking whether the model can recover effects across the range supported by the fit. Practically, a successful conditional test shows that the model can recover the estimated effects if those effects are correct. A successful joint test provides stronger evidence that the effects can be recovered despite uncertainty in their true values. If the conditional test succeeds but the joint test performs poorly, the estimated effects may be recoverable at their fitted values but are not reliably distinguishable across the range of values supported by the data.

```{r joint}
both$sd_rep <- sdreport(both, getJointPrecision = TRUE) # get joint precision matrix
self_test_joint <- simulation_self_test(
  both$data,
  both$parameters,
  both$mapping,
  random = 'ln_RecDevs',
  obj = both,
  rep = both$rep,
  sd_rep = both$sd_rep,
  n_sims = 10,
  newton_loops = 2,
  n_cores = n_cores,
  what = c("SSB", "Rec", "dsem_x_grid"),
  what_par = c("dsem_beta", "ln_dsem_sd"),
  do_par = TRUE,
  sim_type = 'joint'
)
```

```{r, fig.width = 11, fig.height = 4.5}
par(mfrow = c(1,3), mar = c(5, 5.5, 1, 1), mgp = c(3.2, 0.9, 0),
    cex.lab = 1.5, cex.axis = 1.25, las = 1)
plot(apply((self_test_joint$SSB - self_test_joint$truth$SSB) / self_test_joint$truth$SSB, 3, median),
     ylim = c(-0.5, 0.5), xlab = 'Year', ylab = 'Relative Error in SSB',
     type = 'l', lwd = 2)
abline(h = 0, lty = 1)

plot(apply((self_test_joint$Rec - self_test_joint$truth$Rec) / self_test_joint$truth$Rec, 3, median),
     ylim = c(-0.5, 0.5), xlab = 'Year', ylab = 'Relative Error in Recruitment',
     type = 'l', lwd = 2)
abline(h = 0, lty = 1)

# dsem stuff
beta_mat <- (self_test_joint$dsem_beta - self_test_joint$truth$dsem_beta) / self_test_joint$truth$dsem_beta
sd_mat <- (exp(self_test_joint$ln_dsem_sd) - exp(self_test_joint$truth$ln_dsem_sd)) / exp(self_test_joint$truth$ln_dsem_sd)
dsem_par_mat <- rbind(beta_mat, sd_mat)
rownames(dsem_par_mat) <- c("b_xy", "b_x", "b_y", "sd_x", "sd_y", "sd_rec")
stripchart(as.data.frame(t(dsem_par_mat)),
           vertical = TRUE, method = "jitter", pch = 16, cex = 1.1, xaxt = "n", xlab = "",
           ylab = "Relative error", ylim = c(-3, 3))
axis(1, at = 1:nrow(dsem_par_mat), labels = rownames(dsem_par_mat), las = 2)
abline(h = 0, lty = 2)
```

The two direct paths are the informative part of these figures. `b_xy` returns tightly, while `b_x`
and `b_y` scatter widely in both directions even though the biomass and recruitment they generate
come back almost exactly. That is the mediation structure present in the dsem model. The covariate reaches recruitment both directly and through the mediator, and presuambly a series of this length has difficulty identifying how the
total effect divides between the two paths, only what the total is. Nonetheless, based on the similarities between the conditional and joint simulation and recovery of parameters when `perfect_data = TRUE`, it is likely that biases from this model are likely attributed to uneven distribution of data availiability (i.e., fewer data points availiable early and later on in the time-series) rather than an ill conditioned model / poorly specified model.

## Cross-testing
Beyond setting up a simulation self-test in, users can also set up a simulation cross-testing using `dsem`. However, this is a more involved process and requires users to manually setup their own simulation. We will draw on previous vignettes and manually setup a dsem simulation from scratch in the following. Most of these steps follow the Simulation Testing (Cross and Self Tests) vignette and we do not elaborate further except for the newwer code snippets introduced in this vignette. In particular, we now demonstrate a dsem linked process for catchability in the code below. We first setup code that is generally familiar to users that understand how to setup a simulation in `SPoRC`.

```{r}
# dimensions
sim_list <- Setup_Sim_Dim(
  n_sims        = 30,  # number of simulations
  n_yrs         = 30,  # number of years
  n_regions     = 1,   # single region
  n_ages        = 10,  # number of ages
  n_lens        = NULL,# no length structure
  n_sexes       = 1,   # single sex
  n_fish_fleets = 1,   # one fishery fleet
  n_srv_fleets  = 1,    # one survey fleet
  n_pop         = 1,   # number of pops
  # the one switch both sides read: the recruitment and initial age deviations are centered on minus half
  # their variance here, and the estimating models below say the same thing
  bias_correct_pe = "rec"
)

# Create storage containers
sim_list <- Setup_Sim_Containers(sim_list)

# fishing stuff
sim_list <- Setup_Sim_Fishing(sim_list = sim_list, # update simulate list
                              # Logistic selectivity
                              fish_sel_input = replicate(
                                n = sim_list$n_sims,
                                array(rep(1 / (1 + exp(-3 * ((1:sim_list$n_ages) - 5))), each = sim_list$n_yrs),
                                      dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages,
                                              sim_list$n_sexes, sim_list$n_fish_fleets))
                              ),
                              # thin age data, so the year class strengths are not already pinned by the compositions
                              ISS_FishAgeComps = array(25, dim = c(sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas,
                                                                   sim_list$n_sexes, sim_list$n_fish_fleets, sim_list$n_sims))
)

# biologicals
sim_list <- Setup_Sim_Biologicals(
  sim_list = sim_list, # simualtion list
  natmort_input = replicate(n = sim_list$n_sims, array(0.3, dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs,
                                                                    sim_list$n_ages, sim_list$n_sexes))), # natural mortality
  WAA_input = replicate(n = sim_list$n_sims, array(rep(5 / (1 + exp(-3 * ((1:sim_list$n_ages) - 3))), each = sim_list$n_yrs),
                                                   dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages, sim_list$n_sexes))), # weight at age
  WAA_fish_input = replicate(n = sim_list$n_sims, array(rep(5 / (1 + exp(-3 * ((1:sim_list$n_ages) - 3))), each = sim_list$n_yrs),
                                                        dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages, sim_list$n_sexes, sim_list$n_fish_fleets))), # fishery weight at age
  WAA_srv_input = replicate(n = sim_list$n_sims, array(rep(5 / (1 + exp(-3 * ((1:sim_list$n_ages) - 3))), each = sim_list$n_yrs),
                                                       dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages, sim_list$n_sexes, sim_list$n_srv_fleets))), # survey weight at age
  MatAA_input = replicate(n = sim_list$n_sims, array(rep(1 / (1 + exp(-3 * ((1:sim_list$n_ages) - 3))), each = sim_list$n_yrs),
                                                     dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages, sim_list$n_sexes))) # maturity at age
)

# recruitment stuff
sim_list <- Setup_Sim_Rec(
  sim_list = sim_list,
  R0_input = replicate(n = sim_list$n_sims, expr = array(5, dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs))), # R0
  ln_sigmaR = array(log(1), dim = c(2, sim_list$n_pop, sim_list$n_regions)),
  recruitment_opt = 'mean_rec',
  # the population starts in equilibrium, so the initial age deviations are zero rather than drawn at sigmaR
  ln_InitDevs_input = array(0, dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_ages - 1,
                                       sim_list$n_sexes, sim_list$n_sims))
)

# stuff that is not used
sim_list <- Setup_Sim_Tagging(
  sim_list = sim_list, # simulation list
  use_conv_fish_tagging = 0
)

# No Movement
sim_list$Movement <- array(1, dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages, sim_list$n_sexes, sim_list$n_sims))
```

Next we set up the catchability dsem. The usual survey setup is unchanged, but a new optional module,
`Setup_Sim_q_devs`, says where a fleet's catchability deviations come from; `"dsem"` give them to the
arrows. The map we simulate under is a chain rather than a single regression: an AR(1) environment drives
recruitment, and recruitment in turn drives survey catchability. Here, the survey index then moves for two reasons at once, and an estimation model that gets the causal map wrong has to put that misspecification somewhere.

```{r}
sim_list <- Setup_Sim_Survey(
  sim_list = sim_list,
  # Logistic selectivity
  srv_sel_input = replicate(
    n = sim_list$n_sims,
    array(rep(1 / (1 + exp(-1 * ((1:sim_list$n_ages) - 3))), each = sim_list$n_yrs),
          dim = c(sim_list$n_pop, sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_ages,
                  sim_list$n_sexes, sim_list$n_srv))
  ),
  # a believable survey, so a drift in catchability is worth mistaking for abundance
  ObsSrvIdx_SE = array(0.1, dim = c(sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas, sim_list$n_srv_fleets)),
  ISS_SrvAgeComps = array(25, dim = c(sim_list$n_regions, sim_list$n_yrs, sim_list$n_seas,
                                      sim_list$n_sexes, sim_list$n_srv_fleets, sim_list$n_sims))
)

# catchability deviations are handed to the arrows instead of a process error of their own
sim_list <- Setup_Sim_q_devs(sim_list, srv_q_model = "dsem")

# a contribed climate linked map - a basin scale index warms the shelf and moves the prey field, those two drive
# recruitment, temperature also shifts where the fish sit, and availability and recruitment drive catchability
arrows <- c("pdo -> pdo, 1, rho_pdo",
            "pdo <-> pdo, 0, sd_pdo",
            "pdo -> temp, 0, b_pdo_temp",
            "temp -> temp, 1, rho_temp",
            "temp <-> temp, 0, sd_temp",
            "pdo -> prey, 0, b_pdo_prey",
            "prey <-> prey, 0, sd_prey",
            "temp -> rec, 0, b_temp_rec",
            "prey -> rec, 0, b_prey_rec",
            "rec <-> rec, 0, sd_rec",
            "temp -> cond, 0, b_temp_cond",
            "cond <-> cond, 0, sd_cond",
            "cond -> srv_q, 0, b_cond_q",
            "rec -> srv_q, 0, b_rec_q",
            "srv_q <-> srv_q, 0, sd_q")

cov_names <- c("pdo", "temp", "prey", "cond") # the four covariates
env_obs_sd <- 0.2 # each covariate is observed with error

# true values for causual map
om_values <- c(rho_pdo = 0.9, sd_pdo = 0.45, b_pdo_temp = 0.7, rho_temp = 0.4, sd_temp = 0.4,
               b_pdo_prey = -0.5, sd_prey = 0.5, b_temp_rec = 0.45, b_prey_rec = 0.35, sd_rec = 0.35,
               b_temp_cond = 0.6, sd_cond = 0.4, b_cond_q = 0.5, b_rec_q = 0.3, sd_q = 0.2)

# dsem time-series, on two processes at once
sim_list <- Setup_Sim_DSEM(sim_list,
                           dsem_arrows = arrows,
                           dsem_values = om_values,
                           dsem_processes = c("rec", "srv_q"),
                           dsem_cov_obs_sd = setNames(rep(env_obs_sd, length(cov_names)), cov_names))

# simulate data
set.seed(123)
sim_obj <- Simulate_Pop_Static(sim_list = sim_list, output_path = NULL) # get simulated datasets
```

The dsem drawn grid now has six columns, in the order the arrows introduce them: the four true covariates, the
recruitment deviation and the catchability deviation. Note also observed covariates, with their measurement error,
are in `sim_obj$dsem_cov_obs_sim`, and those are the only versions any estimating model gets to see.

```{r, fig.width = 11, fig.height = 7}
par(mfrow = c(2, 3), mar = c(4.5, 4.5, 1.5, 1), cex.lab = 1.2, cex.axis = 1.1)
panels <- list(c("PDO", "1"), c("Temperature", "2"), c("Prey", "3"), c("Availability", "5"),
               c("Rec dev", "4"), c("Q dev", "6"))
for(pan in panels) {
  for(i in 1:sim_obj$n_sims) {
    y <- sim_obj$dsem_x_sim[, as.integer(pan[2]), i]
    if(i == 1) plot(y, type = 'l', xlab = 'Year', ylab = pan[1]) else lines(y)
  }
}
```

In this simualted case, the climate index correlates with the catchability deviation even though no arrow connects the two
directly, because recruitment and availability sit between them and mediate the effect. In the map we have constructed, the index
reaches catchability along three directed paths, and each contributes the product of the coefficients along it. In a similar vein, the recruitment and availability deviations correlate at around 0.6 with no arrow between them either, though for a different reason: both descend from temperature, rather than one reaching the other. In general, a dsem implies a correlation between any two series joined by a directed path, and between any two
that share a mediator / node, so the arrows we write down are not the list of pairs the model treats as related, and
a pair we leave unconnected is not strictly independent. 

Note also that a dsem can be specified from scratch, as we have done here, for recruitment, the numbers at age
and catchability, because the simulation generates those deviation arrays itself and the arrows can therefore
name them directly. However, and movement are different, since we supply growth and movement values directly to the
simulation rather than having it generate them, so there is no deviation array for the arrows to name. For
those two processes we instead draw the dsem series ourselves and apply it to the parameters before they are
supplied, which is what we do below using a von Bertalanffy K example, letting a slowly varying environment move the
growth rate. Movement therefore follows the same idea: specify the movement parameters, then apply dsem linked
deviations to those parameters.

```{r}
# specify dsem model
arrows <- c("env -> env, 1, rho_env",
            "env <-> env, 0, sd_env",
            "env -> K, 0, b_env",
            "K <-> K, 0, sd_K")

truth <- c(rho_env = 0.7, sd_env = 1, b_env = 0.2, sd_K = 0.05)

# get dsem model
dm <- read_dsem_arrows(arrows, variables = c("env", "K"))
av <- SPoRC:::get_dsem_arrow_values(dsem_beta = unname(truth[dm$beta_names]),
                                    ln_dsem_sd = unname(log(truth[dm$ln_sd_names])),
                                    dsem_model = dm)

# draw desem
set.seed(11)
grid <- SPoRC:::draw_dsem_recursive(dm, as.numeric(av), matrix(0, sim_list$n_yrs, 2), n_sims = sim_list$n_sims) 

# input devs into growth pars and derive 
y <- 1
sim <- 1
Linf <- 100
K_y <- 0.18 * exp(grid[,2,])  
len <- Linf * (1 - exp(-K_y[y,sim] * (1:sim_list$n_ages)))
```

After simulating the operating model (OM), we set up the estimation model (EM) in `SPoRC`. In general, EM
settings are identical to the OM, except for fishery selectivity. The function takes a `spec`, which says
which arrows the EM believes and which processes it declares; everything else stays the same between fits, so
that the differences we report below come from the arrows alone.

One default worth specifying through the setup: `Setup_Mod_DSEM` holds every covariate's mean at zero,
since `dsem_mu_spec` defaults to `"zero"`. That is the convention the operating model draws deviations under, 
and it keeps the mean from trading off against the path
coefficient when a covariate is seen in only a few years. It does take the covariate as centered: one in its
own units should be centered first, or given `dsem_mu_spec = "fix"`, which holds it at its observed mean
instead. A warning names any covariate whose observed mean sits further from zero than its own spread.
By contrast, note that `dsem_mu_spec = "est"` estimates the means.

```{r}
setup_em <- function(sim_obj, sim, spec) {

  # Extract simulation data for current year and replicate
  sim_data <- simulation_data_to_SPoRC(sim_env = sim_obj, y = sim_obj$n_years, sim = sim)

  # Setup model dimensions
  input_list <- Setup_Mod_Dim(
    years = 1:sim_obj$n_years,
    ages = 1:sim_obj$n_ages,
    lens = sim_obj$n_lens,
    n_regions = sim_obj$n_regions,
    n_sexes = sim_obj$n_sexes,
    n_fish_fleets = sim_obj$n_fish_fleets,
    n_srv_fleets = sim_obj$n_srv_fleets,
    n_pop = sim_obj$n_pop,
    bias_correct_pe = "rec", # the same switch the operating model was given
    verbose = FALSE
  )

  # Recruitment setup
  input_list <- Setup_Mod_Rec(
    input_list = input_list,
    do_rec_bias_ramp = 0, # not doing bias ramp
    sigmaR_switch = 1, # when to switch from early to late sigmaR (switch in first year)
    ln_sigmaR = array(log(1), dim = c(2, input_list$data$n_pop, input_list$data$n_regions)), # 2 values for early and late sigma
    rec_model = "mean_rec",
    sigmaR_spec = if(spec$rec) "fix" else "fix_early_est_late", # a declared recruitment leaves sigmaR unread
    init_age_strc = "matrix", 
    equil_init_age_strc = "stoch_all", # estimating all intial age deviations
    ln_global_R0 = log(5),
    RecDevs_model = if(spec$rec) "dsem" else "iid" # the arrows own the deviations, or they stay iid
  )

  # the operating model started in equilibrium, so the initial age deviations are known rather than estimated
  input_list$par$ln_InitDevs[] <- 0
  input_list$map$ln_InitDevs <- factor(rep(NA, length(input_list$par$ln_InitDevs)))

  # Biological setup
  input_list <- Setup_Mod_Biologicals(
    input_list = input_list,
    # Data inputs
    WAA = sim_data$WAA,
    MatAA = sim_data$MatAA,
    WAA_fish = sim_data$WAA_fish,
    WAA_srv = sim_data$WAA_srv,
    fit_lengths = 0, # not fitting lengths
    AgeingError = sim_data$AgeingError,
    M_spec = "fix",     # fixing natural mortality
    Fixed_natmort = array(0.3, dim = c(input_list$data$n_pop, input_list$data$n_regions, length(input_list$data$years),  length(input_list$data$ages), input_list$data$n_sexes))
  )

  # Movement and tagging
  input_list <- Setup_Mod_Tagging(input_list = input_list, use_conv_fish_tagging = 0)
  input_list <- Setup_Mod_Movement(
    input_list = input_list,
    use_fixed_movement = 1,
    Fixed_Movement = NA,
    do_recruits_move = 0
  )

  # Fishery catch & fishing mortality
  input_list <- Setup_Mod_Catch_and_F(
    input_list = input_list,
    # Data inputs
    ObsCatch = sim_data$ObsCatch,
    UseCatch = sim_data$UseCatch,
    # Model options
    Use_F_pen = 1,
    sigmaC_spec = "fix",
    # Fixing sigma C and F
    ln_sigmaC = sim_data$ln_sigmaC,
    ln_sigmaF = array(log(1), dim = c(input_list$data$n_regions, input_list$data$n_seas, input_list$data$n_fish_fleets))
    )

  # Survey selectivity and catchability
  input_list <- Setup_Mod_FishIdx_and_Comps(
    input_list = input_list,
    # Data inputs
    ObsFishIdx = sim_data$ObsFishIdx,
    ObsFishIdx_SE = sim_data$ObsFishIdx_SE,
    UseFishIdx = sim_data$UseFishIdx,
    ObsFishAgeComps = sim_data$ObsFishAgeComps,
    ObsFishLenComps = sim_data$ObsFishLenComps,
    UseFishAgeComps = sim_data$UseFishAgeComps,
    UseFishLenComps = sim_data$UseFishLenComps,
    ISS_FishAgeComps = sim_data$ISS_FishAgeComps,
    ISS_FishLenComps = sim_data$ISS_FishLenComps,
    # Model options
    fish_idx_type = c("biom"),
    FishAgeComps_LikeType = c("Multinomial"),
    FishLenComps_LikeType = c("none"),
    FishAgeComps_Type = c("agg_Year_1-terminal_Fleet_1"),
    FishLenComps_Type = c("none_Year_1-terminal_Fleet_1")
  )

  # Survey indices and compositions
  input_list <- Setup_Mod_SrvIdx_and_Comps(
    input_list = input_list,
    # Data inputs
    ObsSrvIdx = sim_data$ObsSrvIdx,
    ObsSrvIdx_SE = sim_data$ObsSrvIdx_SE,
    UseSrvIdx = sim_data$UseSrvIdx,
    ObsSrvAgeComps = sim_data$ObsSrvAgeComps,
    ObsSrvLenComps = sim_data$ObsSrvLenComps,
    UseSrvAgeComps = sim_data$UseSrvAgeComps,
    UseSrvLenComps = sim_data$UseSrvLenComps,
    ISS_SrvAgeComps = sim_data$ISS_SrvAgeComps,
    ISS_SrvLenComps = sim_data$ISS_SrvLenComps,
    # Model options
    srv_idx_type = c("biom"),
    SrvAgeComps_LikeType = c("Multinomial"),
    SrvLenComps_LikeType = c("none"),
    SrvAgeComps_Type = c("agg_Year_1-terminal_Fleet_1"),
    SrvLenComps_Type = c("none_Year_1-terminal_Fleet_1")
  )


  # Fishery selectivity and catchability
  input_list <- Setup_Mod_Fishsel_and_Q(
    input_list = input_list,
    # Model options
    fish_sel_model = c("logist2_Fleet_1"), # fishery selex model 
    fish_fixed_sel_pars_spec = c("est_all"), # whether to estiamte all fixed effects for fishery selectivity
    fish_q_spec = "est_all" # estimate fishery q
  )

  # Survey selectivity and catchability
  input_list <- Setup_Mod_Srvsel_and_Q(
    input_list = input_list,
    # Model options
    srv_sel_model = c("logist2_Fleet_1"), # survey selectivity form
    srv_fixed_sel_pars_spec = c("est_all"), # whether to estimate all fixed effects for survey selectivity
    srv_q_spec = c("est_all"),  # whether to estiamte all fixed effects for survey catchability
    srv_q_model = spec$q, # "dsem" hands the deviations to the arrows, "none" leaves catchability constant
    sigma_srv_q_spec = if(spec$q == "dsem") "fix" else "est_all" # a declared catchability leaves its sigma unread
  )

  # Setup DSEM model here
  if(!is.null(spec$arrows)) {

    # the covariate used here, specific to each sim given obs error
    cov_df <- data.frame(year = 1:sim_obj$n_years)
    for(cov in spec$covs) cov_df[[cov]] <- sim_obj$dsem_cov_obs_sim[, match(cov, cov_names), sim]

    # setup dsem
    input_list <- Setup_Mod_DSEM(
      input_list,
      dsem_data = cov_df,
      dsem_arrows = spec$arrows,
      dsem_processes = c(if(spec$rec) "rec", if(spec$q == "dsem") "srv_q"),
      dsem_family = setNames(rep("normal", length(spec$covs)), spec$covs) # estimat obs error too
    )

  }

  # Data weighting
  input_list <- Setup_Mod_Weighting(
    input_list = input_list,
    Wt_Catch = 1,
    Wt_FishIdx = 1,
    Wt_SrvIdx = 1,
    Wt_Rec = 1,
    Wt_F = 1,
    Wt_Tagging = 0,
    Wt_FishAgeComps = array(1, dim = c(input_list$data$n_regions, length(input_list$data$years), input_list$data$n_seas,
                                       input_list$data$n_sexes, input_list$data$n_fish_fleets)),
    Wt_FishLenComps = array(1, dim = c(input_list$data$n_regions, length(input_list$data$years), input_list$data$n_seas,
                                       input_list$data$n_sexes, input_list$data$n_fish_fleets)),
    Wt_SrvAgeComps = array(1, dim = c(input_list$data$n_regions,length(input_list$data$years), input_list$data$n_seas,
                                      input_list$data$n_sexes, input_list$data$n_srv_fleets)),
    Wt_SrvLenComps = array(0, dim = c(input_list$data$n_regions, length(input_list$data$years), input_list$data$n_seas,
                                      input_list$data$n_sexes, input_list$data$n_srv_fleets))
  )

  return(input_list)

}
```

The same data with two different models, and the only difference between the two fits is how survey catchability is specified: both
use the same catch, indices and compositions, and both see the same four observed covariates.  In `dsem_q`, catchability is additionally given to the arrows, so that its deviations are explained by availability and by recruitment, whereas in `const_q` catchability is constant over time.

```{r}
# dsem map
shared <- c("pdo -> pdo, 1, rho_pdo",
            "pdo <-> pdo, 0, sd_pdo",
            "temp -> temp, 1, rho_temp",
            "pdo -> temp, 0, b_pdo_temp",
            "temp <-> temp, 0, sd_temp",
            "pdo -> prey, 0, b_pdo_prey",
            "prey <-> prey, 0, sd_prey",
            "temp -> rec, 0, b_temp_rec",
            "prey -> rec, 0, b_prey_rec",
            "rec <-> rec, 0, sd_rec",
            "temp -> cond, 0, b_temp_cond",
            "cond <-> cond, 0, sd_cond")

# specifications
specs <- list(
  # the map the data were drawn under
  dsem_q = list(covs = cov_names, rec = TRUE, q = "dsem",
                arrows = c(shared,
                           "cond -> srv_q, 0, b_cond_q",
                           "rec -> srv_q, 0, b_rec_q",
                           "srv_q <-> srv_q, 0, sd_q")),
  # the same map, with catchability constant
  const_q = list(covs = cov_names, rec = TRUE, q = "none", arrows = shared)
)
```

```{r, message = FALSE, warning = FALSE}
n_fit <- 1
n_yrs_em <- sim_obj$n_years

# storage stuff
res <- NULL
ssb_ts <- array(NA_real_, dim = c(n_yrs_em, length(specs), n_fit), dimnames = list(NULL, names(specs), NULL))
q_ts <- ssb_ts

for(sim in 1:n_fit) {
  for(name in names(specs)) {

    # setup EM
    spec <- specs[[name]]
    input_list <- setup_em(sim_obj, sim, spec)

    # figure out what is random
    has_q_dev <- !is.null(input_list$map$ln_srv_q_devs) && any(!is.na(input_list$map$ln_srv_q_devs))
    rand <- c("ln_RecDevs", if(has_q_dev) "ln_srv_q_devs", "dsem_x")
    model <- fit_model(input_list$data, input_list$par, input_list$map, random = rand, newton_loops = 1, silent = F)

    # get paraemters and report
    par_list <- model$env$parList()
    ssb_ts[, name, sim] <- model$rep$SSB[1,1,]
    q_ts[, name, sim] <- model$rep$srv_q[1,,1]

    # bind stuff together
    res <- rbind(res, data.frame(
      sim = sim,
      spec = name,
      max_grad = max(abs(model$gr(model$env$last.par.best[-model$env$random]))),
      betas = paste(round(par_list$dsem_beta, 2), collapse = " "),
      sd_q = if(spec$q == "dsem") exp(par_list$ln_dsem_sd[match("sd_q", input_list$data$dsem_model$ln_sd_names)]) else NA, # truth 0.2
      ssb_mean_abs_err = mean(abs(model$rep$SSB[1,1,] / sim_obj$SSB[1,1,,sim] - 1)),
      q_mean_abs_err = mean(abs(model$rep$srv_q[1,,1] / sim_obj$srv_q[1,,1,sim] - 1)),
      rec_dev_rmse = sqrt(mean((par_list$ln_RecDevs[1,1,] - sim_obj$ln_RecDevs[1,1,,sim])^2))
    ))

  } # end spec loop
} # end sim loop
```

As shown below, mis-specifying the catchability process (i.e., constant q) leads to lard biases, while correctly specifying the casual map results in negiligible bias across model replicates. 

```{r, fig.width = 11, fig.height = 9}
par(mfrow = c(2, 1), mar = c(4.5, 4.8, 1.5, 1), cex.lab = 1.3, cex.axis = 1.2)
cols <- c(dsem_q = "#000000", const_q = "#CC79A7")
yrs <- 1:n_yrs_em

# the error of every replicate comapred acgainst its own operating model values, then the median over them
median_err <- function(ts, truth) {
  err <- sweep(ts, c(1, 3), truth, "/") - 1 # [year, spec, replicate]
  apply(err, c(1, 2), median, na.rm = TRUE)
}

ssb_err <- median_err(ssb_ts, sim_obj$SSB[1,1,,1:n_fit])
q_err <- median_err(q_ts, sim_obj$srv_q[1,,1,1:n_fit])

plot(range(yrs), range(ssb_err, na.rm = TRUE), type = 'n', xlab = 'Year', ylab = 'Median relative error in SSB')
for(name in names(specs)) lines(yrs, ssb_err[, name], col = cols[name], lwd = 2.5)
abline(h = 0, lty = 2)
legend('topright', names(specs), col = cols, lwd = 2.5, bty = 'n', cex = 1.1)

plot(range(yrs), range(q_err, na.rm = TRUE), type = 'n', xlab = 'Year', ylab = 'Median relative error in survey q')
for(name in names(specs)) lines(yrs, q_err[, name], col = cols[name], lwd = 2.5)
abline(h = 0, lty = 2)
```

## Non-stationary dynamics

Lastly, we demonstrate how to implement non-stationary dynamics using dsem. There are several approaches availiable for doing so, ranging from: 1) the moderated variable approach, in which a path coefficient is able to change over time; 2) the same substitution on an sd specification, so that a process error variance changes from year to year; and 3) derived series, which have no process error of their own and are therefore whatever the arrows pointing into them give them, and which can be chained through moderated arrows to write a nonlinear function of a series, such as its square, or an interaction between two covariates. Thorson and Kristensen (2026) introduce all three as one extension of the arrow and lag notation, which they term path-lag-slope notation, and show that the resulting moderated dsem can express non-stationary effects, a nonlinear response, and a statistical interaction while still interpolating the years in which a covariate was not measured. In the following, we first describe what naming an arrow after a series does to the density, and then fit an environmental effect on Dusky Rockfish recruitment that weakens over time, the same effect written as two time blocks, a recruitment variance that changes over time, and a dome shaped recruitment response to a covariate.

### Naming an arrow after a series

Consider an environmental index that increased recruitment appreciably through the 1980s and 1990s but appears to add little thereafter, perhaps because the mechanism it acted through, such as prey availability at first feeding, has itself changed. A single path coefficient estimated over the whole time-series then likely averages the two periods and can report an effect of nearly zero, even though the covariate was informative for the first half of the assessment. As such, we would like the coefficient itself to be a quantity that changes over time.

In particular, when the third field of an arrow names a series rather than a parameter, `SPoRC` reads that arrow's coefficient off that series, in the year the arrow points at, instead of estimating one number for it. Writing that out for a covariate acting on recruitment gives:

$$
\begin{aligned}
rec_t &= b_t\, env_t + \varepsilon^{rec}_t \\
b_t &= \mu_b + \rho_b\,(b_{t-1} - \mu_b) + \varepsilon^{b}_t
\end{aligned}
$$

where $env_t$ is the covariate in year $t$ and $rec_t$ the recruitment deviation in year $t$, both on the scale the grid works on (log scale for recruitment); $b_t$ is the effect of the covariate on recruitment in year $t$, which is now a series of its own and therefore a column of the grid; $\mu_b$ is the mean of that series, which is the one effect a stationary model would have estimated; $\rho_b$ is its first order autocorrelation, which sets how fast the effect may move; and $\varepsilon^{b}_t$ and $\varepsilon^{rec}_t$ are the innovations of the two series, whose standard deviations are named on their own `<->` lines. Note that $b_t$ is a series like any other, so the same lines that would make a covariate autoregressive make the effect autoregressive.

```{r mod_arrows}
mod_arrows <- c("env -> rec, 0, b_env",       # the coefficient is the b_env series, year by year
                "b_env -> b_env, 1, rho_b",   # so the effect follows itself from last year (rho has an ar1 series)
                "b_env <-> b_env, 0, sd_b",   # and moves this far each year
                "env <-> env, 0, sd_env",
                "rec <-> rec, 0, sd_rec")

mod_m <- read_dsem_arrows(mod_arrows, variables = c("env", "b_env", "rec"))
mod_m$beta_names # note that benv becomes a series instead of a fixed effect parameter
```

To see where the effect lands in the density, we read $I - B$ for a three year grid, with the effect taking the values 0.2, 0.45 and 0.35 in those years.

```{r mod_matrix}
mod_cells <- SPoRC:::get_dsem_cells(mod_m, n_grid_yrs = 3)
mod_grid <- cbind(env = c(0.27, -0.63, 0.87), b_env = c(0.20, 0.45, 0.35), rec = c(0.02, 0.22, -0.79))
mod_parts <- SPoRC:::get_dsem_matrices(dsem_beta = c(rho_b = 0.8),
                                       ln_dsem_sd = log(c(sd_b = 0.15, sd_env = 1, sd_rec = 0.6)),
                                       dsem_model = mod_m, dsem_cells = mod_cells, x_grid = mod_grid)
as.matrix(mod_parts$IminusB) # cells run year within series: env, then b_env, then rec
```

Rows 7 to 9 are the recruitment cells, and they hold minus the effect of that year in columns 1 to 3, which are the covariate's cells, whereas the block where those same rows meet columns 4 to 6, the effect's own cells, is empty. Note that this is what separates a moderated arrow from an ordinary path out of `b_env`: the latter would say that recruitment is higher in years when the effect is larger, whatever the covariate did that year, while the former says that recruitment responds to the covariate and that the strength of the response is `b_env`. Rows 4 to 6 are unchanged by the moderation, and hold the autocorrelation of the effect on itself.

Intuitively, the `b_env` column therefore does two jobs at once. It is a state with an equation of its own, which keeps it near where the neighboring years put it, and it is the coefficient in the recruitment equation, which pulls it toward whatever would explain that year's deviation. The sd line on `b_env` is what sets the balance between the two: as it shrinks, the effect stops moving and the model returns to a single coefficient, and as it grows, the effect is free to chase each year's deviation in turn. A moderating series is generally not measured, and it then enters as a column of `dsem_data` that is `NA` in every year, so that each of its cells is a random effect. Lastly, the series has to sit upstream of the arrow it moderates, and the reader refuses a moderator whose value that same arrow sets in the same year, since the density would no longer be exact.

### An effect that weakens over time

We use a contrived covariate built against the packaged Dusky Rockfish deviations, which tracks recruitment before 2000 and is noise thereafter, and which begins in 1990 as a survey based index might.

```{r fade_cov}
years <- base$data$years
n_yrs <- length(years)
eps_std <- as.vector(scale(dusky_rtmb_model$rep$ln_RecDevs[1,1,])) # the packaged deviations, standardized

set.seed(17)
env_fade <- 0.7 * eps_std * (years < 2000) + rnorm(n_yrs, 0, 0.7)
env_fade[years < 1990] <- NA # the index begins in 1990, so the earlier years are states
env_fade <- as.vector(scale(env_fade))
fade_df <- data.frame(year = years, env = env_fade)
c(observed_years = sum(!is.na(env_fade)), cor_before_2000 = cor(env_fade[years %in% 1990:1999], eps_std[years %in% 1990:1999]))
```

The first fit estimates one coefficient for the whole series, which is the stationary model, and the second names that arrow after a series so the effect changes by year. In the second, the autocorrelation and the innovation sd of the effect are kept fixed at 0.9 and 0.15 by writing `NA` in the name field and the value in the fourth, for reasons we return to below, and the effect's mean is estimated through `dsem_mu_spec`, since it is the average effect and there is no reason to hold it at zero.

```{r fade_fits, message = FALSE, warning = FALSE}
dsem_pars <- function(fit, input_list) {
  fixed_names <- names(fit$sd_rep$par.fixed)
  est <- fit$sd_rep$par.fixed[fixed_names == "dsem_beta"]
  se <- sqrt(diag(fit$sd_rep$cov.fixed))[fixed_names == "dsem_beta"]
  names(est) <- names(se) <- input_list$data$dsem_model$beta_names
  round(rbind(est = est, se = se), 3)
}

# one coefficient for the whole series
flat_arrows <- c("env -> env, 1, rho_env",
                 "env <-> env, 0, sd_env",
                 "env -> rec, 0, b_env",
                 "rec <-> rec, 0, sd_rec")

flat_il <- Setup_Mod_DSEM(dsem_base, dsem_arrows = flat_arrows, dsem_data = fade_df)
flat_fit <- fit_model(flat_il$data, flat_il$par, flat_il$map, random = c("ln_RecDevs", "dsem_x"),
                      newton_loops = 2, silent = TRUE)
flat_fit$sd_rep <- sdreport(flat_fit)
dsem_pars(flat_fit, flat_il)

# the same effect, as a series of its own
slope_arrows <- c("env -> env, 1, rho_env",
                  "env <-> env, 0, sd_env",
                  "env -> rec, 0, b_env",         # the effect of the covariate is the b_env series
                  "b_env -> b_env, 1, NA, 0.9",   # how fast it may move, kept fixed
                  "b_env <-> b_env, 0, NA, 0.15", # how far it may move each year, kept fixed
                  "rec <-> rec, 0, sd_rec")

slope_il <- Setup_Mod_DSEM(dsem_base,
                           dsem_arrows = slope_arrows,
                           dsem_data = cbind(fade_df, b_env = NA_real_), # unmeasured, so every year is a state
                           dsem_mu_spec = "b_env")

# start the effect flat at the single coefficient, as any state space fit is started
b_flat <- flat_fit$env$parList()$dsem_beta[match("b_env", flat_il$data$dsem_model$beta_names)]
slope_il$par$dsem_x[, match("b_env", slope_il$data$dsem_model$variables)] <- b_flat
slope_il$par$dsem_mu[match("b_env", slope_il$data$dsem_model$variables)] <- b_flat

slope_fit <- fit_model(slope_il$data, slope_il$par, slope_il$map, random = c("ln_RecDevs", "dsem_x"),
                       newton_loops = 2, silent = TRUE)
slope_fit$sd_rep <- sdreport(slope_fit)
dsem_pars(slope_fit, slope_il)

# the effect by year, and its standard error from the random effect part of the report
b_yr <- slope_fit$rep$dsem_x_grid[, match("b_env", slope_il$data$dsem_model$variables)]
in_b_col <- rep(slope_il$data$dsem_model$variables == "b_env", each = n_yrs) # the grid runs year within series
estimated <- !is.na(as.integer(slope_il$map$dsem_x))
is_x <- names(slope_fit$sd_rep$par.random) == "dsem_x"
b_se <- sqrt(slope_fit$sd_rep$diag.cov.random[is_x])[in_b_col[estimated]]

c(one_coefficient = flat_fit$optim$objective, effect_by_year = slope_fit$optim$objective)
c(mean_before_2000 = mean(b_yr[years < 2000]), mean_after_2000 = mean(b_yr[years >= 2000]))
```

Estimated as one number, the effect is 0.037 with a standard error of 0.154, which is to say that the covariate does nothing. Estimated as a series, it averages 0.464 before 2000 and -0.320 after it, and the marginal likelihood improves by 7.3 units on the same number of fixed parameters, since the single coefficient has been replaced by the mean of the series. The moderated form therefore recovers the structure the covariate was built with, while the stationary form reports the average of a positive effect and no effect and finds nothing.

```{r, fig.width = 11, fig.height = 4.5}
par(mfrow = c(1, 2), mar = c(5, 5.5, 1, 1), mgp = c(3.2, 0.9, 0), cex.lab = 1.4, cex.axis = 1.2, las = 1)

plot(years, env_fade, type = 'l', xlab = 'Year', ylab = 'Environmental covariate')
points(years, env_fade, pch = 16, cex = 0.8)

plot(years, b_yr, type = 'n', ylim = range(c(b_yr - 1.96 * b_se, b_yr + 1.96 * b_se)),
     xlab = 'Year', ylab = 'Effect on log recruitment')
polygon(c(years, rev(years)), c(b_yr - 1.96 * b_se, rev(b_yr + 1.96 * b_se)),
        col = '#00000022', border = NA)
lines(years, b_yr, lwd = 2.5)
abline(h = b_flat, lty = 2, lwd = 2)
legend('topright', c('Effect by year', 'One coefficient'), lwd = c(2.5, 2), lty = c(1, 2), bty = 'n', cex = 1.1)
```

Two points are worth making about this fit. Firstly, the effect's autocorrelation and innovation sd are kept fixed rather than estimated. The innovation sd says how far the effect may move each year and is identified only through the recruitment deviations, of which there are 48 here, and our attempts to estimate it returned a non finite likelihood before the fit converged. Fixing it is therefore a choice about how much movement we are prepared to entertain, and a sensitivity analysis over that value is more informative than an estimate of it. Secondly, Thorson and Kristensen (2026) show by simulation that a moderated coefficient is shrunk toward a constant as the data become less informative, so a series that comes back nearly flat is evidence about the data as much as about the process, and the standard errors on the effect here, which average 0.33 against a difference of about 0.8 between the two periods, are consistent with that.

### The same effect as two time blocks

If the year in which the relationship changed is known beforehand, for instance because a survey was redesigned or a regime shift was identified elsewhere, then the effect can be written as two values rather than as a series. This needs the product of the covariate and an indicator that is one after the break, which is where a derived series comes in: an arrow moderated by the indicator writes that product, and giving the product an sd of zero leaves it as exactly the product rather than a series with process error of its own. The effect is then `b_pre` before the break and `b_pre + b_shift` after it, which is the statistical interaction Thorson and Kristensen (2026) write as an interaction variable.

```{r block_fit, message = FALSE, warning = FALSE}
block_df <- data.frame(year = years,
                       env = env_fade,
                       post = as.numeric(years >= 2000), # one after the break, zero before it
                       env_post = NA_real_)              # the product, which the arrows work out

block_arrows <- c("env -> env, 1, rho_env",
                  "env <-> env, 0, sd_env",
                  "post <-> post, 0, NA, 1",         # observed every year, so this line only adds a constant
                  "env -> env_post, 0, post",        # moderated by the indicator, so this is the product
                  "env_post <-> env_post, 0, NA, 0", # no process error, so the product is all there is
                  "env -> rec, 0, b_pre",
                  "env_post -> rec, 0, b_shift",
                  "rec <-> rec, 0, sd_rec")

block_il <- Setup_Mod_DSEM(dsem_base, dsem_arrows = block_arrows, dsem_data = block_df)
block_fit <- fit_model(block_il$data, block_il$par, block_il$map, random = c("ln_RecDevs", "dsem_x"),
                       newton_loops = 2, silent = TRUE)
block_fit$sd_rep <- sdreport(block_fit)
dsem_pars(block_fit, block_il)

# the product is the covariate times the indicator, which we can read back off the grid
block_grid <- block_fit$rep$dsem_x_grid
colnames(block_grid) <- block_il$data$dsem_model$variables
max(abs(block_grid[,'env_post'] - block_grid[,'post'] * block_grid[,'env']))

# and the effect after the break, with the covariance of the two coefficients
fixed_names <- names(block_fit$sd_rep$par.fixed)
block_beta <- block_fit$sd_rep$par.fixed[fixed_names == "dsem_beta"]
block_cov <- block_fit$sd_rep$cov.fixed[fixed_names == "dsem_beta", fixed_names == "dsem_beta"]
i_pre <- match("b_pre", block_il$data$dsem_model$beta_names)
i_shift <- match("b_shift", block_il$data$dsem_model$beta_names)
c(after_2000 = block_beta[i_pre] + block_beta[i_shift],
  se = sqrt(block_cov[i_pre,i_pre] + block_cov[i_shift,i_shift] + 2 * block_cov[i_pre,i_shift]))
```

The effect is 1.053 with a standard error of 0.231 before 2000 and -0.738 with a standard error of 0.199 after it. Note that these are appreciably larger in magnitude than the series in the previous fit, which is what we should expect: two blocks are estimated without any statement that the effect should move slowly, whereas the series is shrunk toward its own mean by its innovation sd, and the shrinkage is the reason the series reaches only 0.46 and -0.32. In general, the two forms answer slightly different questions, one asking how much the effect differs between two periods chosen in advance, and the other asking whether the effect moved at all without saying when.

The likelihoods of these two fits, however, cannot be compared. The indicator is a series of the grid and contributes its own density terms, so the two models' likelihoods differ by a constant that has nothing to do with recruitment. In general, a likelihood comparison between two dsem specifications is only meaningful when both hold the same set of observed series, as the two fits in the previous section did.

### A recruitment variance that changes over time

Recruitment variability is itself often thought to have changed, whether because the spawning stock became concentrated on fewer spawning areas or because the physical environment became more variable, and an sd line can be moderated in exactly the same way as a path. Here, the sd line for recruitment names a series rather than a parameter, and `mod_var_logscale = TRUE` reads that series as the log of the standard deviation, which keeps it positive whatever value the series takes.

```{r mod_sd_fit, message = FALSE, warning = FALSE}
sd_arrows <- c("ln_sd_rec -> ln_sd_rec, 1, NA, 0.9",   # the spread follows itself from last year
               "ln_sd_rec <-> ln_sd_rec, 0, NA, 0.35", # and moves this far each year, both kept fixed
               "rec <-> rec, 0, ln_sd_rec")            # the sd line names a series, so the spread moves with it

sd_il <- Setup_Mod_DSEM(pen_base,
                        dsem_arrows = sd_arrows,
                        dsem_data = data.frame(year = years, ln_sd_rec = NA_real_),
                        dsem_mu_spec = "ln_sd_rec", # the average spread, on the log scale
                        mod_var_logscale = TRUE)

sd_fit <- fit_model(sd_il$data, sd_il$par, sd_il$map, random = c("ln_RecDevs", "dsem_x"),
                    newton_loops = 2, silent = TRUE)
sd_fit$sd_rep <- sdreport(sd_fit)
sd_yr <- exp(sd_fit$rep$dsem_x_grid[, match("ln_sd_rec", sd_il$data$dsem_model$variables)])

# the same model with one spread for every year
one_sd_il <- Setup_Mod_DSEM(pen_base, dsem_arrows = "rec <-> rec, 0, sd_rec", dsem_data = NULL)
one_sd_fit <- fit_model(one_sd_il$data, one_sd_il$par, one_sd_il$map, random = "ln_RecDevs",
                        newton_loops = 2, silent = TRUE)

c(range_by_year = range(sd_yr), one_value = exp(as.vector(one_sd_fit$env$parList()$ln_dsem_sd)))
c(spread_by_year = sd_fit$optim$objective, one_spread = one_sd_fit$optim$objective)
```

A moderated sd changes the behavior of the recruitment bias correction, and `SPoRC` refuses the combination: there is no single variance to center the deviations on, least of all for the initial ages, which are born before the grid starts, so `bias_correct_pe = "none"` is required, which is why this fit is built on the penalized base rather than the one used above. The initial age deviations then read the series' settled marginal variance under the arrows, taken with the moderating series at its mean, and `ln_sigmaR` goes unread.

```{r, fig.width = 11, fig.height = 4.5}
par(mfrow = c(1, 2), mar = c(5, 5.5, 1, 1), mgp = c(3.2, 0.9, 0), cex.lab = 1.4, cex.axis = 1.2, las = 1)

plot(years, sd_yr, type = 'l', lwd = 2.5, ylim = c(0, max(sd_yr)),
     xlab = 'Year', ylab = 'Recruitment deviation sd')
abline(h = exp(as.vector(one_sd_fit$env$parList()$ln_dsem_sd)), lty = 2, lwd = 2)
legend('bottomleft', c('Spread by year', 'One spread'), lwd = c(2.5, 2), lty = c(1, 2), bty = 'n', cex = 1.1)

plot(years, sd_fit$rep$dsem_x_grid[, match("rec", sd_il$data$dsem_model$variables)], type = 'l', lwd = 2.5,
     xlab = 'Year', ylab = 'Recruitment deviation')
lines(years, one_sd_fit$rep$dsem_x_grid[, match("rec", one_sd_il$data$dsem_model$variables)], col = '#CC79A7', lwd = 2.5)
legend('topleft', c('Spread by year', 'One spread'), col = c('black', '#CC79A7'), lwd = 2.5, bty = 'n', cex = 1.1)
```

The estimated spread runs from 0.18 to 1.72 against a single value of 1.01, and yet the marginal likelihood is 6.6 units worse than the model with one spread, so these data do not support a variance that changes over time even though the series moved a great deal. In general, a moderated sd is weakly identified, because each year's spread is estimated from that year's deviation alone, and the last few years are the clearest case: the terminal deviations are barely informed by data, the density pulls them toward zero, and a small spread there makes that pull stronger still, which is what the two sets of deviations show from 2019 onward, where the moving spread returns values no larger than 0.013 in absolute terms against 0.13 to 0.34 under a single spread. As such, we recommend keeping the movement of a moderated sd tightly constrained, comparing against a single value as we have done here, and treating a collapse in the terminal years as an artifact rather than a finding.

### A dome shaped recruitment response

Recruitment is frequently thought to respond to temperature in a dome shaped way, rising to some intermediate optimum and falling away on either side, which is a quadratic effect of the covariate rather than a changing one. Moderating an arrow by the same series it comes from writes the square: the coefficient on the arrow out of the covariate is the covariate, so the series it points into is the covariate times itself. Giving that series an sd of zero makes it derived, so the square is exactly the square even in years where the covariate was not measured and the model is estimating it, which is the property Thorson and Kristensen (2026) build their Lotka-Volterra and plankton examples on, and the reason a basis expansion supplied as data cannot be used when a covariate has missing years.

One further step is needed before the square is useful in an assessment. The square of a standardized covariate has a mean of about one, so it acts partly as a level on the recruitment deviations, and since the deviations have no intercept of their own, that level is estimated against mean recruitment instead. In our fits the uncentered form returned a quadratic coefficient of 0.45, the wrong sign for the dome the covariate was built with, with mean recruitment moving to absorb the level. We therefore subtract the mean of the square, which takes a series that is one in every year, from which a fixed arrow of minus that mean is an intercept.

```{r dome_fit, message = FALSE, warning = FALSE}
set.seed(31)
dome_raw <- sqrt(pmax(0, (stats::quantile(eps_std, 0.95) - eps_std) / 2)) * sample(c(-1, 1), n_yrs, TRUE)
env_dome <- as.vector(scale(dome_raw + rnorm(n_yrs, 0, 0.25))) # recruitment is highest at intermediate values
sq_mean <- mean(env_dome^2)

dome_df <- data.frame(year = years,
                      one = 1,             # a constant series, so an arrow out of it is an intercept
                      env = env_dome,
                      env_sq = NA_real_,   # the square, which the arrows work out
                      env_ctr = NA_real_)  # and the square less its mean

dome_arrows <- c("one <-> one, 0, NA, 1",
                 "env -> env, 1, rho_env",
                 "env <-> env, 0, sd_env",
                 "env -> env_sq, 0, env",     # moderated by itself, so this is the covariate squared
                 "env_sq <-> env_sq, 0, NA, 0",
                 "env_sq -> env_ctr, 0, NA, 1",
                 paste0("one -> env_ctr, 0, NA, ", -round(sq_mean, 4)), # minus the mean of the square
                 "env_ctr <-> env_ctr, 0, NA, 0",
                 "env -> rec, 0, b_lin",
                 "env_ctr -> rec, 0, b_quad",
                 "rec <-> rec, 0, sd_rec")

cat(dome_arrows, sep = "\n") # the lines the setup reads

dome_il <- Setup_Mod_DSEM(dsem_base, dsem_arrows = dome_arrows, dsem_data = dome_df)
dome_fit <- fit_model(dome_il$data, dome_il$par, dome_il$map, random = c("ln_RecDevs", "dsem_x"),
                      newton_loops = 2, silent = TRUE)
dome_fit$sd_rep <- sdreport(dome_fit)
dsem_pars(dome_fit, dome_il)

# the derived series are the square and the centered square, which we read back off the grid
dome_grid <- dome_fit$rep$dsem_x_grid
colnames(dome_grid) <- dome_il$data$dsem_model$variables
max(abs(dome_grid[,'env_ctr'] - (dome_grid[,'env']^2 - sq_mean)))

# the covariate value at which recruitment is highest
dome_beta <- dsem_pars(dome_fit, dome_il)["est",]
-dome_beta[["b_lin"]] / (2 * dome_beta[["b_quad"]])
```

The linear coefficient is 0.301 with a standard error of 0.188 and the quadratic coefficient is -1.899 with a standard error of 0.302, so the response is dome shaped, with the highest recruitment at a covariate value of 0.08, near the middle of the covariate's range.

```{r, fig.width = 7, fig.height = 5}
par(mar = c(5, 5.5, 1, 1), mgp = c(3.2, 0.9, 0), cex.lab = 1.4, cex.axis = 1.2, las = 1)

env_seq <- seq(min(env_dome), max(env_dome), length.out = 100)
plot(env_dome, dome_grid[,'rec'], pch = 16, xlab = 'Environmental covariate', ylab = 'Recruitment deviation')
lines(env_seq, dome_beta[["b_lin"]] * env_seq + dome_beta[["b_quad"]] * (env_seq^2 - sq_mean), lwd = 2.5)
```

Note that the fitted deviation in the year with the most extreme covariate value, 2000, is -6.2 against -0.05 in the packaged fit, and that the deviations as a whole now run from -6.2 to 2.1 against -0.5 to 1.9. A quadratic is unbounded, so a covariate value out in the tail implies a recruitment far outside anything the assessment has seen, and a year class that the data inform weakly will follow that implication rather than resist it. As such, we recommend reading a dome shaped response only over the range in which the covariate is dense, and comparing the fitted deviations against those of a model without the covariate before believing an extreme year.

Chaining further derived series raises further powers, and a third order series in the log of a variable is the Taylor approximation to the exponential that Thorson and Kristensen (2026) use to write Lotka-Volterra predator prey dynamics as arrows, which suggests that mechanistic relationships of this kind could be written against assessment process errors in the same way. Two cautions apply from our own attempts, however. Firstly, a product of a series with itself makes the joint density non-quadratic in the states, so the inner problem behind the Laplace approximation is no longer a simple least squares one, and with the covariate partly unmeasured we found these fits fragile: some converged cleanly and others stalled with a large final gradient or returned a non finite likelihood, depending on which years were missing. As such, we recommend fitting the fully measured case first, and treating a nonlinear term on an unmeasured covariate as something to be checked rather than assumed. Secondly, the total effect of a change in one series is no longer read off a single path matrix, since a moderated path matrix depends on the values of the series themselves, and Thorson and Kristensen (2026) give the total effect as a sum over successive path matrices, each updated with the effects of the last, rather than as the usual inverse.


## References

Champagnat, J., Monnahan, C. C., Sullivan, J. Y., Thorson, J. T., Shotwell, S. K.,
Rogers, L., and Punt, A. E. 2026. Causal models as a scientific framework for
next-generation ecosystem and climate-linked stock assessments. Fish and Fisheries,
27: 942–959.

Thorson, J. T., Andrews, A. G., Essington, T. E., and Large, S. I. 2024.
Dynamic structural equation models synthesize ecosystem dynamics constrained by
ecological mechanisms. Methods in Ecology and Evolution, 15: 744–755.

Thorson, J.T., Kristensen, K., 2026. Ecological examples of nonstationarity, 
nonlinearity and statistical interactions in dynamic structural equation models. Methods in Ecology and Evolution
17, 2510–2523. https://doi.org/10.1111/2041-210x.70334
