---
title: "Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document is for people editing the `SPoRC` codebase: contributors, successors, future
you. The goal is to keep the package navigable as it grows, especially as spatial model
development adds more moving parts. Update it whenever the pipeline below changes shape (a
new `Setup_Mod_*` stage, a new section of the objective function, a new post fit
diagnostic). A stale architecture document should be treated as a bug.

## Mental model

`SPoRC` runs in three phases, in this order.

1. **Setup.** A chain of R functions builds up three plain lists: `data`, `parameters` and
   `mapping`. No likelihood is evaluated yet. This phase is data wrangling and input
   validation only.
2. **Fit.** `data` and `parameters` go to `RTMB::MakeADFun` wrapped around a single
   objective function, `SPoRC_rtmb()`, which is optimized with `nlminb` plus Newton
   refinement.
3. **Post fit.** The fitted object (`obj$rep`, `obj$sdrep`, `obj$optim`) is passed to
   diagnostics, reference points, plotting and resampling routines.

```
Setup_Mod_Dim()
   -> Setup_Mod_Rec()
   -> Setup_Mod_Biologicals()
   -> Setup_Mod_Movement()
   -> Setup_Mod_Tagging()
   -> Setup_Mod_Catch_and_F()
   -> Setup_Mod_FishIdx_and_Comps()   [+ Setup_Mod_Discard_Comps()]
   -> Setup_Mod_Fishsel_and_Q()       [+ Setup_Mod_Retsel()]
   -> Setup_Mod_SrvIdx_and_Comps()
   -> Setup_Mod_Srvsel_and_Q()
   -> Setup_Mod_Weighting()
        |
        v  input_list$data, input_list$par, input_list$map
   fit_model(data, parameters, mapping)
        |  RTMB::MakeADFun(cmb(SPoRC_rtmb, data), parameters, map = mapping)
        |  nlminb() + Newton steps
        v
   obj  (obj$rep, obj$sdrep, obj$optim)
        |
        +--> Get_Reference_Points()          SPR and MSY
        +--> get_idx_fits() / get_osa()      fit extraction and OSA residuals
        +--> plot_* / get_*_plot             figures and tables
        +--> do_retrospective(), do_jitter(), do_likelihood_profile(), run_francis()
        +--> Do_Population_Projection(), condition_closed_loop_simulations()
```

### The `input_list` accumulator

Every `Setup_Mod_*` function takes an `input_list` (initialized by `Setup_Mod_Dim()` as
`list(data = list(), par = list(), map = list())`) and returns it with more keys filled in.
This is the central structure of the package: setup functions are pipeline stages that
thread one growing list through each other, not independent builders.

Order matters. Later stages read dimensions (`input_list$data$n_regions` and friends) set
by earlier ones, and default arguments frequently reference `input_list$data$...` in their
own definitions. See `Setup_Mod_Rec()`'s defaults for an example.

At the end of the chain `input_list$data`, `input_list$par` and `input_list$map` are
unpacked and passed to `fit_model()`. The `Setup_Sim_*` functions live in the same files as
their `Setup_Mod_*` counterparts and follow the identical accumulator pattern, but build
inputs for the operating model instead of for fitting.

### The objective function

`R/model_objective.R` holds the one RTMB objective function for every model configuration
the package supports. There is no per model branching at the file level. The objective
always runs end to end and is regulated by the switches set during setup, such as
`rec_model`, `move_type`, and the likelihood and OSA flags.

It moves through, in order: parameter transforms (movement, natural mortality,
selectivity), then mortality, recruitment, initial age structure, population projection,
the observation models (fishery, survey, tagging), the likelihood components (retained and
discarded catch, indices, compositions, tags), and finally priors and penalties.

Each section hands off to a module in one of the other `model_*.R` files, so the objective reads as an order of operations rather than as the arithmetic itself. If you are adding a new data source or process, this is where its likelihood contribution gets wired in, and the matching `Setup_Mod_*` function is
where its data, parameters and mapping get prepared beforehand.

Within `model_objective.R` the `## Section` and `### Subsection` comment banner style is
used consistently. Keep using it if you extend the file, since it is the only navigation
aid inside a function of this size.

Note that `R/model_transition.R` is worth knowing about before anything else. It computes what happens to a vector of abundance over one season, given movement, mortality,
season duration and a movement timing. Eleven other files go through it, including the
objective function's dynamics, the reference point solvers and the operating model.

## Naming conventions

### Files

The prefix on a file name says which phase it belongs to.

| Prefix | Phase |
|---|---|
| `setup_` | Builds `data`, `parameters` and `mapping`, one process or data source per file |
| `model_` | The objective function and the modules it delegates to |
| `refpts_`, `projection` | Reference points and forward projection off a fitted model |
| `sim_` | The operating model: self testing and closed loop simulation |
| `diag_` | Post fit diagnostics and residuals |
| `plot_` | Figures and summary tables |
| `utils_` | Helpers shared across phases |
| `data_` | Roxygen documentation for the bundled datasets |

### Functions

| Prefix | Role |
|---|---|
| `Setup_Mod_*` | Builds inputs for a fitted object |
| `Setup_Sim_*` | Same, for the operating model. Lives beside its `Setup_Mod_*` counterpart |
| `do_*_mapping` | Internal helper that builds the RTMB `map` for one parameter block. This is how parameters get fixed, shared across blocks, or turned into random effect deviations |
| `Get_*` | Pulls a derived quantity out of set up data, parameters or a fitted object |
| `do_*` (top level) | A post fit procedure that refits or perturbs a fitted model |
| `get_*_plot`, `get_*_fits` | Post fit plotting and tabulation |

A `do_*_mapping` helper lives in `setup_mapping.R` only if more than one `setup_*` file
calls it. Single caller helpers stay in the file that uses them.

## File map

Every script, what it defines, and what it talks to. This table is generated from the
actual call graph, and shows "what interacts with what". "Calls
into" lists the files whose functions this one calls; "called from" is the reverse.

```{r arch_setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/arch_")
sporc_call_graph <- function(src_dir) {
  files <- sort(list.files(src_dir, pattern = "[.]R$"))
  paths <- file.path(src_dir, files)

  # The trailing .* matters: sub() replaces only what it matches, so without it the
  # rest of the definition line would be left glued onto the function name.
  def_re <- "^([A-Za-z_.][A-Za-z0-9_.]*)[[:space:]]*(<-|=)[[:space:]]*function.*$"
  defs <- character(0)
  defined_in <- setNames(vector("list", length(files)), files)

  for (i in seq_along(files)) {
    lines <- readLines(paths[i], warn = FALSE)
    nm <- unique(sub(def_re, "\\1", grep(def_re, lines, value = TRUE)))
    defined_in[[files[i]]] <- sort(nm)
    defs[setdiff(nm, names(defs))] <- files[i]
  }

  calls <- setNames(vector("list", length(files)), files)
  for (i in seq_along(files)) {
    pd <- utils::getParseData(parse(paths[i], keep.source = TRUE))
    called <- unique(pd$text[pd$token == "SYMBOL_FUNCTION_CALL"])
    called <- called[called %in% names(defs)]
    calls[[files[i]]] <- sort(setdiff(unique(defs[called]), files[i]))
  }

  edges <- do.call(rbind, lapply(files, function(f) {
    if (!length(calls[[f]])) return(NULL)
    data.frame(from = f, to = calls[[f]], stringsAsFactors = FALSE)
  }))
  called_by <- setNames(lapply(files, function(f) sort(edges$from[edges$to == f])), files)

  list(files = files, defined_in = defined_in, calls = calls,
       called_by = called_by, edges = edges)
}

sporc_stage <- function(f) {
  p <- sub("[.]R$", "", sub("_.*$", "", f))
  switch(p, setup = "Setup", model = "Objective",
         refpts = , projection = "Ref points", sim = "Operating model",
         diag = "Diagnostics", plot = "Plotting", utils = "Helpers",
         data = "Data", "Other")
}

STAGES <- c("Setup", "Objective", "Ref points", "Operating model",
            "Diagnostics", "Plotting", "Helpers", "Data")
STAGE_HEADING <- c(
  "Setup"           = "Setup: build data, parameters and mapping",
  "Objective"       = "Objective function and its modules",
  "Ref points"      = "Reference points and projection",
  "Operating model" = "Operating model",
  "Diagnostics"     = "Diagnostics",
  "Plotting"        = "Plotting",
  "Helpers"         = "Shared helpers",
  "Data"            = "Package data")
PAL <- c(Setup = "#7BA7D7", Objective = "#E8A33D", "Ref points" = "#8FBF7F",
         "Operating model" = "#C58FC5", Diagnostics = "#E0796F",
         Plotting = "#6FBFB5", Helpers = "#9E9E9E", Data = "#D0D0D0")

src_dir <- if (dir.exists("../R")) "../R" else if (dir.exists("R")) "R" else NA_character_
have_src <- !is.na(src_dir)
if (have_src) {
  g <- sporc_call_graph(src_dir)
  stage_of <- vapply(g$files, sporc_stage, "")
}
have_igraph <- have_src && requireNamespace("igraph", quietly = TRUE)
```

```{r stage_diagram, fig.width = 9, fig.height = 6.3, dpi = 150, out.width = "100%", fig.alt = "Directed graph of SPoRC pipeline stages, with arrows pointing from each stage to the stages it calls into."}
if (have_igraph) {
  se <- data.frame(from = stage_of[g$edges$from], to = stage_of[g$edges$to])
  se <- se[se$from != se$to, ]
  sw <- aggregate(list(n = rep(1, nrow(se))), by = list(from = se$from, to = se$to), sum)
  sg <- igraph::graph_from_data_frame(sw, directed = TRUE)

  coords <- rbind(
    Plotting          = c( 0.0,  3.0),
    Diagnostics       = c(-2.0,  1.9),
    "Ref points"      = c( 0.7,  1.9),
    "Operating model" = c( 2.9,  1.9),
    Objective         = c(-0.6,  0.8),
    Setup             = c( 2.1,  0.8),
    Helpers           = c( 0.7, -0.4))
  lab <- setNames(rownames(coords), rownames(coords))
  lab["Operating model"] <- "Operating\nmodel"
  keep <- intersect(igraph::V(sg)$name, rownames(coords))
  sg <- igraph::induced_subgraph(sg, keep)

  op <- par(mar = c(0, 0, 0, 0))
  plot(sg, layout = coords[igraph::V(sg)$name, , drop = FALSE],
       vertex.color = PAL[igraph::V(sg)$name], vertex.frame.color = "white",
       vertex.size = 34, vertex.size2 = 20, vertex.shape = "rectangle",
       vertex.label = lab[igraph::V(sg)$name], vertex.label.color = "black",
       vertex.label.cex = 0.8, vertex.label.family = "sans",
       edge.width = 0.8 + 1.6 * sqrt(igraph::E(sg)$n), edge.color = "grey55",
       edge.arrow.size = 0.45, edge.curved = 0.15,
       edge.label = igraph::E(sg)$n, edge.label.cex = 0.78,
       edge.label.color = "grey20", edge.label.family = "sans")
  par(op)
} else {
  cat("igraph is not installed, so the stage diagram was not drawn.\n")
}
```

Edge labels count how many file to file dependencies sit behind each arrow. `Helpers`
is a pure sink and `Plotting` a pure source, which is the shape you want: the objective
function never reaches up into the post fit machinery.

Below the same information at file resolution. A filled cell means the file on that row
calls a function defined in the file in that column, and cells are grouped by stage. Empty
rows call nothing else in the package, empty columns are called by nothing.

```{r dependency_matrix, fig.width = 11.5, fig.height = 10.2, dpi = 150, out.width = "100%", fig.alt = "Dependency matrix of the 52 files in R/, grouped by pipeline stage, where a filled cell means the row file calls a function defined in the column file."}
if (have_src) {
  ord <- order(match(stage_of, STAGES), g$files)
  fl <- g$files[ord]; st <- stage_of[ord]; n <- length(fl)
  A <- matrix(0L, n, n)
  A[cbind(match(g$edges$from, fl), match(g$edges$to, fl))] <- 1L
  lab <- sub("[.]R$", "", fl)

  op <- par(mar = c(0.4, 12.0, 12.0, 8.0))
  plot(NA, xlim = c(0.5, n + 0.5), ylim = c(n + 0.5, 0.5), axes = FALSE,
       xlab = "", ylab = "", xaxs = "i", yaxs = "i")

  blocks <- split(seq_len(n), factor(st, levels = STAGES))
  blocks <- blocks[lengths(blocks) > 0]

  # Alternating stage bands, so the eye can track a row across the empty space.
  for (k in seq_along(blocks)) {
    if (k %% 2 == 1) {
      idx <- blocks[[k]]
      rect(0.5, min(idx) - 0.5, n + 0.5, max(idx) + 0.5, col = "#F4F6F8", border = NA)
      rect(min(idx) - 0.5, 0.5, max(idx) + 0.5, n + 0.5,
           col = adjustcolor("#F4F6F8", alpha.f = 0.6), border = NA)
    }
  }

  # A faint rule along each row that has any dependency, to lead the eye out to
  # the marks in what is otherwise a very sparse grid.
  for (i in seq_len(n)) {
    hits <- which(A[i, ] == 1L)
    if (length(hits)) segments(min(hits) - 0.4, i, max(hits) + 0.4, i,
                               col = adjustcolor(PAL[[st[i]]], alpha.f = 0.28), lwd = 1.1)
  }

  brk <- which(st[-1] != st[-n]) + 0.5
  abline(h = brk, col = "grey62", lwd = 1); abline(v = brk, col = "grey62", lwd = 1)
  rect(0.5, 0.5, n + 0.5, n + 0.5, border = "grey45")

  hit <- which(A == 1L, arr.ind = TRUE)
  rect(hit[, 2] - 0.40, hit[, 1] - 0.40, hit[, 2] + 0.40, hit[, 1] + 0.40,
       col = PAL[st[hit[, 1]]], border = "white", lwd = 0.5)

  # Stage colour strips just inside each margin, and stage names down the side.
  for (k in seq_along(blocks)) {
    idx <- blocks[[k]]; s <- names(blocks)[k]
    rect(-1.9, min(idx) - 0.5, -0.7, max(idx) + 0.5, col = PAL[[s]],
         border = "white", xpd = NA, lwd = 0.6)
    rect(min(idx) - 0.5, -1.9, max(idx) + 0.5, -0.7, col = PAL[[s]],
         border = "white", xpd = NA, lwd = 0.6)
    text(n + 2.6, mean(idx), s, xpd = NA, cex = 0.66, col = PAL[[s]], font = 2, adj = 0)
  }

  axis(2, at = 1:n, labels = lab, las = 2, cex.axis = 0.68, tick = FALSE, line = 0.55)
  axis(3, at = 1:n, labels = lab, las = 2, cex.axis = 0.68, tick = FALSE, line = 0.55)
  mtext("callee  (the file whose function is being called)", side = 3, line = 10.4,
        cex = 0.92, font = 2)
  mtext("caller", side = 2, line = 10.6, cex = 0.92, font = 2)
  par(op)
}
```

The `model_transition` column is the densest in the matrix, which is the point made
above: everything that moves fish through a season goes through that one file.

### Per stage detail

```{r file_map, results = "asis"}
if (!have_src) {
  cat("_The package sources were not available at build time, so the file map",
      "could not be generated._\n")
} else {
  fmt <- function(x) if (!length(x)) "" else paste(sprintf("`%s`", x), collapse = ", ")
  for (s in STAGES) {
    fs <- g$files[stage_of == s]
    if (!length(fs)) next
    cat("\n#### ", STAGE_HEADING[[s]], "\n\n", sep = "")
    cat("| Script | Defines | Calls into | Called from |\n|---|---|---|---|\n")
    for (f in fs) {
      d <- if (length(g$defined_in[[f]])) fmt(g$defined_in[[f]]) else "(documentation only)"
      cat(sprintf("| `%s` | %s | %s | %s |\n", f, d,
                  if (length(g$calls[[f]])) fmt(g$calls[[f]]) else "nothing",
                  fmt(g$called_by[[f]])))
    }
  }
}
```

## How the code is layered

The code is structured as follows and interacts in a hopefully intuitive fashion. Reading it as layers, lowest first:

1. `utils_math.R`, `utils_setup.R`, `utils_postfit.R`, `setup_checks.R`,
   `model_transition.R`, `model_distributions.R`, `model_osa.R`, `model_precision.R`.
   These call nothing else in the package.
2. The model modules (`model_recruitment.R`, `model_biomass.R`, `model_init_naa.R`,
   `model_movement.R`, `model_selectivity.R`, `model_obs_*.R`, `model_lik_*.R`,
   `model_priors_penalties.R`, `model_population_dynamics.R`).
3. `model_objective.R`, which calls only downward into layer 2.
4. The post fit and simulation module: reference points, projection, diagnostics, plotting,
   operating model.

## Tests

Test files in `tests/testthat/` carry the same prefixes as `R/`, so the tests for a given
source file sort next to each other and the name says what is under test rather than which
stock the fixture happens to use.

```{r test-map, results = "asis"}
tdir <- if (dir.exists("../tests/testthat")) "../tests/testthat" else NA_character_
if (is.na(tdir)) {
  cat("_Test directory not available at build time._\n")
} else {
  tf <- sort(list.files(tdir, pattern = "^test-.*[.]R$"))
  grp <- sub("^test-([a-z]+)_.*$", "\\1", tf)
  desc <- c(
    setup       = "Input building: the `map` factor builders and the `Setup_Mod_*` validation",
    model       = "One objective function module each: selectivity, movement, transition, observation models, likelihoods, distributions",
    utils       = "Shared numerical helpers",
    sim         = "Operating model, including simulate then refit self tests",
    refpts      = "SPR and MSY solvers, one file per spatial structure",
    projection  = "Forward projection off a fitted model",
    diag        = "Post fit diagnostics: retrospectives and OSA residuals",
    integration = "Cross cutting agreement between the objective, the reference points and the operating model",
    regression  = "End to end fits pinning `obj$rep` and `nll` for known configurations")
  cat("| Prefix | Files | What it covers |\n|---|---|---|\n")
  for (g in names(desc)) {
    n <- sum(grp == g)
    if (n) cat(sprintf("| `test-%s_*` | %d | %s |\n", g, n, desc[[g]]))
  }
  cat(sprintf("\nThat is %d test files in total.\n", length(tf)))
}
```

Two groups are worth calling out. The `test-regression_*` files pin `obj$rep` and `nll`
values for known configurations against bundled example fits, so they are the ones most
likely to catch an accidental change in the objective function's numerics. The
`test-integration_move_timing_*` files are the tightest constraint on `model_transition.R`
and everything built on it: they check that the reference points, the forward projection
and the operating model all agree under each movement timing, to a tolerance that leaves no
room for a sequencing mistake.

## Extending the model

Adding a new process, fleet type or data source interacts with a predictable set of places.

1. **Setup.** Add or extend a `Setup_Mod_*` function to accept the new spec, validate it,
   and append the results to `input_list$data`, `$par` and `$map`. If parameters need
   fixing or sharing, add a `do_*_mapping` helper in the same file, or reach for the shared
   builders in `setup_mapping.R` if the pattern already exists.
2. **Objective.** Add a `## Section` in `model_objective.R` that consumes the new entries
   and contributes to `jnll` or `nll`. Put it in the section it belongs to (transform,
   observation model, or likelihood and prior) rather than appending at the end. If the
   logic runs to more than a few lines, extract it into a `model_*.R` module instead of
   growing the objective inline.
3. **Simulation.** If the process should be simulateable, add the matching `Setup_Sim_*`
   inputs and the generation logic in `sim_population.R` or `sim_observations.R`.
4. **Post fit.** Expose derived output through a `Get_*` function if other code needs to
   reuse it, and through `plot_figures_tables.R` or `diag_fits.R` if it needs to be seen.
5. **Tests.** Add a targeted test near the feature, and check whether the regression tests
   need updated reference values.
6. **Vignette.** If the feature is user facing, document it in the relevant lettered
   vignette rather than only in code comments.

For the mechanics of getting a change in (branch naming, running tests locally, opening
a PR) see the [Contributing](contributing.html) guide.
