---
title: "Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document is for people editing the `SPoRC` codebase: contributors, successors, future
you. The goal is to keep the package navigable as it grows, especially as spatial model
development adds more moving parts. Update it whenever the pipeline below changes shape (a
new `Setup_Mod_*` stage, a new section of the objective function, a new post fit
diagnostic). A stale architecture document should be treated as a bug.

## Mental model

`SPoRC` runs in three phases, in this order.

1. **Setup.** A chain of R functions builds up three plain lists: `data`, `parameters` and
   `mapping`. No likelihood is evaluated yet. This phase is data wrangling and input
   validation only.
2. **Fit.** `data` and `parameters` go to `RTMB::MakeADFun` wrapped around a single
   objective function, `SPoRC_rtmb()`, which is optimized with `nlminb` plus Newton
   refinement.
3. **Post fit.** The fitted object (`obj$rep`, `obj$sdrep`, `obj$optim`) is passed to
   diagnostics, reference points, plotting and resampling routines.

```
Setup_Mod_Dim()
   -> Setup_Mod_Rec()
   -> Setup_Mod_Biologicals()
   -> Setup_Mod_Movement()
   -> Setup_Mod_Tagging()
   -> Setup_Mod_Catch_and_F()
   -> Setup_Mod_FishIdx_and_Comps()   [+ Setup_Mod_Discard_Comps()]
   -> Setup_Mod_Fishsel_and_Q()       [+ Setup_Mod_Retsel()]
   -> Setup_Mod_SrvIdx_and_Comps()
   -> Setup_Mod_Srvsel_and_Q()
   -> Setup_Mod_Weighting()
        |
        v  input_list$data, input_list$par, input_list$map
   fit_model(data, parameters, mapping)
        |  RTMB::MakeADFun(cmb(SPoRC_rtmb, data), parameters, map = mapping)
        |  nlminb() + Newton steps
        v
   obj  (obj$rep, obj$sdrep, obj$optim)
        |
        +--> Get_Reference_Points()          SPR and MSY
        +--> get_idx_fits() / get_osa()      fit extraction and OSA residuals
        +--> plot_* / get_*_plot             figures and tables
        +--> do_retrospective(), do_jitter(), do_likelihood_profile(), run_francis()
        +--> Do_Population_Projection(), condition_closed_loop_simulations()
```

### The `input_list` accumulator

Every `Setup_Mod_*` function takes an `input_list` (initialized by `Setup_Mod_Dim()` as
`list(data = list(), par = list(), map = list())`) and returns it with more keys filled in.
This is the central structure of the package: setup functions are pipeline stages that
thread one growing list through each other, not independent builders.

Order matters. Later stages read dimensions (`input_list$data$n_regions` and friends) set
by earlier ones, and default arguments frequently reference `input_list$data$...` in their
own definitions. See `Setup_Mod_Rec()`'s defaults for an example.

At the end of the chain `input_list$data`, `input_list$par` and `input_list$map` are
unpacked and passed to `fit_model()`. The `Setup_Sim_*` functions live in the same files as
their `Setup_Mod_*` counterparts and follow the identical accumulator pattern, but build
inputs for the operating model instead of for fitting.

### The objective function

`R/model_objective.R` holds the one RTMB objective function for every model configuration
the package supports. There is no per model branching at the file level. The objective
always runs end to end and is regulated by the switches set during setup, such as
`rec_model`, `move_type`, and the likelihood and OSA flags.

It moves through, in order: parameter transforms (movement, natural mortality,
selectivity), then mortality, recruitment, initial age structure, population projection,
the observation models (fishery, survey, tagging), the likelihood components (retained and
discarded catch, indices, compositions, tags), and finally priors and penalties.

Each section hands off to a module in one of the other `model_*.R` files, so the objective reads as an order of operations rather than as the arithmetic itself. If you are adding a new data source or process, this is where its likelihood contribution gets wired in, and the matching `Setup_Mod_*` function is
where its data, parameters and mapping get prepared beforehand.

Within `model_objective.R` the `## Section` and `### Subsection` comment banner style is
used consistently. Keep using it if you extend the file, since it is the only navigation
aid inside a function of this size.

Note that `R/model_transition.R` is worth knowing about before anything else. It computes what happens to a vector of abundance over one season, given movement, mortality,
season duration and a movement timing. Eleven other files go through it, including the
objective function's dynamics, the reference point solvers and the operating model.

## Naming conventions

### Files

The prefix on a file name says which phase it belongs to.

| Prefix | Phase |
|---|---|
| `setup_` | Builds `data`, `parameters` and `mapping`, one process or data source per file. A source large enough to warrant it is split by topic, as `setup_fishery_*.R` and `setup_survey_*.R` are |
| `model_` | The objective function and the modules it delegates to |
| `refpts_`, `projection` | Reference points and forward projection off a fitted model |
| `sim_` | The operating model: self testing and closed loop simulation |
| `diag_` | Post fit diagnostics and residuals |
| `plot_` | Figures and summary tables |
| `utils_` | Helpers shared across phases |
| `data_` | Roxygen documentation for the bundled datasets |

### Functions

| Prefix | Role |
|---|---|
| `Setup_Mod_*` | Builds inputs for a fitted object |
| `Setup_Sim_*` | Same, for the operating model. Lives beside its `Setup_Mod_*` counterpart, except for the fishery and survey, whose operating model inputs span several topic files and so share `setup_sim_fleets.R` |
| `do_*_mapping` | Internal helper that builds the RTMB `map` for one parameter block. This is how parameters get fixed, shared across blocks, or turned into random effect deviations |
| `Get_*` | Pulls a derived quantity out of set up data, parameters or a fitted object |
| `do_*` (top level) | A post fit procedure that refits or perturbs a fitted model |
| `get_*_plot`, `get_*_fits` | Post fit plotting and tabulation |

A `do_*_mapping` helper lives in `setup_mapping.R` only if more than one `setup_*` file
calls it. Single caller helpers stay in the file that uses them.

## File map

Every script, what it defines, and what it talks to. This table is generated from the
actual call graph, and shows "what interacts with what". "Calls
into" lists the files whose functions this one calls; "called from" is the reverse.

```{r arch_setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/arch_")
sporc_call_graph <- function(src_dir) {
  files <- sort(list.files(src_dir, pattern = "[.]R$"))
  paths <- file.path(src_dir, files)

  # The trailing .* matters: sub() replaces only what it matches, so without it the
  # rest of the definition line would be left glued onto the function name.
  def_re <- "^([A-Za-z_.][A-Za-z0-9_.]*)[[:space:]]*(<-|=)[[:space:]]*function.*$"
  defs <- character(0)
  defined_in <- setNames(vector("list", length(files)), files)

  for (i in seq_along(files)) {
    lines <- readLines(paths[i], warn = FALSE)
    nm <- unique(sub(def_re, "\\1", grep(def_re, lines, value = TRUE)))
    defined_in[[files[i]]] <- sort(nm)
    defs[setdiff(nm, names(defs))] <- files[i]
  }

  calls <- setNames(vector("list", length(files)), files)
  for (i in seq_along(files)) {
    pd <- utils::getParseData(parse(paths[i], keep.source = TRUE))
    called <- unique(pd$text[pd$token == "SYMBOL_FUNCTION_CALL"])
    called <- called[called %in% names(defs)]
    calls[[files[i]]] <- sort(setdiff(unique(defs[called]), files[i]))
  }

  edges <- do.call(rbind, lapply(files, function(f) {
    if (!length(calls[[f]])) return(NULL)
    data.frame(from = f, to = calls[[f]], stringsAsFactors = FALSE)
  }))
  called_by <- setNames(lapply(files, function(f) sort(edges$from[edges$to == f])), files)

  list(files = files, defined_in = defined_in, calls = calls,
       called_by = called_by, edges = edges)
}

sporc_stage <- function(f) {
  p <- sub("[.]R$", "", sub("_.*$", "", f))
  switch(p, setup = "Setup", model = "Objective",
         refpts = , projection = "Ref points", sim = "Operating model",
         diag = "Diagnostics", plot = "Plotting", utils = "Helpers",
         data = "Data", "Other")
}

STAGES <- c("Setup", "Objective", "Ref points", "Operating model",
            "Diagnostics", "Plotting", "Helpers", "Data")
STAGE_HEADING <- c(
  "Setup"           = "Setup: build data, parameters and mapping",
  "Objective"       = "Objective function and its modules",
  "Ref points"      = "Reference points and projection",
  "Operating model" = "Operating model",
  "Diagnostics"     = "Diagnostics",
  "Plotting"        = "Plotting",
  "Helpers"         = "Shared helpers",
  "Data"            = "Package data")
PAL <- c(Setup = "#7BA7D7", Objective = "#E8A33D", "Ref points" = "#8FBF7F",
         "Operating model" = "#C58FC5", Diagnostics = "#E0796F",
         Plotting = "#6FBFB5", Helpers = "#9E9E9E", Data = "#D0D0D0")

src_dir <- if (dir.exists("../R")) "../R" else if (dir.exists("R")) "R" else NA_character_
have_src <- !is.na(src_dir)
if (have_src) {
  g <- sporc_call_graph(src_dir)
  stage_of <- vapply(g$files, sporc_stage, "")
}
have_igraph <- have_src && requireNamespace("igraph", quietly = TRUE)
```

```{r stage_diagram, fig.width = 9, fig.height = 6.3, dpi = 150, out.width = "100%", fig.alt = "Directed graph of SPoRC pipeline stages, with arrows pointing from each stage to the stages it calls into."}
if (have_igraph) {
  se <- data.frame(from = stage_of[g$edges$from], to = stage_of[g$edges$to])
  se <- se[se$from != se$to, ]
  sw <- aggregate(list(n = rep(1, nrow(se))), by = list(from = se$from, to = se$to), sum)
  sg <- igraph::graph_from_data_frame(sw, directed = TRUE)

  coords <- rbind(
    Plotting          = c( 0.0,  3.0),
    Diagnostics       = c(-2.0,  1.9),
    "Ref points"      = c( 0.7,  1.9),
    "Operating model" = c( 2.9,  1.9),
    Objective         = c(-0.6,  0.8),
    Setup             = c( 2.1,  0.8),
    Helpers           = c( 0.7, -0.4))
  lab <- setNames(rownames(coords), rownames(coords))
  lab["Operating model"] <- "Operating\nmodel"
  keep <- intersect(igraph::V(sg)$name, rownames(coords))
  sg <- igraph::induced_subgraph(sg, keep)

  op <- par(mar = c(0, 0, 0, 0))
  plot(sg, layout = coords[igraph::V(sg)$name, , drop = FALSE],
       vertex.color = PAL[igraph::V(sg)$name], vertex.frame.color = "white",
       vertex.size = 34, vertex.size2 = 20, vertex.shape = "rectangle",
       vertex.label = lab[igraph::V(sg)$name], vertex.label.color = "black",
       vertex.label.cex = 0.8, vertex.label.family = "sans",
       edge.width = 0.8 + 1.6 * sqrt(igraph::E(sg)$n), edge.color = "gray55",
       edge.arrow.size = 0.45, edge.curved = 0.15,
       edge.label = igraph::E(sg)$n, edge.label.cex = 0.78,
       edge.label.color = "gray20", edge.label.family = "sans")
  par(op)
} else {
  cat("igraph is not installed, so the stage diagram was not drawn.\n")
}
```

Edge labels count how many file to file dependencies sit behind each arrow. `Helpers`
is a pure sink and `Plotting` a pure source, which is the shape you want: the objective
function never reaches up into the post fit machinery.

Below the same information at file resolution. A filled cell means the file on that row
calls a function defined in the file in that column, and cells are grouped by stage. Empty
rows call nothing else in the package, empty columns are called by nothing.

```{r dependency_matrix, fig.width = 11.5, fig.height = 10.2, dpi = 150, out.width = "100%", fig.alt = "Dependency matrix of the 52 files in R/, grouped by pipeline stage, where a filled cell means the row file calls a function defined in the column file."}
if (have_src) {
  ord <- order(match(stage_of, STAGES), g$files)
  fl <- g$files[ord]; st <- stage_of[ord]; n <- length(fl)
  A <- matrix(0L, n, n)
  A[cbind(match(g$edges$from, fl), match(g$edges$to, fl))] <- 1L
  lab <- sub("[.]R$", "", fl)

  op <- par(mar = c(0.4, 12.0, 12.0, 8.0))
  plot(NA, xlim = c(0.5, n + 0.5), ylim = c(n + 0.5, 0.5), axes = FALSE,
       xlab = "", ylab = "", xaxs = "i", yaxs = "i")

  blocks <- split(seq_len(n), factor(st, levels = STAGES))
  blocks <- blocks[lengths(blocks) > 0]

  # Alternating stage bands, so the eye can track a row across the empty space.
  for (k in seq_along(blocks)) {
    if (k %% 2 == 1) {
      idx <- blocks[[k]]
      rect(0.5, min(idx) - 0.5, n + 0.5, max(idx) + 0.5, col = "#F4F6F8", border = NA)
      rect(min(idx) - 0.5, 0.5, max(idx) + 0.5, n + 0.5,
           col = adjustcolor("#F4F6F8", alpha.f = 0.6), border = NA)
    }
  }

  # A faint rule along each row that has any dependency, to lead the eye out to
  # the marks in what is otherwise a very sparse grid.
  for (i in seq_len(n)) {
    hits <- which(A[i, ] == 1L)
    if (length(hits)) segments(min(hits) - 0.4, i, max(hits) + 0.4, i,
                               col = adjustcolor(PAL[[st[i]]], alpha.f = 0.28), lwd = 1.1)
  }

  brk <- which(st[-1] != st[-n]) + 0.5
  abline(h = brk, col = "gray62", lwd = 1); abline(v = brk, col = "gray62", lwd = 1)
  rect(0.5, 0.5, n + 0.5, n + 0.5, border = "gray45")

  hit <- which(A == 1L, arr.ind = TRUE)
  rect(hit[, 2] - 0.40, hit[, 1] - 0.40, hit[, 2] + 0.40, hit[, 1] + 0.40,
       col = PAL[st[hit[, 1]]], border = "white", lwd = 0.5)

  # Stage color strips just inside each margin, and stage names down the side.
  for (k in seq_along(blocks)) {
    idx <- blocks[[k]]; s <- names(blocks)[k]
    rect(-1.9, min(idx) - 0.5, -0.7, max(idx) + 0.5, col = PAL[[s]],
         border = "white", xpd = NA, lwd = 0.6)
    rect(min(idx) - 0.5, -1.9, max(idx) + 0.5, -0.7, col = PAL[[s]],
         border = "white", xpd = NA, lwd = 0.6)
    text(n + 2.6, mean(idx), s, xpd = NA, cex = 0.66, col = PAL[[s]], font = 2, adj = 0)
  }

  axis(2, at = 1:n, labels = lab, las = 2, cex.axis = 0.68, tick = FALSE, line = 0.55)
  axis(3, at = 1:n, labels = lab, las = 2, cex.axis = 0.68, tick = FALSE, line = 0.55)
  mtext("callee  (the file whose function is being called)", side = 3, line = 10.4,
        cex = 0.92, font = 2)
  mtext("caller", side = 2, line = 10.6, cex = 0.92, font = 2)
  par(op)
}
```

The `model_transition` column is the densest in the matrix, which is the point made
above: everything that moves fish through a season goes through that one file.

### Per stage detail

```{r file_map, results = "asis"}
if (!have_src) {
  cat("_The package sources were not available at build time, so the file map",
      "could not be generated._\n")
} else {
  fmt <- function(x) if (!length(x)) "" else paste(sprintf("`%s`", x), collapse = ", ")
  for (s in STAGES) {
    fs <- g$files[stage_of == s]
    if (!length(fs)) next
    cat("\n#### ", STAGE_HEADING[[s]], "\n\n", sep = "")
    cat("| Script | Defines | Calls into | Called from |\n|---|---|---|---|\n")
    for (f in fs) {
      d <- if (length(g$defined_in[[f]])) fmt(g$defined_in[[f]]) else "(documentation only)"
      cat(sprintf("| `%s` | %s | %s | %s |\n", f, d,
                  if (length(g$calls[[f]])) fmt(g$calls[[f]]) else "nothing",
                  fmt(g$called_by[[f]])))
    }
  }
}
```

## Module walkthroughs

The file map above shows what calls what. It does not show *why* a module is split into
the particular functions it has, which matters more once you are adding to one. Two
modules below are worked through in full as examples of the pattern; most others in
`model_*.R`/`setup_*.R` follow the same shape.

### Growth

Growth is the mean length (and its spread) at age, built from up to six parameters
(`L1`, `L2`, `K`, `CV1`, `CV2`, and `rho` under the Richards form) and turned into the
size-age transition matrix every length composition, conditional age-at-length
observation, and length-based selectivity reads.

```
Setup_Mod_Biologicals()                                    [setup_biologicals.R]
  validates growth_model/growth_tv_model/growth_semipar, populates $data and $par
  -> do_growth_mapping()                                    [setup_biologicals.R]
       builds $map for ln_growth_pars, ln_growth_devs, ln_growth_semipar_devs
        |
        v
SPoRC_rtmb()                                                [model_objective.R]
  -> growth_args_from_model()                                [model_growth.R]
       pulls growth's ~20 arguments out of the RTMB frame by name
  -> growth_tv_type == "curve": Get_Growth() builds every year up front
                                                                [model_growth.R]
     growth_tv_type == "cohort": one year at a time, from inside the
                                  population's year-by-year loop, because the
                                  plus group's size depends on how many old vs.
                                  really-old fish are alive THAT year, which
                                  growth cannot know in advance
       -> Get_Growth_Year()                                    [model_growth.R]
            builds this year's growth, given the population's numbers at age
       -> growth_take_year()                                   [model_growth.R]
            copies that year's growth into the objective's own arrays
       -> mortality_year()                                     [model_objective.R]
            rebuilds that year's fishing/total mortality; only actually
            changes anything for a length-based gear, whose age-specific
            selectivity is read off growth's key, but runs every year
            regardless rather than special-casing age-based gears out
  -> (both) growth_selected_waa_year()                        [model_growth.R]
       called from the objective's mortality_year step, once per year, only for
       fleets flagged fish_waa_selected / srv_waa_selected
  -> ### Growth (Process Error): Get_PE_loglik()          [model_priors_penalties.R]
       penalizes ln_growth_devs and ln_growth_semipar_devs for departing from the
       process they are supposed to follow, the same function selectivity uses
```

| Function | Why it is its own function |
|---|---|
| `get_laa_curve()` | The pure closed-form curve: mean length and spread at a vector of real ages, from the six parameters alone. No notion of years, fleets, or cohorts, so it is unit-testable against the textbook von Bertalanffy/Richards formula directly (`test-model_growth.R` does exactly that) without any model scaffolding around it. |
| `growth_start_state()` | The state a year starts from: `get_laa_curve()` at every integer age, with the plus group adjusted. Both `growth_tv_type`s need a "year 1 initial state" and need it built identically, so it is factored out rather than duplicated in `Get_Growth()` and `Get_Growth_Year()`. |
| `growth_laa_at()` | The within-year evaluator at one elapsed time `e`. Under `"curve"` growth every age is just `get_laa_curve()` at its real age; under `"cohort"` growth, ages already propagated carry a length history the closed form does not know about, so they instead grow from `L_beg` via `grow_increment()`. One function, branching on `cohort`, rather than two, because everything else in it (the plus group, the CV, the semi-parametric deviation) is identical either way. |
| `grow_increment()` | The one piece of math with no dependency on ages, `A1`, or `A2` at all: grow a length `e` further given `K`, `Linf`, `rho`. Reused in two genuinely different contexts, evaluating a fleet's within-year timing (`growth_laa_at`) and advancing a cohort a full year (`Get_Growth_Year`), which is why it is not inlined into either. |
| `growth_fill_year()` | "Every fleet, timing, and spawning point in one year," given a start-of-year state: walks each distinct `t_fish`/`t_srv`/`t_spawn` value once (deduplicated), builds the key via `get_alk()`, and writes it into every fleet that reads at that timing. Both `Get_Growth()` (looping over years on the "curve" path) and `Get_Growth_Year()` (one year at a time on the "cohort" path) call it with a different start-of-year state, so the fleet-timing bookkeeping is written once. |
| `plus_group_size()` | The survivorship-weighted mixture for a plus group's mean length. Kept out of `get_laa_curve()` because it needs `Linf` and the curve's own last-age value together, plus the `growth_plus_group` flag; `get_laa_curve()`'s job is only ever the raw curve. |
| `get_growth_pars_year()` | Applies a year's time-varying deviations (log or logit link) to the base parameters once, before anything else runs, so `growth_start_state()`, `growth_laa_at()`, and `growth_fill_year()` never have to know deviations exist; they only ever see "this year's realized parameters." |
| `do_growth_mapping()` | Setup side: builds every growth map (parameters, time-varying deviations, semi-parametric surface) in one place, called from `Setup_Mod_Biologicals()`'s `# Mapping Options` section next to `do_natmort_mapping()`, so that function's own body stays validate-then-hand-off like every other `Setup_Mod_*`. |

Two separate things are going on here, worth telling apart.

Growth has to run *inside* the population loop at all, one year at a time, only because of
the plus group. Under `growth_tv_type = "cohort"`, the plus group's mean size is a mix of
the fish that were already old and the fish that just grew into being old, weighted by how
many of each there are, and that count is only known once the population loop has actually
reached that year. This is true whatever selectivity looks like: even a model with plain
age-based selectivity still needs growth for weight at age, spawning biomass, and length
compositions, and still has this same plus-group problem. `growth_tv_type = "curve"` has no
such problem (nothing carries a length history from year to year) and so builds every
year's growth up front, before the population loop even starts.

Rebuilding that year's mortality right alongside growth, though, *is* specifically about
length-based selectivity. `get_population_projection()` (in `model_population_dynamics.R`)
takes an argument that lets it call back out to the objective once a year; `model_objective.R`
uses it to run growth for that year, copy the results in, and then also rerun that year's
fishing and total mortality before handing control back to the population loop. That last
step only changes anything when a fleet's selectivity is defined on length rather than age,
since that is the one case where a fleet's age-specific selectivity is read off growth's
key and so moves whenever growth does. Under purely age-based selectivity the rerun leaves
mortality exactly as it already was; the code does not bother telling the two cases apart,
since redoing an unchanged calculation is cheap and the alternative is a second, more
complicated code path only for a case that costs nothing to just repeat.

### Selectivity (fishery, retention, survey)

A model has up to three kinds of selectivity: the fishery's, retention's (what fraction of
what the fishery catches is kept vs. thrown back), and the survey's. All three are set up
and evaluated by the exact same code, run three separate times with a short label
(`"fish"`, `"ret"`, `"srv"`) telling it which fleet count and which set of fields to read
and write. This is not three similar-looking implementations; it is one implementation
called three times, which is why retention setup (`Setup_Mod_Retsel()`) simply calls the
same functions fishery setup (`Setup_Mod_Fishsel_and_Q()`) does.

```
Setup_Mod_Fishsel_and_Q()/Setup_Mod_Retsel()/Setup_Mod_Srvsel_and_Q()
                                              [setup_fishery_selectivity.R / setup_survey_selectivity.R]
  validate blocks, functional form per block, time-varying spec
  -> do_fixed_sel_pars_mapping(..., "fish"/"ret"/"srv")            [setup_mapping.R]
       maps the block-level functional-form parameters (b50, slope, double normal's 6, ...)
  -> do_sel_pe_pars_mapping(..., "fish"/"ret"/"srv")                [setup_mapping.R]
       maps the process-error settings for whichever time-varying form is chosen
  -> do_sel_devs_mapping(..., "fish"/"ret"/"srv")                   [setup_mapping.R]
       maps the year-by-year deviations themselves
        |
        v
SPoRC_rtmb()                                                      [model_objective.R]
  ## Selectivity: Get_Selex_Array(), once for each of the three     [model_selectivity.R]
       works out which block and which deviation applies to a given
       region/year/fleet, then hands the actual math off to:
    -> Get_Selex()                                                 [model_selectivity.R]
         given one set of parameters and a vector of bins, the selectivity at
         each bin; knows nothing about years, blocks, or fleets
  ### Selectivity (Penalty): Get_PE_loglik(), once for each of the three
                                                                [model_priors_penalties.R]
       penalizes the year-by-year deviations for departing from whatever
       process they are supposed to follow (iid, random walk, ...); the same
       penalty function growth uses
  ### Selectivity Smoothness (Penalty)                             [model_objective.R]
       a separate penalty just for the bicubic spline form, since a smooth
       surface is not "deviations from a base curve" the way the others are
  ### Selectivity (Prior) / Parameter Centering (Penalty)
    -> get_selex_prior() / get_selex_fixed_penalty()            [model_priors_penalties.R]
```

| Function | Why it is its own function |
|---|---|
| `do_fixed_sel_pars_mapping()`, `do_sel_pe_pars_mapping()`, `do_sel_devs_mapping()` | Three separate functions because a fleet can turn each piece on independently: the curve's own parameters, whether it varies over time and under what process, and the actual deviation values. A fleet might hold the curve fixed with no deviations, vary it under one process with the process's own settings fixed, or estimate both together. Keeping the three apart means each one's sharing options do not have to be reconciled against the other two inside a single function. |
| `Get_Selex_Array()` | The only place that knows about SPoRC's blocks, year-by-year deviations, and fleet counts. Works out which block and which deviation apply to a given region/year/fleet, then hands off to `Get_Selex()`. |
| `Get_Selex()` | Given one functional form and its parameters, the selectivity at a vector of bins, nothing else. Because it does not know about years, blocks, or fleets, `test-model_selectivity.R` can check every functional form directly against its formula without building a whole model around it, and `Get_Selex_Array()`'s only job is the block/year/fleet bookkeeping around it. |
| `Get_PE_loglik()` | Lives in `model_priors_penalties.R`, not `model_selectivity.R` or `model_growth.R`, because it does not need to know which one is calling it: give it a deviation array, its map, and a code for which process (iid, random walk, ...) it is supposed to follow, and it penalizes departures from that process. That is what lets growth and selectivity call the same function with the same five arguments. |

### The same shape shows up elsewhere too

The setup side of this pattern is used everywhere in the package: every `Setup_Mod_*`
function ends by handing its parameter blocks to one or more `do_*_mapping` helpers
(`do_RecDevs_mapping()` for recruitment, `do_conv_tag_fish_reporting_pars_mapping()` for
tagging, and so on). Each one stays in the topic's own `setup_*.R` file unless more than
one file needs it, in which case it moves to the shared `setup_mapping.R`.

The scoring side is not as uniform, which is worth knowing. Growth and selectivity's
year-by-year deviations both go through the same `Get_PE_loglik()`. Movement's own
deviations (`move_devs`) instead have their own penalty function, `Get_move_PE_loglik()`, sitting
right next to `Get_PE_loglik()` in `model_priors_penalties.R` rather than sharing it,
because a movement deviation has to know about which regions are adjacent to which and
what kind of movement model is in use, which `Get_PE_loglik()` has no way to be told about.
Tagging's reporting rate is different again: it is a single estimated number rather than
something that varies by year or bin, so it gets a plain prior (`get_tagrep_prior()`)
instead of a deviation series
at all, since it is a single estimated rate rather than something that varies by year or
bin. When adding a new time-varying process, the question to ask is which of these three
shapes it actually is, and whether it truly fits `Get_PE_loglik()`'s five arguments or
needs a sibling function the way movement's does, rather than assuming growth and
selectivity's choice generalizes.

## How the code is layered

The code is structured as follows and interacts in a hopefully intuitive fashion. Reading it as layers, lowest first:

1. `utils_math.R`, `utils_setup.R`, `utils_postfit.R`, `setup_checks.R`,
   `model_transition.R`, `model_distributions.R`, `model_osa.R`, `model_precision.R`.
   These call nothing else in the package.
2. The model modules (`model_recruitment.R`, `model_biomass.R`, `model_init_naa.R`,
   `model_movement.R`, `model_selectivity.R`, `model_obs_*.R`, `model_lik_*.R`,
   `model_priors_penalties.R`, `model_population_dynamics.R`).
3. `model_objective.R`, which calls only downward into layer 2.
4. The post fit and simulation module: reference points, projection, diagnostics, plotting,
   operating model.

## Tests

Test files in `tests/testthat/` carry the same prefixes as `R/`, so the tests for a given
source file sort next to each other and the name says what is under test rather than which
stock the fixture happens to use.

```{r test-map, results = "asis"}
tdir <- if (dir.exists("../tests/testthat")) "../tests/testthat" else NA_character_
if (is.na(tdir)) {
  cat("_Test directory not available at build time._\n")
} else {
  tf <- sort(list.files(tdir, pattern = "^test-.*[.]R$"))
  grp <- sub("^test-([a-z]+)_.*$", "\\1", tf)
  desc <- c(
    setup       = "Input building: the `map` factor builders and the `Setup_Mod_*` validation",
    model       = "One objective function module each: selectivity, movement, transition, observation models, likelihoods, distributions",
    utils       = "Shared numerical helpers",
    sim         = "Operating model, including simulate then refit self tests",
    refpts      = "SPR and MSY solvers, one file per spatial structure",
    projection  = "Forward projection off a fitted model",
    diag        = "Post fit diagnostics: retrospectives and OSA residuals",
    integration = "Cross cutting agreement between the objective, the reference points and the operating model",
    regression  = "End to end fits pinning `obj$rep` and `nll` for known configurations")
  cat("| Prefix | Files | What it covers |\n|---|---|---|\n")
  for (g in names(desc)) {
    n <- sum(grp == g)
    if (n) cat(sprintf("| `test-%s_*` | %d | %s |\n", g, n, desc[[g]]))
  }
  cat(sprintf("\nThat is %d test files in total.\n", length(tf)))
}
```

Two groups are worth calling out. The `test-regression_*` files pin `obj$rep` and `nll`
values for known configurations against bundled example fits, so they are the ones most
likely to catch an accidental change in the objective function's numerics. The
`test-integration_move_timing_*` files are the tightest constraint on `model_transition.R`
and everything built on it: they check that the reference points, the forward projection
and the operating model all agree under each movement timing, to a tolerance that leaves no
room for a sequencing mistake.

## Extending the model

Adding a new process, fleet type or data source interacts with a predictable set of places.

1. **Setup.** Add or extend a `Setup_Mod_*` function to accept the new spec, validate it,
   and append the results to `input_list$data`, `$par` and `$map`. If parameters need
   fixing or sharing, add a `do_*_mapping` helper in the same file, or use the shared
   builders in `setup_mapping.R` if the pattern already exists.
2. **Objective.** Add a `## Section` in `model_objective.R` that consumes the new entries
   and contributes to `jnll` or `nll`. Put it in the section it belongs to (transform,
   observation model, or likelihood and prior) rather than appending at the end. If the
   logic runs to more than a few lines, extract it into a `model_*.R` module instead of
   growing the objective inline.
3. **Simulation.** If the process should be simulateable, add the matching `Setup_Sim_*`
   inputs and the generation logic in `sim_population.R` or `sim_observations.R`.
4. **Post fit.** Expose derived output through a `Get_*` function if other code needs to
   reuse it, and through `plot_figures_tables.R` or `diag_fits.R` if it needs to be seen.
5. **Tests.** Add a targeted test near the feature, and check whether the regression tests
   need updated reference values.
6. **Vignette.** If the feature is user facing, document it in the relevant lettered
   vignette rather than only in code comments.

For the mechanics of getting a change in (branch naming, running tests locally, opening
a PR) see the [Contributing](contributing.html) guide.
