---
title: "Starting Values and Fixing (and Sharing) Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{j_starting_mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In addition to the `Setup_xxx` functions, users can access several advanced model features. These include:

- Specifying starting values (closely tied to parameter fixing)  
- Fixing parameters  
- Sharing parameters across model partitions  

## Starting Values

Starting values can be specified in two ways:

1. Directly within the `Setup_xxx` functions. Each setup function accepts starting values through the `...` argument. The inputs must match the model's parameter names and dimensions exactly. For more information on parameter dimensions, see the *Description of Model Parameters* vignette.

- `Setup_Mod_Rec`  
  *ln_global_R0*, *rec_region_prop_pars*, *rec_seas_prop_pars*, *steepness_h*, *ln_InitDevs*, *ln_RecDevs*, *ln_sigmaR*

- `Setup_Mod_Biologicals`  
  *ln_M*, *M_offset*

- `Setup_Mod_Movement`  
  *move_pars*, *move_devs*, *move_pe_pars*, *log_move_diffusion_pars*, *move_preference_pars*

- `Setup_Mod_Tagging`  
  *ln_init_conv_tag_mort*, *ln_conv_tag_shed*, *ln_conv_fish_tag_theta*, *conv_tag_fish_reporting_pars*

- `Setup_Mod_Catch_and_F`  
  *ln_sigmaC*, *ln_sigmaF*, *ln_F_mean*, *ln_F_devs*

- `Setup_Mod_FishIdx_and_Comps`  
  *ln_FishAge_theta*, *FishAge_corr_pars*, *ln_FishAge_theta_agg*, *FishAge_corr_pars_agg*,  
  *ln_FishLen_theta*, *FishLen_corr_pars*, *ln_FishLen_theta_agg*, *FishLen_corr_pars_agg*

- `Setup_Mod_SrvIdx_and_Comps`  
  *ln_SrvAge_theta*, *SrvAge_corr_pars*, *ln_SrvAge_theta_agg*, *SrvAge_corr_pars_agg*,  
  *ln_SrvLen_theta*, *SrvLen_corr_pars*, *ln_SrvLen_theta_agg*, *SrvLen_corr_pars_agg*

- `Setup_Mod_Fishsel_and_Q`  
  *fish_fixed_sel_pars*, *ln_fish_q*, *fishsel_pe_pars*, *ln_fishsel_devs*

- `Setup_Mod_Srvsel_and_Q`  
  *ln_srv_fixed_sel_pars*, *ln_srv_q*, *srvsel_pe_pars*, *ln_srvsel_devs*

2. Post-hoc modification of starting values. Alternatively, you can first call the setup functions without specifying starting values, then access and modify the internally created parameter list (`input_list$par`) before running the model.

In the following, we will illustrate both methods using the recruitment module (`Setup_Mod_Rec`) and specify starting values for `ln_global_R0` and `ln_sigmaR`. First let us load in the package and define the model dimensions.

```{r Setup, message = FALSE, eval = TRUE}
# Load in packages
library(SPoRC) 
data("sgl_rg_sable_data") # load in data

input_list <- Setup_Mod_Dim(years = 1:length(sgl_rg_sable_data$years), # vector of years 
                            # (corresponds to year 1960 - 2024)
                            ages = 1:length(sgl_rg_sable_data$ages), # vector of ages
                            lens = seq(41,99,2), # number of lengths
                            n_regions = 1, # number of regions
                            n_sexes = sgl_rg_sable_data$n_sexes, # number of sexes == 1,
                            # female, == 2 male
                            n_fish_fleets = sgl_rg_sable_data$n_fish_fleets, # number of fishery
                            # fleet == 1, fixed gear, == 2 trawl gear
                            n_srv_fleets = sgl_rg_sable_data$n_srv_fleets, # number of survey fleets
                            n_pop = sgl_rg_sable_data$n_pop, # number of populations
                            verbose = FALSE
                            )
```

We can specify starting values directly using the `Setup_Mod_Rec` function. Note that all inputs passed via the `...` argument must exactly match the parameter names and their expected dimensions in the model (see the *Description of Model Parameters* vignette for details).

```{r}
input_list <- Setup_Mod_Rec(
  input_list = input_list,        # input data list from above
  # Model options
  do_rec_bias_ramp = 0,       # disable bias ramp
  sigmaR_switch = as.integer(length(1960:1975)),  # switch from early to late sigmaR
  dont_est_recdev_last = 1,       # do not estimate last recruitment deviate
  rec_model = "mean_rec",         # recruitment model type
  init_age_strc = 1,              # geometric series for initial age structure
  
  # Specify starting values
  ln_global_R0 = log(30),         # starting value for global R0
  ln_sigmaR = array(log(1.5), dim = c(2, input_list$data$n_pop, input_list$data$n_regions))  # starting values for early and late sigmaR
)
```

In this example, the starting value for *ln_global_R0* is set to `log(30)`, while *ln_sigmaR* is set to `log(1.5)` for both the early (index 1 along the first dimension) and late (index 2) periods across all populations and regions.

Alternatively, starting values can be assigned after running the setup functions. Users can extract the internal parameter list and modify starting values as needed:

```{r}
input_list <- Setup_Mod_Rec(input_list = input_list, # input data list from above
                            # Model options
                            do_rec_bias_ramp = 0, # don't do bias ramp 
                            sigmaR_switch = as.integer(length(1960:1975)), # when to switch from early to late sigmaR
                            dont_est_recdev_last = 1, # don't estimate last recruitment deviate
                            rec_model = "mean_rec", # recruitment model
                            init_age_strc = 1 # geometric series to derive age structure
                            )

# Specify starting values post-hoc
# R0
input_list$par$ln_global_R0 # default starting value
input_list$par$ln_global_R0 <- log(30) # user specified starting value

# sigmaR
input_list$par$ln_sigmaR # default starting value
input_list$par$ln_sigmaR[,,] <- log(1.5) # user specified starting value
```

## Mapping
Mapping is a core feature of TMB and RTMB models that controls which parameters are estimated and how they relate to each other. It works through a named list (`input_list$map`) where each entry corresponds to a parameter array flattened into a vector of factors. The rules are simple:

- **`factor(NA)`** at a position means that parameter is **fixed**, it is kept at its value in `input_list$par` and not estimated.
- **Shared integers** at two or more positions means those parameters are **constrained to be equal** during estimation, only one value is estimated for the group.
- **Unique integers** at each position means each parameter is **independently estimated**.

Because fixing a parameter holds it at its current value in `input_list$par`, starting values and fixing must always be set together. If you fix a parameter without also setting a starting value, it will be kept at the default value, which may not be what you intend.

### Fixing Parameters
The `Setup_xxx` functions include arguments that allow certain parameters to be fixed. As an example, we use `Setup_Mod_Rec` to show how *ln_sigmaR* can be fixed using the convenience argument `sigmaR_spec = "fix"`.

```{r}
input_list <- Setup_Mod_Rec(input_list = input_list, # input data list from above
                            # Model options
                            do_rec_bias_ramp = 0, # don't do bias ramp 
                            sigmaR_switch = as.integer(length(1960:1975)), # when to switch from early to late sigmaR
                            dont_est_recdev_last = 1, # don't estimate last recruitment deviate
                            rec_model = "mean_rec", # recruitment model
                            init_age_strc = 1, # geometric series to derive age structure
                            
                            # Parameter Fixing
                            sigmaR_spec = 'fix'
                            )

input_list$map$ln_sigmaR # both values are fixed and not estimated (specified as factor(rep(NA, 2)))
input_list$par$ln_sigmaR # ln_sigmaR is then fixed at the default starting value
```

To fix *ln_sigmaR* at a specific value rather than the default, supply the desired starting value at the same time:

```{r}
input_list <- Setup_Mod_Rec(input_list = input_list, # input data list from above
                            # Model options
                            do_rec_bias_ramp = 0, # don't do bias ramp 
                            sigmaR_switch = as.integer(length(1960:1975)), # when to switch from early to late sigmaR
                            dont_est_recdev_last = 1, # don't estimate last recruitment deviate
                            rec_model = "mean_rec", # recruitment model
                            init_age_strc = 1, # geometric series to derive age structure
                            
                            # Parameter Fixing
                            sigmaR_spec = 'fix',
  ln_sigmaR = array(log(1.5), dim = c(2, input_list$data$n_pop, input_list$data$n_regions))  # starting values for early and late sigmaR
  )

input_list$map$ln_sigmaR # both values are fixed and not estimated (specified as factor(rep(NA, 2)))
input_list$par$ln_sigmaR # ln_sigmaR is then fixed at the user specified starting value
```

Not all parameters have a convenience argument like `sigmaR_spec = "fix"`. For example, *ln_global_R0* does not. In such cases, you can fix the parameter manually by setting its map entry to `NA` and assigning the desired value in the parameter list. Remember: the map entry fixes the parameter at whatever is currently in `input_list$par`, so always set the starting value first.

```{r}
input_list <- Setup_Mod_Rec(input_list = input_list, # input data list from above
                            # Model options
                            do_rec_bias_ramp = 0, # don't do bias ramp 
                            sigmaR_switch = as.integer(length(1960:1975)), # when to switch from early to late sigmaR
                            dont_est_recdev_last = 1, # don't estimate last recruitment deviate
                            rec_model = "mean_rec", # recruitment model
                            init_age_strc = 1, # geometric series to derive age structure
                            )

input_list$map$ln_global_R0 <- factor(NA)
input_list$par$ln_global_R0 <- log(30)
```

### Sharing Parameters

The `Setup_xxx` functions also support sharing parameters across model partitions via built-in convenience arguments. These cover common sharing structures (e.g., sharing selectivity across sexes or regions), but for finer control, such as sharing a subset of parameters while estimating others independently, users can construct the map manually.

Below, we demonstrate both approaches using fishery selectivity as an example. We define two fishery fleets: Fleet 1 uses a logistic selectivity model (2 parameters: *a50* and slope), and Fleet 2 uses a gamma dome-shaped model (2 parameters: *amax* and slope).

#### Built-in sharing: estimate all parameters independently

We first use `fish_fixed_sel_pars_spec = "est_all"` to estimate all selectivity parameters independently across all model partitions. With 2 sexes, 2 parameters per fleet, and 2 fleets, we expect 8 unique values in the map.

```{r}
input_list <- SPoRC::Setup_Mod_Fishsel_and_Q(input_list = input_list,

                                      # Model options
                                      cont_tv_fish_sel = c("none_Fleet_1", "none_Fleet_2"),
                                      fish_sel_blocks = c("none_Fleet_1", "none_Fleet_2"),
                                      fish_sel_model = c("logist1_Fleet_1", "gamma_Fleet_2"),
                                      fish_q_blocks = c("none_Fleet_1", "none_Fleet_2"),
                                      fish_q_spec = c("fix", "fix"),

                                      # Share selectivity parameters across all partitions
                                      fish_fixed_sel_pars_spec = c("est_all", "est_all"))

input_list$map$fish_fixed_sel_pars # 8 unique numbers, 4 for each sex and fleet combination
```

#### Built-in sharing: share parameters across sexes

Using `fish_fixed_sel_pars_spec = "est_shared_s"` links selectivity parameters between sexes within each fleet. The same two parameters are used for both sexes, so we expect only 4 unique values (2 per fleet).

```{r}
input_list <- SPoRC::Setup_Mod_Fishsel_and_Q(input_list = input_list,

                                      # Model options
                                      cont_tv_fish_sel = c("none_Fleet_1", "none_Fleet_2"),
                                      fish_sel_blocks = c("none_Fleet_1", "none_Fleet_2"),
                                      fish_sel_model = c("logist1_Fleet_1", "gamma_Fleet_2"),
                                      fish_q_blocks = c("none_Fleet_1", "none_Fleet_2"),
                                      fish_q_spec = c("fix", "fix"),

                                      # Share selectivity parameters across all partitions
                                      fish_fixed_sel_pars_spec = c("est_shared_s", "est_shared_s"))

input_list$map$fish_fixed_sel_pars # 4 unique numbers, 2 for each sex and fleet combination
```

#### Manual sharing: fine-grained control

For sharing structures not supported by the convenience arguments, the map can be constructed manually. In this example, we want:

- Fleet 1 (logistic): *a50* is sex-specific, but the slope is shared across sexes
- Fleet 2 (gamma): all parameters are sex-specific

We start from `est_all` to get the correctly dimensioned parameter array, then overwrite the map. The parameter array `fish_fixed_sel_pars` has dimensions `[n_regions, n_max_sel_pars, n_max_sel_blocks, n_sexes, n_fish_fleets]`, in this example `1 × 2 × 1 × 2 × 2`. The table below summarizes the intended sharing structure before we implement it:

| Fleet | Parameter | Female index | Male index | Shared? |
|-------|-----------|-------------|------------|---------|
| 1 | *a50* | 1 | 2 | No |
| 1 | slope | 3 | 3 | Yes |
| 2 | *amax* | 4 | 5 | No |
| 2 | slope | 6 | 7 | No |

```{r}
input_list <- Setup_Mod_Fishsel_and_Q(
  input_list = input_list,
  cont_tv_fish_sel = c("none_Fleet_1", "none_Fleet_2"),
  fish_sel_blocks = c("none_Fleet_1", "none_Fleet_2"),
  fish_sel_model = c("logist1_Fleet_1", "gamma_Fleet_2"),
  fish_q_blocks = c("none_Fleet_1", "none_Fleet_2"),
  fish_q_spec = c("fix", "fix"),
  fish_fixed_sel_pars_spec = c("est_all", "est_all")  # start with all independent
)

# Extract the parameter array to use as a mapping template
map_fish_fixed_sel_pars <- input_list$par$fish_fixed_sel_pars

# Fleet 1 (logistic): a50 sex-specific, slope shared
map_fish_fixed_sel_pars[1, 1, 1, , 1] <- c(1, 2)  # a50: unique per sex
map_fish_fixed_sel_pars[1, 2, 1, , 1] <- c(3, 3)  # slope: same integer = shared

# Fleet 2 (gamma): all parameters sex-specific
map_fish_fixed_sel_pars[1, 1, 1, , 2] <- c(4, 5)  # amax: unique per sex
map_fish_fixed_sel_pars[1, 2, 1, , 2] <- c(6, 7)  # slope: unique per sex

# Flatten to a factor vector and assign to the map
input_list$map$fish_fixed_sel_pars <- factor(as.vector(map_fish_fixed_sel_pars))

# Inspect: 7 unique integers (slope for Fleet 1 appears twice with the same value)
input_list$map$fish_fixed_sel_pars
```

This custom mapping approach gives full control over parameter sharing structures beyond what the high-level convenience arguments provide. The key principle is always the same: positions in the flattened map with the same integer are estimated as a single shared parameter, while `NA` positions are fixed at their value in `input_list$par`.
