---
title: "Defining Priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k_defining_priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In some cases, prior information might be available for use to inform model estimation. Priors can either serve as informative priors that have a strong influence on model estimation (e.g., natural mortality, catchability), but can also serve as regularizing priors that help stabilize model estimation. These regularizing priors can be helpful particularly for stabilizing tag reporting, movement, and selectivity parameters because some parameters are not well informed by data (e.g., slope of selectivity when it approaches infinity) such that the gradient is zero with respect to the likelihood surface.

In `SPoRC`, priors can be placed on several parameters. These include:

1. Steepness,
2. Natural Mortality,
3. Catchability,
4. Selectivity,
5. Movement,
6. Tag Reporting Rates, 
7. Recruitment Regional, Seasonal Proportions, and R0,
8. Stray rates

Two penalties are specified differently from the priors above, without a
`Use_*_prior` toggle or a hyperparameter dataframe, and are covered here as well:
selectivity smoothness, and a stock-recruit curve fitted against the recruitment
series.


In the following, we will demonstrate how priors can be specified for each of these parameters. For further mathematical details on the formulation of these priors, refer to the vignette describing model equations.

Let us first load in any required packages, data we may use, and define model dimensions.

```{r setup}
library(SPoRC) 
data("sgl_rg_sable_data")

input_list <- Setup_Mod_Dim(years = 1:length(sgl_rg_sable_data$years),
                            ages = 1:length(sgl_rg_sable_data$ages),
                            lens = seq(41, 99, 2),
                            n_regions = 1,
                            n_sexes = sgl_rg_sable_data$n_sexes,
                            n_fish_fleets = sgl_rg_sable_data$n_fish_fleets,
                            n_srv_fleets = sgl_rg_sable_data$n_srv_fleets,
                            n_seas = 1,
                            n_pop = sgl_rg_sable_data$n_pop,
                            verbose = TRUE)
```

# Steepness 
To specify priors on the steepness of a Beverton-Holt stock-recruitment relationship, we use the `Setup_Mod_Rec` function. First, we define a dataframe that includes the population (`pop`), region, mean (`mu`), and standard deviation (`sd`) of the steepness prior. These values are specified in normal space but are internally transformed to a beta distribution bounded between 0.2 and 1. To activate steepness priors, set `Use_h_prior = 1` and provide the dataframe to the `h_prior` argument:

```{r}
steepness_prior <- data.frame(
  pop    = 1,   # population 1
  region = 1,   # region 1
  mu     = 0.6, # prior mean
  sd     = 0.2  # prior sd
)

input_list <- Setup_Mod_Rec(
  input_list = input_list,
  rec_model = "bh_rec",
  Use_h_prior = 1,
  h_prior = steepness_prior,
  h_spec = "est_shared_pop_r"
)
```

In a spatial model where region-specific steepness priors are required, the prior dataframe can include multiple entries. Each row corresponds to a population-region combination along with its associated mean and standard deviation:

```{r}
steepness_prior <- data.frame(
  pop    = 1,
  region = c(1, 2, 3),
  mu     = c(0.6, 0.8, 0.3),
  sd     = c(0.2, 0.2, 0.2)
)

steepness_prior
```

# Natural Mortality
To define natural mortality priors, we use the `Setup_Mod_Biologicals` function. Setting `Use_M_prior = 1` activates the prior, and a dataframe of values is provided via the `M_prior` argument. Each row corresponds to a specific combination of population, region, year, age, and sex blocks, along with their associated mean and standard deviation.

```{r}
M_prior <- data.frame( 
  popblk    = 1,
  regionblk = 1,
  yearblk   = 1,
  ageblk    = 1,
  sexblk    = c(1, 2),
  mu        = 0.085,
  sd        = 0.1
)

input_list <- Setup_Mod_Biologicals(
  input_list = input_list,
  WAA          = sgl_rg_sable_data$WAA,
  MatAA        = sgl_rg_sable_data$MatAA,
  AgeingError  = as.matrix(sgl_rg_sable_data$age_error),
  SizeAgeTrans = sgl_rg_sable_data$SizeAgeTrans,
  Use_M_prior  = 1,
  M_prior      = M_prior,
  M_spec       = "est_ln_M",
  M_sexblk_spec = list(1, 2)
)
```

# Catchability
Defining catchability priors for the fishery and survey follows the same process. For demonstration purposes, the following section will illustrate how catchability priors can be defined for the survey using the `Setup_Mod_Srvsel_and_Q` function. A dataframe of catchability priors is first defined that determines the region, fleet, and block to which the prior is applied. The dataframe also includes columns defining the mean (`mu`) and standard deviation (`sd`) of the catchability prior. These priors are provided in normal space, but are internally transformed to a lognormal density. Setting `Use_srv_q_prior = 1` activates the prior.

```{r}
srv_q_prior <- data.frame(
  region = 1,
  fleet  = c(1, 2, 3),
  block  = 1,
  mu     = c(6, 0.5, 4),
  sd     = c(0.1, 0.35, 0.1)
)

srv_q_prior

input_list <- Setup_Mod_Srvsel_and_Q(
  input_list = input_list,
  cont_tv_srv_sel        = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_blocks         = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_model          = c("logist1_Fleet_1", "exponential_Fleet_2", "logist1_Fleet_3"),
  srv_q_blocks           = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_fixed_sel_pars_spec = c("est_all", "est_all", "est_all"),
  srv_q_spec             = c("est_all", "est_all", "est_all"),
  Use_srv_q_prior        = 1,
  srv_q_prior            = srv_q_prior
)
```

# Selectivity
Defining selectivity priors for the fishery and survey follows the same process. Here, we will focus on defining selectivity priors for the survey for brevity, using the `Setup_Mod_Srvsel_and_Q` function. We first define a dataframe with the following columns: `region`, `par`, `block`, `sex`, `fleet`, `mu`, and `sd`, which defines the region, parameter, time block, sex, and fleet combination to apply the prior to. Given the complexity of the number of selectivity parameters that can be estimated across model partitions (regions, sexes, blocks, and fleets), users will need to be cautious about how the selectivity prior dataframe is defined to ensure they are correctly applied.

In the following, we will apply priors to the logistic selectivity parameters for both sexes in fleet 1, and priors to the power parameter describing exponential selectivity in fleet 2.

```{r}
srv_selex_prior <- data.frame(
  region = 1,
  par    = c(1, 2, 1, 2, 1, 1),
  sex    = c(1, 1, 2, 2, 1, 2),
  fleet  = c(1, 1, 1, 1, 2, 2),
  block  = 1,
  mu     = rep(0, 6),
  sd     = rep(5, 6)
)

srv_selex_prior
```

In the example provided above, the first row of the dataframe indicates that a prior is applied to region 1, parameter 1, sex 1, fleet 1, and block 1, which corresponds to the $b_{50}$ parameter for logistic selectivity specified for the first survey fleet. The last row indicates that a prior is applied to region 1, parameter 1, sex 2, fleet 2, and block 1, which corresponds to the male $\phi$ power parameter for exponential selectivity specified for the second survey fleet.

Next, we can activate selectivity priors by setting `Use_srv_selex_prior = 1` and providing the associated selectivity prior dataframe to `srv_selex_prior`.

```{r}
input_list <- Setup_Mod_Srvsel_and_Q(
  input_list = input_list,
  cont_tv_srv_sel         = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_blocks          = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_model           = c("logist1_Fleet_1", "exponential_Fleet_2", "logist1_Fleet_3"),
  srv_q_blocks            = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_fixed_sel_pars_spec = c("est_all", "est_all", "est_all"),
  srv_q_spec              = c("est_all", "est_all", "est_all"),
  Use_srv_selex_prior     = 1,
  srv_selex_prior         = srv_selex_prior
)
```

## Selectivity smoothness penalties

Separately from priors on selectivity parameters, `SPoRC` can penalize the shape
of the realized selectivity curve. These act on $\log(\text{Sel})$ after every
transformation, so they apply whatever functional form produced the curve, and
they are specified in `Setup_Mod_Weighting` rather than in the selectivity setup.
Passing them to `Setup_Mod_Fishsel_and_Q` or `Setup_Mod_Srvsel_and_Q` leaves them
read as starting values and every penalty evaluates to zero.

`fish_sel_pen_wts`, `ret_sel_pen_wts` and `srv_sel_pen_wts` each take one named
list per fleet, or a single named list shared by every fleet. Six terms are
available and anything left out is zero, so a fleet names only what it
constrains:

| Term | Penalizes |
|---|---|
| `smooth_bin_curve` | second difference across bins, curvature |
| `smooth_bin_diff` | first difference across bins, both directions |
| `smooth_yr_diff` | first difference between years, a random walk |
| `smooth_yr_curve` | second difference between years |
| `smooth_dome` | decreases across bins only, so a flat or rising curve is free |
| `smooth_mean_center` | the per-year mean of log selectivity, toward zero |

Each weight is either a scalar used in every year, or a vector with one value per
model year, which lets a term act only in the years selectivity is allowed to
change. Years given zero are skipped. Three further settings sit alongside the
six. `bin_range` is a length two vector giving the first and last bin the terms
run over, defaulting to every bin. `yr_diff_ref` is a value to hold the walk's
first penalized year against, which anchors an otherwise free series before the
data begin.

`normalize`, `TRUE` by default, divides a penalty by the number of terms in its
sum, so it becomes a mean squared difference rather than a total: the two
bin-wise terms divide by the number of bins in `bin_range` and the two year-wise
terms divide by the number of model years. A weight then means the same thing
whatever the bin or year count, where `FALSE` makes a longer series cost
proportionally more at the same weight. It can be a single value for all terms,
or a list naming them individually. `smooth_dome` and `smooth_mean_center` are
never normalized, the first because it sums only the bins that decrease and the
second because its statistic is already a mean.

```{r, eval = FALSE}
# One fleet, curvature and dome on ages 4 to 10, and a random walk that acts
# only in the years the curve is allowed to move.
yr_wt <- rep(0, n_yrs)
yr_wt[block_change_years] <- 1 / (2 * 0.2^2)   # a walk with sigma = 0.2

input_list <- Setup_Mod_Weighting(
  input_list = input_list,
  fish_sel_pen_wts = list(list(
    smooth_bin_curve = 0.5,
    smooth_dome      = 3.1,
    smooth_yr_diff   = yr_wt,
    bin_range        = c(4, 10),
    normalize        = FALSE
  ))
)
```

With `normalize = FALSE`, a first-difference weight of $1/(2\sigma_{y}^{2})$
makes the year term exactly the negative log kernel of a random walk with
standard deviation $\sigma_{y}$, written on the realized curve rather than on
deviation parameters. The equations for all six are in
`vignette("c_model_equations")`.

# Movement
Defining priors for movement, tag reporting rates, and recruitment proportions (see sections below) is best illustrated with a spatial model. Here, we will load in data from the multi-region sablefish case study.

```{r}
data(mlt_rg_sable_data)

input_list <- Setup_Mod_Dim(years = 1:length(mlt_rg_sable_data$years),
                            ages = 1:length(mlt_rg_sable_data$ages),
                            lens = mlt_rg_sable_data$lens,
                            n_regions = mlt_rg_sable_data$n_regions,
                            n_sexes = mlt_rg_sable_data$n_sexes,
                            n_fish_fleets = mlt_rg_sable_data$n_fish_fleets,
                            n_srv_fleets = mlt_rg_sable_data$n_srv_fleets,
                            n_seas = 1,
                            n_pop = mlt_rg_sable_data$n_pop,
                            verbose = TRUE)
```

Because movement parameters are correlated and must sum to 1, `SPoRC` uses a Dirichlet prior to model movement probabilities. Movement priors are enabled by setting `Use_Movement_Prior = 1`. The `Movement_prior` argument must be provided as a dataframe where each row corresponds to a combination of population, origin region, year, age, season, and sex indices, and includes a list-column `alpha` specifying the Dirichlet concentration parameters that define the prior over movement probabilities to all destination regions.

```{r}
prior <- expand.grid(
  pop         = 1,
  region_from = 1:input_list$data$n_regions,
  year        = 1,
  age         = 1,
  seas        = 1,
  sex         = 1,
  alpha       = I(list(rep(2, input_list$data$n_regions)))
)

input_list <- Setup_Mod_Movement(
  input_list           = input_list,
  Movement_ageblk_spec = "constant",
  Movement_yearblk_spec = "constant",
  Movement_sexblk_spec = "constant",
  Use_Movement_Prior   = 1,
  Movement_prior       = prior
)
```

# Tag Reporting Rates
Defining priors for tag reporting rates is done by constructing a dataframe and passing it to the `Setup_Mod_Tagging` function. The dataframe must contain the following columns: `region`, `block`, `fleet`, `mu`, `sd`, and `type`. These columns define the region and time block the prior applies to, the mean and standard deviation of the prior distribution, and the type of beta prior to use. Two prior types are supported: a symmetric beta prior (`type = 0`), which penalizes extreme values near 0 and 1, and a standard beta prior (`type = 1`), which allows users to specify both the mean and standard deviation directly.

In the example below, symmetric beta priors are applied to two time blocks in region 1. Because the symmetric form does not require a mean, the `mu` column is set to `NA`:

```{r}
tag_prior <- data.frame(
  region = 1,
  block  = c(1, 2),
  fleet  = 1,
  mu     = NA,
  sd     = 5,
  type   = 0
)
tag_prior
```

The first row of the dataframe indicates that a symmetric beta prior is applied to tag reporting rates in region 1 and block 1, with a standard deviation of 5. The second row applies the same prior structure to block 2. Because `type = 0`, the prior is symmetric around 0.5 and discourages extreme reporting rate estimates. If `type = 1` were used instead, the `mu` column would need to be specified with the prior mean.

Tag reporting rate priors are activated by setting `use_conv_tag_fishrep_prior = 1` and providing the dataframe to the `conv_tag_fishrep_prior` argument.

```{r}
input_list <- Setup_Mod_Tagging(
  input_list = input_list,
  use_conv_fish_tagging      = c(1, 0),
  conv_tag_max_liberty       = 15,

  # Data inputs
  conv_tag_release_indicator = mlt_rg_sable_data$conv_tag_release_indicator,
  conv_tagged_fish           = mlt_rg_sable_data$conv_tagged_fish,
  obs_conv_tag_fish_recap    = mlt_rg_sable_data$obs_conv_tag_fish_recap,

  # Model options
  conv_fish_tag_like           = "Poisson",
  use_conv_tag_fishrep_prior   = 1,
  conv_tag_fishrep_prior       = tag_prior,
  init_conv_tag_mort_spec      = "fix",
  conv_tag_shed_spec           = "fix",
  conv_tagrep_spec             = "est_shared_r_f",
  conv_tag_fish_reporting_blocks = c(
    apply(expand.grid(1:input_list$data$n_regions, 1:input_list$data$n_fish_fleets), 1, function(x)
      paste0("Block_1_Year_1-35_Region_", x[1], "_Fleet_", x[2])),
    apply(expand.grid(1:input_list$data$n_regions, 1:input_list$data$n_fish_fleets), 1, function(x)
      paste0("Block_2_Year_36-terminal_Region_", x[1], "_Fleet_", x[2]))
  ),

  # Starting / fixed values
  ln_init_conv_tag_mort = log(0.1),
  ln_conv_tag_shed      = log(0.02)
)
```

# Recruitment Proportions and R0

## Recruitment Regional Apportionment
In the context of a spatial model, Dirichlet priors can also be placed on recruitment proportions. This is done through the `Setup_Mod_Rec` function by setting `use_rec_region_prop_prior = 1` and providing a dataframe to `rec_region_prop_prior`. Each row of the dataframe corresponds to a population and contains a list-column `alpha` specifying the Dirichlet concentration parameters for movement among destination regions. If a uniform Dirichlet prior is desired, all concentration parameters can be set to the same value. Providing different values for each region will apply a weighted Dirichlet prior.

```{r}
input_list <- Setup_Mod_Rec(
  input_list = input_list,
  do_rec_bias_ramp = 0,
  rec_model = "mean_rec",
  use_rec_region_prop_prior = 1,
  rec_region_prop_prior = data.frame(
    pop   = 1,
    alpha = I(list(rep(5, mlt_rg_sable_data$n_regions)))
  )
)
```

## Recruitment Seasonal Apportionment
Likewise, priors on recruitment seasonal apportionment can be applied in an analogous manner. This is done by setting `use_rec_seas_prop_prior = 1` and providing a dataframe to `rec_seas_prop_prior`. Note that seasonal recruitment proportion priors are only relevant when `n_seas > 1` and `use_fixed_rec_seas_prop = 0` (i.e., seasonal proportions are being estimated rather than fixed). Each row of the dataframe corresponds to a population and contains a list-column `alpha` specifying the Dirichlet concentration parameters for the seasonal apportionment of recruitment.

## R0
Lognormal priors on $R_0$ can be applied per population via `use_r0_prior` and `r0_prior`.
Each row of `r0_prior` corresponds to a population, with `mu` specifying the prior mean on
the natural scale and `sd` the standard deviation on the log scale.

```{r}
input_list <- Setup_Mod_Rec(
  input_list = input_list,
  use_r0_prior = 1,
  do_rec_bias_ramp = 0,
  rec_model = "mean_rec",
  r0_prior = data.frame(
    pop = 1,
    mu  = 1e6,
    sd  = 1
  )
)
```

## Stock-Recruit Residual

Under `rec_model = "mean_rec"`, a stock-recruit curve can be fitted as a penalty rather than used to generate recruitment. That penalty is a prior in the same sense as the ones above, a normal density placed on the log residual $\log R_y - \log \hat{R}_y$, but it is specified differently from everything else in this vignette: there is no `Use_*_prior` toggle and no hyperparameter dataframe. Instead, `sr_penalty` names the curve, `sr_pen_sigma` gives the residual's standard deviation, `sr_pen_yrs` gives the years it is summed over, and `sr_R0_spec` sets where the curve's scale comes from. The curve's steepness remains an ordinary parameter, so `Use_h_prior` and `h_prior` described above apply to it exactly as they do to a curve that generates recruitment.

```{r, eval = FALSE}
input_list <- Setup_Mod_Rec(
  input_list   = input_list,
  rec_model    = "mean_rec",       # the curve is fitted, not used to generate recruitment
  sr_penalty   = "bh",             # "none" (default), "bh", or "ricker"
  sr_pen_sigma = 0.5,              # standard deviation of the log residual
  sr_pen_yrs   = 1980:2015,        # years summed over; defaults to all with a lagged SSB
  sr_R0_spec   = "shared"          # "shared", "est", or "rinit"
)
```

`sr_R0_spec` sets where the curve takes its scale: `"shared"` reuses
`ln_global_R0`, which under mean recruitment is the recruitment level itself,
`"est"` gives the curve its own estimated `ln_sr_R0`, and `"rinit"` takes it from
`ln_rinit` so one parameter sets both the unfished age structure and the curve,
which requires `use_rinit = 1`. The fitted scale is reported as `sr_R0` and the
curve's prediction as `SR_pred`.

## Stray Rates

When stray rates are estimated (`stray_rate_spec` is not `"fix"`), a beta prior should almost always be used to keep estimation stable. Estimation of stray rates are only applicable when `n_pop > 1`. The prior is specified via a dataframe passed to `stray_rate_prior`, with columns `pop`, `block`, `mu`, and `sd`. The mean (`mu`) and standard deviation (`sd`) are supplied in probability space and are internally converted to beta shape parameters via method-of-moments. Note that `sd` must satisfy $\sigma < \sqrt{\mu(1-\mu)}$ to ensure a valid (unimodal) beta density.

In the example below, stray rates are estimated independently per population using a single time block, with population 1 given a prior centered at 0.1 and population 2 centered at 0.5:

```{r, eval = FALSE}

input_list$data$n_pop <- 2 # modify to two populations for demo purposes

stray_prior <- data.frame(
  pop   = c(1, 2),
  block = c(1, 1),
  mu    = c(0.1, 0.5),
  sd    = c(0.1, 0.1)
)

input_list <- Setup_Mod_Rec(
  input_list       = input_list,
  do_rec_bias_ramp = 0,
  rec_model        = "bh_rec",
  rec_dd           = "local",
  stray_rate_blocks = c(
    "Block_1_Year_1-terminal_Pop_1",
    "Block_1_Year_1-terminal_Pop_2"
  ),
  stray_rate_spec      = "est_all",
  use_fixed_stray_rate = 0,
  use_stray_rate_prior = 1,
  stray_rate_prior     = stray_prior
)
```

Tighter values of `sd` produce a stronger regularizing effect and are recommended when data contain little information about cross-population spawning contributions. If `sd` is set too large relative to `mu * (1 - mu)`, the beta density becomes U-shaped (mass near 0 and 1), which can cause numerical instability during optimization; `SPoRC` guards against this internally by squishing the logistic transform away from the boundaries, but specifying a sensible `sd` is still good practice.

## Time-varying stray rates

Stray rates can also be allowed to vary over time by defining multiple blocks. The `stray_rate_blocks` argument follows the same blocking syntax used elsewhere in `SPoRC`. For example, to allow stray rates to differ between an early and late period for each population:

```{r, eval = FALSE}
stray_prior_tv <- data.frame(
  pop   = c(1, 1, 2, 2),
  block = c(1, 2, 1, 2),
  mu    = c(0.1, 0.2, 0.4, 0.5),
  sd    = c(0.05, 0.05, 0.1, 0.1)
)

input_list <- Setup_Mod_Rec(
  input_list       = input_list,
  do_rec_bias_ramp = 0,
  rec_model        = "bh_rec",
  rec_dd           = "local",
  stray_rate_blocks = c(
    "Block_1_Year_1-20_Pop_1",
    "Block_2_Year_21-terminal_Pop_1",
    "Block_1_Year_1-20_Pop_2",
    "Block_2_Year_21-terminal_Pop_2"
  ),
  stray_rate_spec      = "est_all",
  use_fixed_stray_rate = 0,
  use_stray_rate_prior = 1,
  stray_rate_prior     = stray_prior_tv
)
```

Each unique block index in `stray_rate_blocks` requires a corresponding row in `stray_rate_prior`, one per population × block combination.

