---
title: "Defining Priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k_defining_priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In some cases, prior information might be available for use to inform model estimation. Priors can either serve as informative priors that have a strong influence on model estimation (e.g., natural mortality, catchability), but can also serve as regularizing priors that help stabilize model estimation. These regularizing priors can be helpful particularly for stabilizing tag reporting, movement, and selectivity parameters because some parameters are not well informed by data (e.g., slope of selectivity when it approaches infinity) such that the gradient is zero with respect to the likelihood surface.

In `SPoRC`, priors can be placed on several parameters. These include:

1. Steepness,
2. Natural Mortality,
3. Catchability,
4. Selectivity,
5. Movement,
6. Tag Reporting Rates, 
7. Recruitment Regional and Seasonal Proportions,
8. Stray rates


In the following, we will demonstrate how priors can be specified for each of these parameters. For further mathematical details on the formulation of these priors, refer to the vignette describing model equations.

Let us first load in any required packages, data we may use, and define model dimensions.

```{r setup}
library(SPoRC) 
data("sgl_rg_sable_data")

input_list <- Setup_Mod_Dim(years = 1:length(sgl_rg_sable_data$years),
                            ages = 1:length(sgl_rg_sable_data$ages),
                            lens = seq(41, 99, 2),
                            n_regions = 1,
                            n_sexes = sgl_rg_sable_data$n_sexes,
                            n_fish_fleets = sgl_rg_sable_data$n_fish_fleets,
                            n_srv_fleets = sgl_rg_sable_data$n_srv_fleets,
                            n_seas = 1,
                            n_pop = sgl_rg_sable_data$n_pop,
                            verbose = TRUE)
```

# Steepness 
To specify priors on the steepness of a Beverton-Holt stock-recruitment relationship, we use the `Setup_Mod_Rec` function. First, we define a dataframe that includes the population (`pop`), region, mean (`mu`), and standard deviation (`sd`) of the steepness prior. These values are specified in normal space but are internally transformed to a beta distribution bounded between 0.2 and 1. To activate steepness priors, set `Use_h_prior = 1` and provide the dataframe to the `h_prior` argument:

```{r}
steepness_prior <- data.frame(
  pop    = 1,   # population 1
  region = 1,   # region 1
  mu     = 0.6, # prior mean
  sd     = 0.2  # prior sd
)

input_list <- Setup_Mod_Rec(
  input_list = input_list,
  rec_model = "bh_rec",
  Use_h_prior = 1,
  h_prior = steepness_prior,
  h_spec = "est_shared_pop_r"
)
```

In a spatial model where region-specific steepness priors are required, the prior dataframe can include multiple entries. Each row corresponds to a population–region combination along with its associated mean and standard deviation:

```{r}
steepness_prior <- data.frame(
  pop    = 1,
  region = c(1, 2, 3),
  mu     = c(0.6, 0.8, 0.3),
  sd     = c(0.2, 0.2, 0.2)
)

steepness_prior
```

# Natural Mortality
To define natural mortality priors, we use the `Setup_Mod_Biologicals` function. Setting `Use_M_prior = 1` activates the prior, and a dataframe of values is provided via the `M_prior` argument. Each row corresponds to a specific combination of population, region, year, age, and sex blocks, along with their associated mean and standard deviation.

```{r}
M_prior <- data.frame( 
  popblk    = 1,
  regionblk = 1,
  yearblk   = 1,
  ageblk    = 1,
  sexblk    = c(1, 2),
  mu        = 0.085,
  sd        = 0.1
)

input_list <- Setup_Mod_Biologicals(
  input_list = input_list,
  WAA          = sgl_rg_sable_data$WAA,
  MatAA        = sgl_rg_sable_data$MatAA,
  AgeingError  = as.matrix(sgl_rg_sable_data$age_error),
  SizeAgeTrans = sgl_rg_sable_data$SizeAgeTrans,
  Use_M_prior  = 1,
  M_prior      = M_prior,
  M_spec       = "est_ln_M",
  M_sexblk_spec = list(1, 2)
)
```

# Catchability
Defining catchability priors for the fishery and survey follows the same process. For demonstration purposes, the following section will illustrate how catchability priors can be defined for the survey using the `Setup_Mod_Srvsel_and_Q` function. A dataframe of catchability priors is first defined that determines the region, fleet, and block to which the prior is applied. The dataframe also includes columns defining the mean (`mu`) and standard deviation (`sd`) of the catchability prior. These priors are provided in normal space, but are internally transformed to a lognormal density. Setting `Use_srv_q_prior = 1` activates the prior.

```{r}
srv_q_prior <- data.frame(
  region = 1,
  fleet  = c(1, 2, 3),
  block  = 1,
  mu     = c(6, 0.5, 4),
  sd     = c(0.1, 0.35, 0.1)
)

srv_q_prior

input_list <- Setup_Mod_Srvsel_and_Q(
  input_list = input_list,
  cont_tv_srv_sel        = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_blocks         = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_model          = c("logist1_Fleet_1", "exponential_Fleet_2", "logist1_Fleet_3"),
  srv_q_blocks           = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_fixed_sel_pars_spec = c("est_all", "est_all", "est_all"),
  srv_q_spec             = c("est_all", "est_all", "est_all"),
  Use_srv_q_prior        = 1,
  srv_q_prior            = srv_q_prior
)
```

# Selectivity
Defining selectivity priors for the fishery and survey follows the same process. Here, we will focus on defining selectivity priors for the survey for brevity, using the `Setup_Mod_Srvsel_and_Q` function. We first define a dataframe with the following columns: `region`, `par`, `block`, `sex`, `fleet`, `mu`, and `sd`, which defines the region, parameter, time block, sex, and fleet combination to apply the prior to. Given the complexity of the number of selectivity parameters that can be estimated across model partitions (regions, sexes, blocks, and fleets), users will need to be cautious about how the selectivity prior dataframe is defined to ensure they are correctly applied.

In the following, we will apply priors to the logistic selectivity parameters for both sexes in fleet 1, and priors to the power parameter describing exponential selectivity in fleet 2.

```{r}
srv_selex_prior <- data.frame(
  region = 1,
  par    = c(1, 2, 1, 2, 1, 1),
  sex    = c(1, 1, 2, 2, 1, 2),
  fleet  = c(1, 1, 1, 1, 2, 2),
  block  = 1,
  mu     = rep(0, 6),
  sd     = rep(5, 6)
)

srv_selex_prior
```

In the example provided above, the first row of the dataframe indicates that a prior is applied to region 1, parameter 1, sex 1, fleet 1, and block 1, which corresponds to the $b_{50}$ parameter for logistic selectivity specified for the first survey fleet. The last row indicates that a prior is applied to region 1, parameter 1, sex 2, fleet 2, and block 1, which corresponds to the male $\phi$ power parameter for exponential selectivity specified for the second survey fleet.

Next, we can activate selectivity priors by setting `Use_srv_selex_prior = 1` and providing the associated selectivity prior dataframe to `srv_selex_prior`.

```{r}
input_list <- Setup_Mod_Srvsel_and_Q(
  input_list = input_list,
  cont_tv_srv_sel         = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_blocks          = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_sel_model           = c("logist1_Fleet_1", "exponential_Fleet_2", "logist1_Fleet_3"),
  srv_q_blocks            = c("none_Fleet_1", "none_Fleet_2", "none_Fleet_3"),
  srv_fixed_sel_pars_spec = c("est_all", "est_all", "est_all"),
  srv_q_spec              = c("est_all", "est_all", "est_all"),
  Use_srv_selex_prior     = 1,
  srv_selex_prior         = srv_selex_prior
)
```

# Movement
Defining priors for movement, tag reporting rates, and recruitment proportions (see sections below) is best illustrated with a spatial model. Here, we will load in data from the multi-region sablefish case study.

```{r}
data(mlt_rg_sable_data)

input_list <- Setup_Mod_Dim(years = 1:length(mlt_rg_sable_data$years),
                            ages = 1:length(mlt_rg_sable_data$ages),
                            lens = mlt_rg_sable_data$lens,
                            n_regions = mlt_rg_sable_data$n_regions,
                            n_sexes = mlt_rg_sable_data$n_sexes,
                            n_fish_fleets = mlt_rg_sable_data$n_fish_fleets,
                            n_srv_fleets = mlt_rg_sable_data$n_srv_fleets,
                            n_seas = 1,
                            n_pop = mlt_rg_sable_data$n_pop,
                            verbose = TRUE)
```

Because movement parameters are correlated and must sum to 1, `SPoRC` uses a Dirichlet prior to model movement probabilities. Movement priors are enabled by setting `Use_Movement_Prior = 1`. The `Movement_prior` argument must be provided as a dataframe where each row corresponds to a combination of population, origin region, year, age, season, and sex indices, and includes a list-column `alpha` specifying the Dirichlet concentration parameters that define the prior over movement probabilities to all destination regions.

```{r}
prior <- expand.grid(
  pop         = 1,
  region_from = 1:input_list$data$n_regions,
  year        = 1,
  age         = 1,
  seas        = 1,
  sex         = 1,
  alpha       = I(list(rep(2, input_list$data$n_regions)))
)

input_list <- Setup_Mod_Movement(
  input_list           = input_list,
  Movement_ageblk_spec = "constant",
  Movement_yearblk_spec = "constant",
  Movement_sexblk_spec = "constant",
  Use_Movement_Prior   = 1,
  Movement_prior       = prior
)
```

# Tag Reporting Rates
Defining priors for tag reporting rates is done by constructing a dataframe and passing it to the `Setup_Mod_Tagging` function. The dataframe must contain the following columns: `region`, `block`, `fleet`, `mu`, `sd`, and `type`. These columns define the region and time block the prior applies to, the mean and standard deviation of the prior distribution, and the type of beta prior to use. Two prior types are supported: a symmetric beta prior (`type = 0`), which penalizes extreme values near 0 and 1, and a standard beta prior (`type = 1`), which allows users to specify both the mean and standard deviation directly.

In the example below, symmetric beta priors are applied to two time blocks in region 1. Because the symmetric form does not require a mean, the `mu` column is set to `NA`:

```{r}
tag_prior <- data.frame(
  region = 1,
  block  = c(1, 2),
  fleet  = 1,
  mu     = NA,
  sd     = 5,
  type   = 0
)
tag_prior
```

The first row of the dataframe indicates that a symmetric beta prior is applied to tag reporting rates in region 1 and block 1, with a standard deviation of 5. The second row applies the same prior structure to block 2. Because `type = 0`, the prior is symmetric around 0.5 and discourages extreme reporting rate estimates. If `type = 1` were used instead, the `mu` column would need to be specified with the prior mean.

Tag reporting rate priors are activated by setting `use_conv_tag_fishrep_prior = 1` and providing the dataframe to the `conv_tag_fishrep_prior` argument.

```{r}
input_list <- Setup_Mod_Tagging(
  input_list = input_list,
  use_conv_fish_tagging      = c(1, 0),
  conv_tag_max_liberty       = 15,

  # Data inputs
  conv_tag_release_indicator = mlt_rg_sable_data$conv_tag_release_indicator,
  conv_tagged_fish           = mlt_rg_sable_data$conv_tagged_fish,
  obs_conv_tag_fish_recap    = mlt_rg_sable_data$obs_conv_tag_fish_recap,

  # Model options
  conv_fish_tag_like           = "Poisson",
  use_conv_tag_fishrep_prior   = 1,
  conv_tag_fishrep_prior       = tag_prior,
  init_conv_tag_mort_spec      = "fix",
  conv_tag_shed_spec           = "fix",
  conv_tagrep_spec             = "est_shared_r_f",
  conv_tag_fish_reporting_blocks = c(
    apply(expand.grid(1:input_list$data$n_regions, 1:input_list$data$n_fish_fleets), 1, function(x)
      paste0("Block_1_Year_1-35_Region_", x[1], "_Fleet_", x[2])),
    apply(expand.grid(1:input_list$data$n_regions, 1:input_list$data$n_fish_fleets), 1, function(x)
      paste0("Block_2_Year_36-terminal_Region_", x[1], "_Fleet_", x[2]))
  ),

  # Starting / fixed values
  ln_init_conv_tag_mort = log(0.1),
  ln_conv_tag_shed      = log(0.02)
)
```

# Recruitment Proportions

## Recruitment Regional Apportionment
In the context of a spatial model, Dirichlet priors can also be placed on recruitment proportions. This is done through the `Setup_Mod_Rec` function by setting `use_rec_region_prop_prior = 1` and providing a dataframe to `rec_region_prop_prior`. Each row of the dataframe corresponds to a population and contains a list-column `alpha` specifying the Dirichlet concentration parameters for movement among destination regions. If a uniform Dirichlet prior is desired, all concentration parameters can be set to the same value. Providing different values for each region will apply a weighted Dirichlet prior.

```{r}
input_list <- Setup_Mod_Rec(
  input_list = input_list,
  do_rec_bias_ramp = 0,
  rec_model = "mean_rec",
  use_rec_region_prop_prior = 1,
  rec_region_prop_prior = data.frame(
    pop   = 1,
    alpha = I(list(rep(5, mlt_rg_sable_data$n_regions)))
  )
)
```

## Recruitment Seasonal Apportionment
Likewise, priors on recruitment seasonal apportionment can be applied in an analogous manner. This is done by setting `use_rec_seas_prop_prior = 1` and providing a dataframe to `rec_seas_prop_prior`. Note that seasonal recruitment proportion priors are only relevant when `n_seas > 1` and `use_fixed_rec_seas_prop = 0` (i.e., seasonal proportions are being estimated rather than fixed). Each row of the dataframe corresponds to a population and contains a list-column `alpha` specifying the Dirichlet concentration parameters for the seasonal apportionment of recruitment.

## Stray Rates (in development)
