---
title: "Overview of Model Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{t_model_options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

SPoRC models are assembled through a pipeline of `Setup_Mod_*` functions (for the estimation model) and `Setup_Sim_*` functions (for the operating model / simulation). Each function appends to an `input_list` (or `sim_list`) that accumulates data, parameter starting values, and RTMB factor maps. This vignette catalogs every user-facing option across that pipeline. Where both an integer code and a string alias exist, either is accepted. Strings are generally preferred for readability. When `n_sexes > 1`, the first sex index is always female. When `n_pop > 1`, populations follow the order in which they are defined. Throughout this document, the notation `[p × r × y × τ × a × s]` is shorthand for array dimensions: population, region, year, seasons, age, sex.

# Parameter sharing conventions

Many hyperparameters throughout the pipeline (process-error standard
deviations, catchability, natural mortality, tag reporting rates, ...)
are controlled by a `_spec` string argument that follows the same
convention: a single named dimension (e.g. region, season, fleet, sex)
is either estimated independently for every level, shared across some
subset of levels, or fixed. Concretely, for a hyperparameter varying
across dimensions abbreviated `r` (region), `seas` (season), and `f`
(fleet), the accepted strings are:

| String | Meaning |
|---|---|
| `"est_all"` | Independent parameter for every combination of dimensions |
| `"est_shared_r"` | Shared across regions; independent per remaining dimension |
| `"est_shared_seas"` | Shared across seasons |
| `"est_shared_f"` | Shared across fleets |
| `"est_shared_r_seas"`, `"est_shared_r_f"`, `"est_shared_seas_f"` | Shared across the named pair of dimensions |
| `"est_shared_r_seas_f"` | A single parameter shared across everything |
| `"fix"` | Held at its starting value (not estimated) |

Which dimension abbreviations are valid, and in what order they can be
combined, depends on the specific parameter (see each section below for
its abbreviations, e.g. `sigmaC_spec` additionally has a `y` (year)
dimension). `sigmaF_spec`, `sigmaC_spec`, `sigmaR_spec`, and
`Fdev_rho_spec` all follow this convention via a single shared internal
helper (`build_pe_map`/`build_shared_spec_map`), so they behave
identically. Other `_spec` arguments using the same string vocabulary
(e.g. `sigmaC_pop_spec`, and selectivity's `fish_fixed_sel_pars_spec`,
`fishsel_pe_pars_spec`, `fish_sel_devs_spec`, and their retention/survey
equivalents) are implemented separately with their own hand-written
logic, and some additionally support a fleet-sharing escape hatch,
`"est_shared_f_<n>"`, which copies the sharing structure of fleet `<n>`
wholesale rather than collapsing a dimension.

---

# Spatial and Demographic Structure

Everything starts with `Setup_Mod_Dim()` (estimation model) or `Setup_Sim_Dim()` (operating model; simulation). These functions define the skeleton that every subsequent setup call validates against.

Core dimensions

| Argument | Type | Description |
|---|---|---|
| `years` | integer vector | Calendar years included in the assessment (length determines $n_y$) |
| `ages` | integer vector | Modeled age classes; the final element is the plus group |
| `lens` | numeric vector or `NULL` | Length-bin midpoints; `NULL` disables length-based features |
| `n_regions` | integer | Spatial regions |
| `n_sexes` | integer | `1` (aggregated) or `2` (sex-structured) |
| `n_fish_fleets` | integer | Fishery fleets |
| `n_srv_fleets` | integer | Survey fleets |

Multi-population structure

SPoRC decouples biological populations from spatial regions. Populations have their own stock-recruit dynamics and natal assignment; regions define the arena through which all populations move.

| Argument | Type | Description |
|---|---|---|
| `n_pop` | integer | Number of biological populations (default `1`) |
| `natal_region` | integer vector (length `n_pop`) | Maps each population to its home region. Inferred automatically when `n_pop == n_regions` (one-to-one) or `n_regions == 1` (all share region 1). Must be supplied explicitly otherwise |

When `n_pop > n_regions`, multiple populations share a natal region, a contingent-population or mixed-stock structure. When `n_pop < n_regions`, the model represents a single population spread across more regions than there are spawning origins.

Seasonal sub-stepping

| Argument | Type | Description |
|---|---|---|
| `n_seas` | integer | Seasons per year (default `1`) |
| `seasdur` | numeric vector (length `n_seas`) | Duration of each season as a fraction of the year (must sum to 1). Defaults to equal-length seasons |

Within each year, processes execute in sequence: recruitment → movement → mortality (with age advancement at the end of the final season).

Other dimension controls (EM only)

| Argument | Description |
|---|---|
| `n_proj_yrs_devs` | How many projection-year slots to pre-allocate for deviation parameters (`ln_RecDevs`, `move_devs`, selectivity deviations). Default `0` |
| `store_config` | If `TRUE`, archives all `Setup_Mod_*` arguments inside `input_list$config` for reproducibility |

Additional simulation dimensions (OM only)

| Argument | Description |
|---|---|
| `n_sims` | Number of Monte Carlo replicates |
| `n_yrs` | Projection horizon (years) |
| `n_obs_ages` | Number of observed age bins (can differ from `n_ages` when compositions pool ages differently) |
| `run_feedback` | `TRUE` for closed-loop MSE; `FALSE` (default) for open-loop simulation |
| `feedback_start_yr` | First year of feedback when `run_feedback = TRUE` |

---

# Population Initialization

Controlled via `Setup_Mod_Rec()` (argument `init_age_strc`) and `Setup_Sim_Rec()`.

SPoRC offers four methods for deriving equilibrium numbers-at-age, plus a fifth, non-equilibrium option. The equilibrium methods project a constant recruitment ($R_0$, or a separate `ln_rinit` scalar when `use_rinit = 1`) through seasonal mortality, fishing, and (optionally) movement until the age structure stabilizes. After the equilibrium is computed, multiplicative log-scale initial deviations (`ln_InitDevs`) are applied to ages $2, \ldots, A$.

When `use_rinit = 1`, the regional recruitment scalar feeding the equilibrium calculation is bias-corrected: $\zeta_{p,r}\,\text{rinit}_p\exp(-\sigma_{\text{Rec,early}}^2/2)$, treating `ln_rinit` as the median of the assumed lognormal recruitment process and converting to the corresponding mean before it is used as a deterministic (non-stochastic) equilibrium seed. This mirrors the same correction applied to `Setup_Sim_Rec()`'s equivalent `rinit_input` pathway during simulation, so fitted and simulated equilibria remain on a consistent scale.

### Equilibrium solution method (`init_age_strc`)

| Value | String | How the equilibrium is solved |
|---|---|---|
| 0 | `"iterative"` | Brute-force iteration: runs the full seasonal cycle $n_\text{ages} \times 10$ times |
| 1 | `"scalar_no_move"` | Closed-form geometric series assuming no movement at any age. Plus group: $N_{A^+} = N_{A-1} e^{-Z_{A-1}} / (1 - e^{-Z_{A^+}})$ |
| 2 | `"matrix"` | Builds seasonal transition matrices $\mathbf{T}_a = \prod_\tau \mathbf{M}_{a,\tau} \mathbf{S}_{a,\tau}$ that combine movement and survival, then solves $\mathbf{N}_{A^+} = (\mathbf{I} - \mathbf{T}_{A^+})^{-1} \mathbf{T}_{A-1} \mathbf{N}_{A-1}$. Default |
| 3 | `"scalar_plus_only"` | Hybrid: uses the matrix approach for ages below the plus group but switches to the scalar geometric-series for the plus group itself |
| 4 | `"free"` | No equilibrium at all: `ln_InitDevs` *are* the initial log numbers-at-age (ages 2+, apportioned by sex ratio), with age 1 still taken from recruitment |

Recommendations. Use the default `"matrix"` (2) for spatial models; `"scalar_no_move"` (1) is fine, and marginally cheaper, for single-region models. Use `"free"` (4) when the initial age structure carries no information about $R_0$ and should not be pulled toward an equilibrium, matching assessments in which initial numbers-at-age are freely estimated parameters. Note two consequences of `"free"`: the initial condition becomes independent of `init_F_par` and of `ln_rinit`, and the deviations are on the scale of log-*numbers* rather than log-ratios about an equilibrium, so any penalty applied via `equil_init_age_strc` acts as a prior on log initial abundance. Pair `"free"` with `InitDevs_pen_center = "own_mean"` (see Recruitment below) to penalize only the roughness of the initial age structure rather than its level.

For no prior at all, note that `equil_init_age_strc` bundles two decisions that do not move together: `"equil"` (0) neither penalizes nor estimates the deviations, and `"stoch_all"` (2) both estimates and penalizes them. Under `"free"` you usually want the combination it does not offer, estimated and unpenalized, because the deviations are the numbers themselves. Either route reaches it with one extra line, and they give an identical fit:

```{r, eval = FALSE}
# route A: no penalty from the setting, then re-free the deviations it fixed
input_list <- Setup_Mod_Rec(input_list, ..., init_age_strc = "free",
                            equil_init_age_strc = "equil")
input_list$map$ln_InitDevs <- factor(seq_along(input_list$par$ln_InitDevs))

# route B: estimated by the setting, then switch its penalty off
input_list <- Setup_Mod_Rec(input_list, ..., init_age_strc = "free",
                            equil_init_age_strc = "stoch_all")
input_list$data$init_devs_pen_use[] <- 0
```

Either line must come after the setup calls that touch `ln_InitDevs`, since a later call would overwrite an earlier assignment. Leaving it out under `"equil"` is quiet rather than loud: the deviations stay fixed at zero, which under `"free"` is one fish at every initial age, and the model still fits.

### Initial deviation structure (`equil_init_age_strc`)

| Value | String | Which ages receive initial deviations |
|---|---|---|
| 0 | `"equil"` | None, strict equilibrium |
| 1 | `"stoch_no_plus"` | All ages except the plus group. Default |
| 2 | `"stoch_all"` | All ages including the plus group |
| 3 | `"stoch_shared_ages"` | User-defined age sharing via `init_age_devs_shared` |

`init_F_par` (array `[n_regions × n_seas × n_fish_fleets]`) optionally introduces fishing mortality into the equilibrium calculation, producing a fished initial condition. `init_F_form` sets what it means: `"prop"` treats it as a proportion of the mean fishing mortality (inverse-logit scale, bounded to (0,1)), so the initial age structure moves with mean F; `"abs"` treats it as an absolute rate (log scale) that is independent of mean F. `init_F_spec` (`"fix"` or `"est"`) sets whether it is estimated, independently of the form. Prefer `"abs"` when the historical fishing mortality that shaped the initial condition is distinct from the mean F of the modeled period; under `"prop"` one parameter both depletes the initial age structure and scales the F series, and because catch constrains only their product the optimizer can fit catch equally well with a smaller, harder-fished stock. The older `init_F_prop` argument is still accepted and is converted to the `"prop"` form.

---

# Recruitment

Controlled via `Setup_Mod_Rec()` (EM) or `Setup_Sim_Rec()` (OM).

### Stock-recruit function (`rec_model`)

| String | Description |
|---|---|
| `"mean_rec"` | Estimate mean $\ln R_0$ with annual deviations; no SSB feedback |
| `"bh_rec"` | Beverton-Holt: $R = 4hR_0 \cdot \text{SSB} / \left[(1-h)S_0 + (5h-1)\text{SSB}\right]$. Unfished spawning biomass per recruit ($S_0$) is computed internally by projecting a single recruit through all ages and seasons with movement |
| `"ricker_rec"` | Ricker in depletion form: $R = R_0 (S/S_0)\exp(\alpha(1 - S/S_0))$ with $\alpha = \log(4h/(1-h))$ (the "Dorn form" used by the EBS pollock assessment). The curve passes through $(S_0, R_0)$ and shares the Beverton-Holt's compensation ratio at a given $h$, not the textbook $R(0.2S_0) = hR_0$ definition, so steepness values are not interchangeable between `"bh_rec"` and `"ricker_rec"`, and a steepness prior calibrated for one should not be reused for the other |

`SR_ref_yr` sets the year index whose biological inputs (WAA, maturity, $M$, movement) feed the unfished spawning-biomass-per-recruit calculation, and hence $S_0$ and the scale of the stock-recruit curve. Default `1` (first model year, the long-standing behavior); set `length(years)` to condition the curve on terminal biologicals, a convention several assessments use. With time-varying weight-at-age the two choices give different $S_0$; pick whichever the assessment you are bridging or the reference-point convention you follow uses, and keep it consistent with how reference points are computed.

### Density dependence scope (`rec_dd`)

| String | When to use |
|---|---|
| `"local"` | SSB and $S_0$ computed per population and/or per region. Required when `n_pop > 1` with BH recruitment |
| `"global"` | SSB summed across all regions before entering the SRR. Single-population only |

### Spawning timing

| Argument | Description |
|---|---|
| `spawn_seas` | Season index in which spawning occurs |
| `t_spawn` | Fraction of the spawning season elapsed before spawning occurs (0 = start, 0.5 = midpoint) |
| `rec_lag` | Delay (in seasons) between spawning and recruitment entry. `1` (default): recruitment driven by SSB from `rec_lag` seasons prior, may enter in any season. `0`: age-0 recruitment, recruitment driven by that *same* year's own SSB. Since that SSB isn't known until `spawn_seas` is reached, recruits may only enter in `spawn_seas` itself or a later season in the same year (`rec_seas_prop` must be zero for every season before `spawn_seas`), and the recruit age class must have zero maturity everywhere |

### Recruitment allocation

| Argument | Description |
|---|---|
| `rec_region_prop_spec` | How regional recruitment proportions are estimated/fixed |
| `rec_seas_prop_spec` | How seasonal recruitment proportions are estimated. Default `"fix"` (all recruitment enters in season 1) |
| `sexratio_spec` | Sex ratio at recruitment estimation. Default `"fix"` (equal or user-supplied) |
| `sexratio_blocks` | Time blocks for sex-ratio parameters, specified as `"none_Pop_<p>_Region_<r>"` or `"Block_<b>_Year_<s>-<e>_Pop_<p>_Region_<r>"` |

### Recruitment and Initial age deviations

| Argument | Description |
|---|---|
| `RecDevs_spec` | Estimation structure for annual $\ln(\text{RecDevs})$ |
| `InitDevs_spec` | Estimation structure for initial age deviations |
| `InitDevs_sex_spec` | `"est_shared_s"` (default): one initial age curve read by every sex. `"est_all"`: one curve per sex, each penalized, with an `"own_mean"` `InitDevs_pen_center` pooled across sexes so the sexes share a single estimated level. This is how assessments with a common mean-log-initial and sex-specific initial deviations are written. Requires `n_sexes > 1` |
| `sigmaR_spec` | Recruitment variability: `"est_all"`, `"fix_early_est_late"`, `"est_shared_all"`, `"fix"` |
| `sigmaR_switch` | Year index separating the early and late $\sigma_R$ periods |
| `do_rec_bias_ramp` | Methot & Taylor bias adjustment (0 = off, 1 = on) |
| `bias_year`, `max_bias_ramp_fct` | Bias ramp breakpoints and maximum correction factor |
| `dont_est_recdev_last` | Number of terminal-year recruitment deviations to fix at zero |
| `init_age_devs_shared` | Integer vector of length `n_ages - 1` specifying age-sharing for `ln_InitDevs`. Positions with the same value share a single estimated parameter (e.g. `c(1:42, rep(42, 9))`). Required when `equil_init_age_strc = 3`; `NULL` (default) uses standard behavior |
| `use_rinit` | 0 = population initialized using `ln_global_R0` (default); 1 = separate `ln_rinit` used for initialization, with `ln_global_R0` governing only the recruitment relationship.|
| `Use_rinit_pen`, `rinit_pen_sd` | 0/1 and a standard deviation. With `use_rinit = 1` the two recruitment levels are linked only through the data, and the initial age structure is often the thinnest part of it. `1` places a normal on $\log(\text{rinit}/R_0)$, reported as `rinit_nLL`, saying how far the stock's equilibrium recruitment may have sat from the level the modeled years show. An equilibrium recruitment averages over several year classes rather than one, so its spread is narrower than $\sigma_R$. A reasonable choice is $\sigma_R/(1/M - 0.5)$, the recruitment variability shrunk by the number of year classes the equilibrium averages over |

A deviation shared across cells of the parameter array through the map (`est_shared_pop_r`, `est_shared_r`, `est_shared_s`, `init_age_devs_shared`) is one parameter and carries one penalty: each cell the penalty loop visits takes $1/n$ of it, $n$ the number of such cells holding the level. Under the bias ramp the initial age deviations are centered on the ramp read at each age's own birth year, deviation index $1 - a$ for the deviation on age $a$, which is how a ramp defined on calendar years treats the years before the first model year; `bias_year` can therefore be negative.

### Deviation penalty centering (`RecDevs_pen_center`, `InitDevs_pen_center`, `Fdev_pen_center`)

The recruitment, initial age, and (iid) fishing mortality deviation penalties can each be centered in one of two ways:

| String | Penalty mean | What it constrains |
|---|---|---|
| `"fixed"` | The asserted prior mean: zero, or the bias-corrected $-\sigma_R^2/2$ for recruitment under the bias ramp. Default | Both the level and the spread of the deviations |
| `"own_mean"` | The mean of the estimated deviations themselves | Only their spread; the level is left free (a sum of squares about the mean, matching assessments whose deviation vectors sum to zero) |

Recommendations. `"fixed"` is the statistically coherent choice when $\sigma_R$ is estimated or the deviations are treated as random effects: the penalty is then a genuine distributional assumption, and the bias ramp machinery depends on the mean being asserted. `"own_mean"` is primarily a *bridging* device: it reproduces assessments where the mean parameter (`ln_global_R0`, `ln_F_mean`) carries the level and the deviations carry only shape, so the level is not penalized twice. Two cautions: under `"own_mean"` the deviations' level must be pinned elsewhere (an $R_0$ prior, a fixed deviation, or informative data) or the likelihood is flat along it, and `RecDevs_pen_center = "own_mean"` cannot be combined with `do_rec_bias_ramp = 1` (the $-\sigma^2/2$ offset is meaningless once the mean is estimated rather than asserted; setup errors out).

### Recruitment level penalty

Separate from the deviation penalty, an optional penalty on the log recruitment *series itself*:

| Argument | Description |
|---|---|
| `Use_rec_level_pen` | 0 (default) / 1 toggle |
| `rec_level_pen_sigma` | Standard deviation of the penalty. A sum of squares with weight $w$ corresponds to $\sigma = 1/\sqrt{2w}$. Default 1 |
| `rec_level_pen_center` | `"own_mean"` (default; penalizes only the series' variability) or `"fixed"` (centers on zero) |
| `rec_level_pen_yrs` | Calendar years the penalty applies over; `NULL` (default) = all years |

Rationale. Under a stock-recruit relationship the deviations are residuals about the predicted curve, so a model that also wants the *realized* recruitment series to stay regular has nowhere else to say so. This penalty is that second, independent statement, and reproduces the recruitment regularity penalties several existing assessments carry (e.g., a penalty on log recruitment variability). Leave it off unless recruitment in data-poor years is wandering unreasonably; it is a tuning penalty, not a probability model, and it will shrink genuine recruitment variability if over-weighted (hence the `rec_level_pen_sigma` is left as data rather than an estimated parameter).

### Initial age deviations, tie between sexes

When the sexes carry their own initial age curves (`InitDevs_sex_spec = "est_all"`), a second statement can be made about them: not how variable each curve is (that is the initial-age penalty) but how far apart the two curves may sit,

$$-\ln L^{\text{InitSex}} = \sum_{s > 1}\sum_{i} -\ln \phi\left(\epsilon_{p,r,i,s}^{\text{Init}} - \epsilon_{p,r,i,1}^{\text{Init}};\ 0,\ \sigma_{\text{InitSex}}\right)$$

| Argument | Description |
|---|---|
| `Use_init_sex_pen` | 0 (default) / 1 toggle. Requires `n_sexes > 1` and `InitDevs_sex_spec = "est_all"`; under `"est_shared_s"` the difference is identically zero and setup refuses it |
| `init_sex_pen_sigma` | $\sigma_{\text{InitSex}}$. A sum of squares with weight $w$ corresponds to $\sigma = 1/\sqrt{2w}$. Default 1 |

It runs over the same ages the initial-age penalty covers and enters the objective unweighted (reported as `Init_Sex_nLL`). Its natural use is the one flatfish assessments put it to: male and female initial numbers are estimated separately, the first few years of compositions are all that informs them, and the tie keeps the two from drifting apart where those data are thin. Like the recruitment level penalty it is a tuning prior rather than a probability model, so its sigma is held as data.

### Stock-recruit curve as a penalty

Under `rec_model = "mean_rec"` a stock-recruit curve can be evaluated alongside the dynamics without ever advancing them, its residual penalized instead. Recruitment stays a mean with deviations, $R_y = R_0^{\text{mean}}\exp(\varepsilon_y)$, and the curve enters the objective only through the log residual $\xi^{\text{SR}}_y = \ln R_y - \ln \hat{R}_y$:

$$-\ln L^{\text{SR}} = \sum_{y \in \mathcal{Y}_{\text{SR}}} -\ln \phi\left(\xi^{\text{SR}}_y;\ 0,\ \sigma_{\text{SR}}\right)$$

| Argument | Description |
|---|---|
| `sr_penalty` | `"none"` (default), `"bh"`, or `"ricker"`, naming the curve $\hat{R}_y$ follows. Valid only with `rec_model = "mean_rec"`; under `"bh_rec"` or `"ricker_rec"` the curve already generates recruitment, which makes the deviation the residual by construction, so penalizing the residual as well would count it twice and setup rejects the combination |
| `sr_pen_sigma` | $\sigma_{\text{SR}}$, the standard deviation of the residual. Held as data rather than estimated. Default 1 |
| `sr_pen_yrs` | Calendar years $\mathcal{Y}_{\text{SR}}$ the penalty is summed over; `NULL` (default) is every year that has a lagged spawning biomass, i.e. all but the first `rec_lag`. Years outside it keep their recruitment deviation estimated and contribute nothing, which is how a restricted stock-recruit window is expressed. Naming a year with no lagged spawning biomass is an error rather than a silent fallback: the curve's prediction there is the equilibrium, so the residual would not be a stock-recruit residual |
| `sr_R0_spec` | Where the curve's scale $R_0^{\text{SR}}$ comes from (see below) |

| String | The curve's scale $R_0^{\text{SR}}$ |
|---|---|
| `"shared"` | `ln_global_R0`, i.e. $R_0^{\text{mean}}$ itself, which under mean recruitment is the mean recruitment level. One parameter and well posed, but the curve is anchored on mean recruitment rather than on an unfished level. Default |
| `"est"` | The curve's own estimated `ln_sr_R0`, identified by the curve fit alone. Reproduces templates carrying separate mean-recruitment and unfished-recruitment parameters, though nothing ties the two together, so the pair can slide |
| `"rinit"` | `ln_rinit`, the initial-equilibrium recruitment, so one parameter sets both the unfished age structure and the curve. Requires `use_rinit = 1` and errors otherwise. This is the usual ADMB arrangement |

Steepness is reachable in this mode: `h_spec`, `Use_h_prior`, and `h_prior` act on the penalty curve exactly as they do on a curve that generates recruitment.

What this is, and what it is not. It is a penalty on a derived quantity, a statement about how far realized recruitment sits from a curve, and not a data-generating process. The recruitment deviations remain free and keep their own penalty, so a weakly determined relationship informs the recruitment series rather than dictating it. The simulation side does not see it at all: `Setup_Sim_Rec()` and the operating model generate recruitment from the mean and its deviations under `"mean_rec"`, and `sr_penalty` enters the estimation model's objective alone. The curve's prediction is reported as `SR_pred` and its penalty as `SR_pen_nLL`; see `vignette("d_model_report")`. MSY is the maximum of equilibrium yield over a curve that drives the dynamics, and this curve does not, so `Get_Reference_Points()` refuses an MSY request on a fit carrying the penalty and directs it to the SPR variants.

### Recruitment penalty weighting (`Setup_Mod_Weighting()`)

| Argument | Description |
|---|---|
| `Wt_Rec` | Weight on the recruitment deviation penalty. Scalar (default 1), or an array `[n_pop × n_regions × n_est_rec_devs]` for per-deviation weighting; note the third dimension follows `ln_RecDevs` (moved by `dont_est_recdev_last` and `n_proj_yrs_devs`), not the number of years |
| `Wt_Init_Rec` | Weight on the initial age deviation penalty, `[n_pop × n_regions × (n_ages - 1)]` or scalar. `NULL` (default) inherits a scalar `Wt_Rec`; must be supplied explicitly when `Wt_Rec` is an array, since the two penalties are dimensioned differently |

A per-deviation weight of zero removes that deviation from the penalty while it remains estimated, which is how a stock-recruit relationship is fit over a chosen window of years while recruitment stays effectively free elsewhere. That is distinct from `dont_est_recdev_last`, which removes the deviations themselves (recruitment reverts to the deterministic prediction), and from mapping a deviation off (fixed *and* unpenalized; see [Which deviations are penalized]). Beware the $\sigma_R$ interaction: deviations excluded from the penalty no longer inform an estimated `ln_sigmaR`, so weight-based windows are safest with $\sigma_R$ fixed.

### Composition and index zero-guards (`Setup_Mod_Weighting()`)

| Argument | Description |
|---|---|
| `addtocomp` | Small constant guarding $\log(0)$ in composition likelihoods (default `1e-3`) |
| `comp_const_obs` | 0/1. Whether `addtocomp` is also added to the observed proportions used as multinomial weights. `1` (default) is the long-standing SPoRC behavior; `0` adds it only inside the logarithms, a convention several existing assessments use. The difference slightly reweights every bin, so set `0` when bridging such assessments and otherwise leave the default. The Dirichlet-multinomial also honors it, and there the constant is not neutral: every bin with no observed and no expected mass contributes $\log(\theta/(1+\theta))$, so compositions with many structurally empty bins (conditional age-at-length above all) bias $\theta$ upward under `1`; use `0` for those. `Setup_Mod_Weighting` warns if any conditional age-at-length fleet uses the Dirichlet-Multinomial while this is `1` |
| `addtofishidx`, `addtosrvidx` | Small constants guarding $\log(0)$ in the fishery/survey index likelihoods (default `1e-4` each) |
| `addtotag` | Small constant guarding zero tag releases/recoveries (default `1e-10`) |

`addtocomp`/`comp_const_obs`/`addtofishidx`/`addtosrvidx`/`addtotag` used to be arguments of `Setup_Mod_Biologicals()`. They still work there, forwarded with a message, but new code should pass them here instead.

### Steepness priors

When a stock-recruit curve is present, either because it generates recruitment (`rec_model = "bh_rec"` or `"ricker_rec"`) or because it enters as a penalty (`rec_model = "mean_rec"` with `sr_penalty`), steepness ($h$) can be penalized with a Beta distribution scaled to [0.2, 1] by default:

| Argument | Description |
|---|---|
| `Use_h_prior` | 0 = no prior, 1 = apply prior |
| `h_prior` | Data frame with columns `pop`, `region`, `mu`, `sd`, and optional `lb`, `ub` giving the beta's support (default 0.2 and 1) |
| `h_spec` | Estimation structure for steepness parameters |

The support matters, not just the mean and SD: a beta on $(0,1)$ is a genuinely different function of $h$ than one on $(0.2,1)$ (it carries $\log(h)$ where the rescaled form carries $\log(h - 0.2)$), and no choice of shape parameters reconciles the two. When bridging an assessment, match its prior's support via `lb`/`ub` rather than approximating with the default.

### R0 priors

A lognormal prior on $R_0$ can be applied per population:

| Argument | Description |
|---|---|
| `use_r0_prior` | 0 = no prior (default), 1 = apply lognormal prior on $\ln R_0$ |
| `r0_prior` | Data frame with columns `pop` (population index), `mu` (prior mean on natural scale), and `sd` (prior SD on log scale). Required when `use_r0_prior = 1` |

### Population straying

When `n_pop > 1`, a fraction of recruits produced by population $p$ can "stray" and recruit into regions associated with other populations.

| Argument | Description |
|---|---|
| `stray_rate_spec` | `"fix"` (default), `"est_all"`, etc. |
| `stray_rate_blocks` | Time blocks: `"none_Pop_<p>"` or `"Block_<b>_Year_<s>-<e>_Pop_<p>"` |
| `use_stray_rate_prior` | 0/1 toggle for Beta priors on stray rates |
| `stray_rate_prior` | Data frame with columns `pop`, `block`, `mu` (in (0,1)), `sd` |

### Spawning movement (single-season, multi-population)

When `n_pop > 1` and `n_seas == 1`, individuals cannot physically move to their natal region within the seasonal cycle, so a separate spawning-movement matrix (`sgl_seas_spawning_movement`) routes SSB back to natal regions for the SRR calculation. If not supplied, SPoRC defaults to 100% natal homing.

---

# Selectivity

Selectivity configuration is shared across fishery fleets (`Setup_Mod_Fishsel_and_Q()`), survey fleets (`Setup_Mod_Srvsel_and_Q()`), and retention curves (`ret_sel_model` within `Setup_Mod_Fishsel_and_Q()`). All three accept the same functional forms and time-varying structures.

### Functional forms

Specified as character strings following the pattern `"<form>_Fleet_<f>"` (constant across all time blocks) or `"<form>_Fleet_<f>_Block_<b>"` (block-specific). The available forms are:

| String | Functional form | Free parameters |
|---|---|---|
| `"logist1"` | Ascending logistic: $1 / (1 + \exp(-k(\text{bin} - a_{50})))$ | 2 ($\ln a_{50}$, $\ln k$) |
| `"logist2"` | Ascending logistic using $a_{50}$ and $a_{95}$: $1 / (1 + 19^{(a_{50} - \text{bin})/a_{95}})$ | 2 ($\ln a_{50}$, $\ln a_{95}$) |
| `"gamma"` | Dome-shaped gamma: $(b/a_\text{max})^{a_\text{max}/p} \exp((a_\text{max} - b)/p)$ | 2 ($\ln a_\text{max}$, $\ln \delta$) |
| `"exponential"` | Descending power: $1/\text{bin}^\beta$ | 1 ($\ln \beta$) |
| `"dbnrml"` | Double-normal with ascending and descending widths, plateau, and endpoint control | 6 |
| `"nonpar"` | Non-parametric: one logit-scale parameter per bin, transformed via $\text{logit}^{-1}$ | $n_\text{bins}$ |
| `"nonparlog"` | Non-parametric on the log scale, standardized so each year's selectivity averages to 1 over a window of bins (every bin by default; see [Normalization and length-based selectivity]) | $n_\text{bins}$ |
| `"nonparfree"` | Non-parametric on the log scale with no standardization at all: $\exp(\theta_\text{bin})$, so the values carry the height of the curve as well as its shape | $n_\text{bins}$ |
| `"asymplogist1"` | Logistic with asymptote $\alpha \in (0,1)$: $\alpha / (1 + \exp(-k(\text{bin} - a_{50})))$ | 3 ($\text{logit}(\alpha)$, $\ln a_{50}$, $\ln k$) |
| `"asymplogist2"` | Logistic with asymptote, $a_{50}/a_{95}$ parameterization | 3 ($\text{logit}(\alpha)$, $\ln a_{50}$, $\ln a_{95}$) |
| `"bicubic"` | Bicubic natural-cubic-spline surface over a bin-node $\times$ year-node grid | $n_\text{bin\_nodes} \times n_\text{yr\_nodes}$ (see below) |

The three non-parametric forms differ only in what happens after the transform, and that difference is what decides whether selectivity can carry a scale. Write $\theta_{a,b}$ for the estimated parameter at bin $a$ in block $b$, $b(y)$ for the block year $y$ falls in, and $s_{y,a}$ for the selectivity the model actually uses. Then

$$
\texttt{"nonpar"}: \quad
s_{y,a} = \frac{\operatorname{logit}^{-1}\!\left(\theta_{a,b(y)}\right)}
               {\dfrac{1}{n_\text{yrs} n_\text{bins}}\sum_{y'}\sum_{a'} \operatorname{logit}^{-1}\!\left(\theta_{a',b(y')}\right)}
$$

$$
\texttt{"nonparlog"}: \quad
s_{y,a} = \frac{\exp\!\left(\theta_{a,b(y)}\right)}
               {\dfrac{1}{|B|}\sum_{a' \in B} \exp\!\left(\theta_{a',b(y)}\right)}
$$

$$
\texttt{"nonparfree"}: \quad
s_{y,a} = \exp\!\left(\theta_{a,b(y)}\right)
$$

where $B$ is the set of bins named by `*_sel_norm_bins` (every bin by default). The inverse logit bounds each raw value below one before anything else happens; the exponential does not.

Read the denominators. `"nonpar"` divides by a single number computed over years *and* bins together, so it fixes the grand mean of the whole surface at one: one degree of freedom is removed from the surface as a whole. `"nonparlog"` divides by a number computed within each year, so it fixes every year's mean at one: one degree of freedom is removed *per year*, and only within-year contrasts survive. `"nonparfree"` divides by nothing, so every $\theta_{a,b}$ is a selectivity in its own right and the level is estimated along with the shape.

That last point is the practical one. A standardized form cannot hold a level, so the level has to live somewhere else, in catchability or in the fishing mortality mean. A free form can hold it, which is what makes it the right choice for an index fit age by age, where the age multiplier is the whole quantity being estimated. It is also what makes it possible to estimate a level twice by accident: if selectivity carries a level *and* the mean it multiplies is free, the two trade a common factor, the Hessian stops being positive definite, and the selectivity standard errors inflate while the fitted trajectory looks fine. Pin one bin, or one block's reference group, by leaving it out of `*_sel_nonpar_est_bins`.

Choosing among the three non-parametric forms. `"nonpar"` bounds every raw value below one (logistic transform) and mean-standardizes jointly over years and bins; `"nonparlog"` leaves the scale free and centers within each year. Under `"nonparlog"` only the within-year *differences* among parameters are identified, since the level is absorbed by catchability or fishing mortality, so pair it with the selectivity parameter centering penalty (below) or fix a bin group via `*_sel_nonpar_est_bins` to pin the scale. Prefer `"nonpar"` when you want selectivity interpretable as a proportion without further constraints.

`"nonparfree"` does not standardize at all, so its values hold the level as well as the shape. That is what an index fit age by age needs: a free catchability per age and a selectivity estimated at age are one quantity written two ways, so an at-age index puts the whole age multiplier in selectivity and leaves catchability alone. Group ages the way a catchability key would through `*_sel_nonpar_est_bins`, and leave an age out of the list entirely when the fleet never observes it, which maps it off rather than leaving it free and uninformed. Because nothing is canceled by standardization, `"nonparfree"` also accepts a `"scale"` sex offset, which `"nonpar"` and `"nonparlog"` refuse. Do not pair it with the centering penalty! 

### Bicubic spline selectivity (`"bicubic"`)

Unlike the other functional forms above, `"bicubic"` selectivity is specified with its own extended syntax:

```
"bicubic_Bin_<n_bin_nodes>_Yr_<n_yr_nodes>_Fleet_<f>[_Block_<b>][_SelStyr_<year>][_NSelBins_<n>]"
```

A smooth 2-dimensional selectivity-at-bin-and-year surface is built from a small grid of $n_\text{bin\_nodes} \times n_\text{yr\_nodes}$ freely estimated log-scale node values (`fish_fixed_sel_pars` / `srv_fixed_sel_pars`, flattened column-major into a `[yr_node × bin_node]` matrix). Two natural-cubic-spline weight matrices are precomputed once at setup:

- $\mathbf{W}^{\text{bin}}$ ($n_\text{bins} \times n_\text{bin\_nodes}$), mapping bin-node values onto every bin,
- $\mathbf{W}^{\text{yr}}$ ($n_\text{yrs} \times n_\text{yr\_nodes}$), mapping year-node values onto every year,

and combined via a two-pass tensor product (bin-node values spline-interpolated across bins for every year-node, then those curves spline-interpolated across years) to give the full log-selectivity surface, which is then exponentiated. Setting `n_yr_nodes = 1` collapses the surface to a time-invariant bin-only spline; combining `n_yr_nodes = 1` with `fish_sel_blocks`/`srv_sel_blocks` re-fits an independent bin-only spline within each block.

Two optional suffixes restrict the fitted region of the surface, edge-holding (flat-lining) outside it:

- `_SelStyr_<year>`: a calendar year within the block. Only years from `SelStyr` through the block's end are actually spline-fit across the year dimension; years within the block before `SelStyr` are held constant at the `SelStyr` year's fitted curve ("previous years are filled").
- `_NSelBins_<n>`: restricts the spline fit to the first `n` bins (ages or lengths, per `fish_selex_type`/`srv_selex_type`). Bins beyond `n` are held constant at the last fitted bin's value (a plateau), rather than continuing the spline extrapolation.

Both suffixes only change which years/bins are considered part of the *fitted* surface; they do not change the total number of estimated node parameters.

### Temporal variation

SPoRC provides two mutually exclusive mechanisms for time-varying selectivity within a fleet. You can use discrete blocks or continuous deviations, but not both on the same fleet.

Discrete blocks (`fish_sel_blocks` / `srv_sel_blocks` / `ret_sel_blocks`): defined as `"Block_<b>_Year_<s>-<e>_Fleet_<f>"` or `"Block_<b>_Year_<s>-terminal_Fleet_<f>"`. Blocks must be non-overlapping and collectively span all model years. Each block gets its own set of fixed-effect selectivity parameters (and can even use a different functional form).

Continuous deviations (`cont_tv_fish_sel` / `cont_tv_srv_sel` / `cont_tv_ret_sel`): specified as `"<type>_Fleet_<f>"`. For parametric forms (`"logist1"`, `"gamma"`, etc.), IID and random-walk deviations act multiplicatively on the transformed base parameters. For semi-parametric forms (`"3dmarg"`, `"3dcond"`, `"2dar1"`), deviations act multiplicatively on the selectivity curve at the bin level.

| String | Description |
|---|---|
| `"none"` | Time-invariant (default) |
| `"iid"` | Independent annual deviations on selectivity parameters |
| `"rw"` | Random walk on selectivity parameters |
| `"3dmarg"` | 3D Gaussian Markov random field, marginal variance parameterization |
| `"3dcond"` | 3D GMRF, conditional variance parameterization |
| `"2dar1"` | Separable 2D AR(1) over bin × year |

Ancillary controls for continuous time-variation include `fishsel_pe_pars_spec` (hyperparameter estimation), `fish_sel_devs_spec` (deviation estimation structure), `fishsel_devs_shared_bins` (bin grouping for shared deviations), and `corr_opt_semipar` (which correlation components to suppress in semi-parametric forms). Bin-sharing specs (`"est_shared_b"` variants) require the deviations to be indexed by bin: any GMRF/AR1 form, or `"iid"`/`"rw"` on a non-parametric fleet (`"nonpar"`/`"nonparlog"`, where each deviation already belongs to one bin); `"iid"`/`"rw"` on a parametric form indexes deviations by *parameter* and is rejected.

Three further controls tune the process-error penalty itself:

| Argument | Description |
|---|---|
| `fishsel_pe_wt` / `retsel_pe_wt` / `srvsel_pe_wt` | Per-fleet multiplier on the selectivity process-error likelihood (default 1). `0` removes the distributional penalty entirely while the deviations remain estimated; use it when deviations should float subject only to explicit smoothness penalties, as several existing assessments do. Anything other than 0 or 1 makes an estimated PE sigma reinterpretable, so prefer 0/1 unless deliberately down-weighting |
| `fishsel_rw_init_sigma` / `retsel_rw_init_sigma` / `srvsel_rw_init_sigma` | Standard deviation on the first year of an `"rw"` deviation series. Default 5 (first year effectively free). `NA` starts the walk at zero under the walk's own estimated sigma, making the first year as smooth as every later step; appropriate when the base parametric curve already describes the first year well |
| `*_sel_bin_dev_bins`, `cont_tv_*sel_bin_devs` | Bin-override deviations; see below |

### Bin-override selectivity deviations

Individual bins can be cut loose from the functional form entirely: bins named in `fish_sel_bin_dev_bins` / `ret_sel_bin_dev_bins` / `srv_sel_bin_dev_bins` (a list with one element per fleet, `NULL` for fleets with no overrides) take a freely estimated annual value $\exp(\epsilon_{y,b})$ in place of whatever the form produced, applied *after* every other transformation including standardization. The rest of the curve keeps its parametric shape.

| Argument | Description |
|---|---|
| `fish_sel_bin_dev_bins` etc. | Which bins each fleet overrides (e.g., `list(1, NULL)` frees bin 1 of fleet 1 only) |
| `cont_tv_fishsel_bin_devs` etc. | Process error on the override deviations, per fleet: `"none"`, `"iid"`, or `"rw"` (with its own estimated sigma per bin and `*sel_bin_devs_rw_init_sigma` for the first year) |

When to use. The canonical case is a gear whose curve is well described by a parametric form except for one bin governed by *availability* rather than the gear (e.g., age-1 availability to a trawl survey varying with year-class strength). Overriding that bin gives it free annual variation without abandoning the parametric form (or paying for a full semi-parametric surface) elsewhere. Give the override `"rw"` process error unless the bin genuinely jumps independently between years; with `"none"`, each year's value is informed only by that year's compositions and can be poorly determined in sparse years.

### Selectivity parameter centering penalty

| Argument | Description |
|---|---|
| `Use_fish_selex_penalty` / `Use_ret_selex_penalty` / `Use_srv_selex_penalty` | 0 (default) / 1 toggle |
| `fish_selex_penalty` etc. | Data frame with columns `region`, `fleet`, `block`, `sex`, `par` (a single index or list-column of integer vectors naming a set), `wt`. Each row penalizes $w\,[\log(\overline{\exp(\theta)})]^2$ over the named set |

This pushes the *average selectivity* of a parameter set toward one, pinning the scalar of a `"nonparlog"` curve that catchability or fishing mortality would otherwise absorb. It is a softer alternative to fixing a bin outright, and the standard companion to `"nonparlog"` selectivity (weights of 10 to 100 are typical starting points; the penalty only needs to break a ridge, not dominate the fit). Because the expression averages on the natural scale it is only meaningful for log-scale parameter sets, so do not apply it to `"nonpar"` (logit-scale) or to logit-scale asymptote parameters.

### Parameter sharing and fixing

The `fish_fixed_sel_pars_spec` / `srv_fixed_sel_pars_spec` argument controls how base selectivity parameters are estimated:

| String | Meaning |
|---|---|
| `"est_all"` | Fully region-, sex-, and fleet-specific |
| `"est_shared_r"` | Shared across regions |
| `"est_shared_s"` | Shared across sexes |
| `"est_shared_r_s"` | Shared across both regions and sexes |
| `"fix"` | Fixed at starting values |

### Selectivity plateau (`_NSelBins_<n>`)

Any functional form's model string accepts an optional `_NSelBins_<n>` suffix (e.g. `"logist1_Fleet_1_NSelBins_17"`, composable with `_Block_<b>` in either order): bins beyond `n` are held at bin `n`'s computed value rather than evaluated through the form. This is the plateau convention many existing assessments carry, where selectivity is the functional form up to a chosen bin and constant after it. The bicubic form already had this suffix and builds its plateau into its spline weights; for every other form the hold is applied after the form and its parameter deviations, before bin-level semi-parametric deviations and bin overrides. The plateau changes the model whenever the curve has not saturated by bin `n`, so treat it as part of the selectivity specification rather than a display convention.

### Sex offsets (`fish_sel_sex_offset` / `ret_sel_sex_offset` / `srv_sel_sex_offset`)

Between fully independent sexes (`"est_all"`) and identical ones (`"est_shared_s"`) sits a family of models where the sexes are *linked*: one sex's curve is the other's, shifted or scaled by a small number of estimated offsets. These per-fleet options express that directly for total fishery, retention, and survey selectivity alike (requires `n_sexes > 1`):

| String | Meaning |
|---|---|
| `"none"` | Each sex's stored parameters are its own. Default |
| `"par"` | The stored parameter slots of every sex beyond the first hold additive offsets on the first sex's transformed-scale parameters, $\theta_s = \theta_1 + \delta_s$. For log-scale parameters the second sex's natural value is the first sex's times $e^{\delta}$, e.g. $k_M = k_F e^{\delta_k}$, $a^{50}_M = a^{50}_F e^{\delta_{50}}$ |
| `"scale"` | Each sex keeps its own parameters, and the realized curve of every sex beyond the first is multiplied by a constant $e^{\gamma_s}$ (`ln_fishsel_sex_scale` / `ln_retsel_sex_scale` / `ln_srvsel_sex_scale`, estimated per region, block, and sex). The scaled curve may exceed one |
| `"par_scale"` | A par offset and a scale offset |
| `"apical"` | Double normal only. Each sex keeps its own parameters, and for every sex beyond the first the ascending and descending limbs are built up to $A_s = e^{\gamma_s}$ rather than to one, with the plateau sitting there too. The selectivity at the first and last bins stays where that sex's own parameters put it |
| `"par_apical"` | A par offset and an apical offset |

Under `"par"`, an offset fixed at zero reproduces sex-shared parameters, so the option nests `"est_shared_s"` while letting the link be estimated; no new parameters are involved, the existing sex slots simply change meaning. Under `"scale"`, the offset is a log-scale constant on the whole curve, which is how several existing assessments carry a male selectivity offset; because a constant multiplier is canceled by mean standardization, setup refuses `"scale"` for the non-parametric forms (`"nonpar"`, `"nonparlog"`) and the semi-parametric time-varying structures.

Choosing between `"scale"` and `"apical"`. A scale offset multiplies the finished curve, so selectivity at the first and last bins moves with it. An apical offset is the height the two limbs are built up to, so those bins stay where $p_5$ and $p_6$ put them and only the middle of the curve moves. Either can draw any single curve, so a plot will not separate them, but their derivatives differ and a fit responds to that. Use `"apical"` where the offset is meant as the sex's maximum selectivity. It requires the double normal; setup refuses it elsewhere and points to `"scale"`.

The anchor bin. By default the ascending limb is anchored at the first bin,
so its selectivity-at-first-bin parameter means what it says. When the
compositions begin above the population's first length bin, that parameter is
describing a bin the data never see. `fish_sel_dbnrml_startbin` and
`srv_sel_dbnrml_startbin` (integer vectors, one bin index per fleet) anchor the
limb at the first *data* bin instead, so the parameter is the selectivity at a
bin the data inform, and every bin below it takes
$(b/b_{\text{start}})^{2}$ times the selectivity there. Set it to the first bin
the compositions carry whenever those start above the population's first bin.

Unanchored limbs. `fish_sel_dbnrml_raw` and `srv_sel_dbnrml_raw` (0/1 matrices, one row per fleet, columns for the ascending and descending limbs) leave a limb as the raw Gaussian $\exp(-(x - \text{peak})^2 / \text{width})$ built up to the apical value, with no rescaling to hit an endpoint; the default anchors both limbs. The two forms differ by the Gaussian's value at the end bin, which is negligible when the peak sits many widths from that bin and not otherwise. Leave a limb raw when its end-bin parameter is not something the data can speak to, and anchored when the selectivity at the end bin is a quantity worth estimating.

Time-varying deviations are untouched by any of these options and still apply per sex to the effective parameters.

### Normalization and length-based selectivity

Selectivity can be normalized relative to a specific bin, to the maximum, or to the mean across a bin range. When `fit_lengths = 1`, selectivity operates on length bins and is mapped to age-space via a user-supplied size-age transition matrix (`SizeAgeTrans`).

The `"nonparlog"` form carries its own standardization: each year's curve is divided by the mean of $\exp(\theta)$ over a window of bins $\mathcal{B}$,

$$\text{sel}_b = \exp(\theta_b) \Big/ \frac{1}{|\mathcal{B}|}\sum_{b' \in \mathcal{B}} \exp(\theta_{b'})$$

and that window is an input:

| Argument | Description |
|---|---|
| `fish_sel_norm_bins` / `srv_sel_norm_bins` | The bins $\mathcal{B}$ each fleet standardizes over, supplied to `Setup_Mod_Fishsel_and_Q()` and `Setup_Mod_Srvsel_and_Q()` respectively. A list with one element per fleet naming that fleet's bins, or `NULL` (the default) for every bin; a fleet's element may also be `NULL` on its own. Stored as a 0/1 array `[n_bins × n_fleets]` and read only by `"nonparlog"` fleets |

Why the window matters. A gear whose catchability is defined against only part of the bin range standardizes over that part. Two windows differ by a constant multiplier on the whole curve, which catchability would absorb completely *if* $q$ were free. Under an informative prior on $q$ it is not free, and the mismatch is instead levered onto the selectivity parameters by the factor $\partial(\text{index nLL}) / \partial \log q$, which in the BSAI Atka mackerel bridge is $-14.055$ (see `vignette("ab_bsai_atka_mackerel_case_study")`). A $q$ prior is also stated against a particular standardization, so the window and the prior mean have to follow the same convention or the same prior statement lands on a different number. With $q$ free and unpenalized the choice makes no difference to the fit.

### Selectivity priors

Lognormal priors on selectivity parameters are toggled via `Use_fish_selex_prior` / `Use_srv_selex_prior`, with hyperparameters supplied in a data frame (`fish_selex_prior` / `srv_selex_prior`) containing `region`, `fleet`, `block`, `sex`, `par`, `mu`, and `sd`.

### Selectivity smoothness penalty weights

All selectivity smoothness/regularization penalty weights are configured in a single place, `Setup_Mod_Weighting()`, via the `fish_sel_pen_wts`, `ret_sel_pen_wts`, and `srv_sel_pen_wts` arguments (one list per selectivity surface). Each is a named list/vector; any name not supplied defaults to `0` (off). All six terms are evaluated directly on a fleet's *realized* selectivity-at-bin-at-year surface, so they apply to any selectivity functional form and any fleet, regardless of block/deviation structure.

| Name | Description |
|---|---|
| `smooth_dome` | Hinge penalty discouraging decreases across adjacent bins (dome-shape control) |
| `smooth_bin_curve` | Second-difference penalty across bins, normalized by the number of fitted bins |
| `smooth_bin_diff` | Unconditional first-difference penalty across bins (both increases *and* decreases contribute, unlike `smooth_dome`), normalized by the number of fitted bins |
| `smooth_yr_diff` | First-difference penalty across years, normalized by the number of fitted years |
| `smooth_yr_curve` | Second-difference penalty across years, normalized by the number of fitted years |
| `smooth_mean_center` | Penalizes the per-year mean of log-selectivity away from zero; resolves the scale indeterminacy of the bicubic surface (a uniform per-year shift in log-selectivity otherwise trades off exactly against that year's fishing mortality) |

Per-fleet specifications. A single named specification is shared by every fleet; alternatively, pass an *unnamed list* with one named specification per fleet (use `list()` or `NULL` for fleets with no penalties) so, e.g., two surveys can carry different smoothing.

Per-year weights, bin ranges, and normalization. Each weight may also be a vector with one value per model year (`0` skips that year), so a penalty can act only in years where selectivity changes, or with year-specific strength. A specification may additionally carry:

| Name | Description |
|---|---|
| `bin_range` | Length-two vector `c(first, last)` restricting the bins the penalties act over; or a named list giving each term its own range. Confine a shape penalty (dome, curvature) to the older ages where the curve should flatten without constraining the ascending limb |
| `normalize` | `TRUE` (default) divides bin-wise weights by the number of penalized bins and year-wise weights by the number of years; settable per term. Turn off when weights are calibrated as explicit variances |
| `yr_diff_ref` | Reference log-selectivity vector for `smooth_yr_diff`'s first penalized year, which otherwise has no predecessor and goes unpenalized. Anchors an otherwise free series to a known selectivity before the data begin |

A useful identity: with `normalize = FALSE`, a per-year `smooth_yr_diff` weight of $1/(2\sigma_y^2)$ is exactly the negative log-kernel of a random walk on the realized log-selectivity with year-specific standard deviation $\sigma_y$. This is how selectivity random walks with tabulated per-year sigmas are reproduced as penalties on the curve rather than as deviation parameters.

Recommendations. Prefer the distributional process-error forms (`cont_tv_*`) when you want an estimated, interpretable sigma; prefer these penalties when bridging assessments whose selectivity smoothing is a tuned penalty, or when regularizing a `"bicubic"`/`"nonparlog"` surface. `smooth_mean_center` (or the centering penalty above) should accompany any form with a free scale. Weights are on the scale of squared log-selectivity differences; start small (1 to 10), inspect the realized surfaces, and remember that normalized and unnormalized weights differ by a factor of $n_\text{bins}$ or $n_\text{yrs}$. 

---

# Catchability

Configured alongside selectivity in `Setup_Mod_Fishsel_and_Q()` / `Setup_Mod_Srvsel_and_Q()`.

### Estimation structure (`fish_q_spec` / `srv_q_spec`)

| String | Description |
|---|---|
| `"est_all"` | Free parameter for each region × fleet combination |
| `"est_shared_r"` | One parameter shared across regions (per fleet) |
| `"fix"` | Held at user-supplied value |

### Time blocks (`fish_q_blocks` / `srv_q_blocks`)

Same syntax as selectivity blocks: `"none_Fleet_<f>"` for a single block, or `"Block_<b>_Year_<s>-<e>_Fleet_<f>"` for structured change points.

### Analytic catchability (`srv_q_type` / `fish_q_type`)

Per fleet, catchability can be concentrated out of the likelihood rather than estimated. `srv_q_type` is set in `Setup_Mod_Srvsel_and_Q()` and `fish_q_type` in `Setup_Mod_Fishsel_and_Q()`; the two take the same strings and behave identically, applied to whichever index that fleet fits:

| String | Description |
|---|---|
| `"est"` | Estimate `ln_srv_q` / `ln_fish_q` (default) |
| `"arith"` | Solve analytically as the ratio of mean observed to mean predicted index |
| `"geo"` | Solve analytically on the log scale, $\hat{q} = \exp(\overline{\log \text{obs} - \log \text{pred}})$ |

Both analytic forms solve one $q$ per region and fleet from only the years with observations, ignore block structure, and automatically fix that fleet's catchability parameter regardless of `srv_q_spec` / `fish_q_spec`; they are incompatible with catchability covariates and priors on that fleet.

Recommendations. Use `"geo"` with a lognormal index likelihood: it is the exact maximum-likelihood solution under a shared SE, removes one ridge-prone parameter per fleet, and typically speeds and stabilizes optimization. `"arith"` exists to match assessments that use the arithmetic ratio. Stay with `"est"` whenever you need a $q$ prior, covariates, time blocks, or when the index's absolute scale is genuinely informative (e.g., a swept-area survey with a strong prior near 1, where concentrating $q$ out would discard that information).

### Priors

| Argument | Description |
|---|---|
| `Use_fish_q_prior` / `Use_srv_q_prior` | 0/1 toggle |
| `fish_q_prior` / `srv_q_prior` | Data frame with `region`, `fleet`, `block`, `mu` (natural scale), `sd` (log scale). Penalty: $\text{Normal}(\ln(\mu), \sigma)$ |

---

# Natural Mortality

Controlled via `Setup_Mod_Biologicals()`.

### Estimation vs. fixing (`M_spec`)

| String | Description |
|---|---|
| `"est_ln_M"` | Estimate $\ln M$ across the defined block structure |
| `"fix"` | Fix at values supplied via `Fixed_natmort` (array `[p × r × y × a × s]`) |

### Block structure

Natural mortality parameters can be shared or made specific along any combination of five axes. Each argument takes either `'constant'` (all levels pooled) or a list of integer vectors defining blocks:

| Argument | Axis | Example |
|---|---|---|
| `M_popblk_spec` | Population | `list(c(1,2), 3)` → pops 1-2 share $M$, pop 3 is separate |
| `M_regionblk_spec` | Region | `list(1:3, 4:5)` |
| `M_yearblk_spec` | Year | `list(1:20, 21:40)` |
| `M_ageblk_spec` | Age | `list(1:5, 6:30)` → young vs. old |
| `M_sexblk_spec` | Sex | `list(1, 2)` → sex-specific $M$ |

All block specifications are crossed to produce unique $M$ parameters. For instance, two age blocks × two sex blocks = four estimated $\ln M$ values (assuming everything else is `'constant'`).

### Priors

Lognormal priors on $M$ are activated via `Use_M_prior = 1`, supplying a data frame with columns `popblk`, `regionblk`, `yearblk`, `ageblk`, `sexblk`, `mu`, and `sd`.

---

# Movement

Configured via `Setup_Mod_Movement()` (EM) or `Setup_Sim_Movement()` (OM).

SPoRC implements two movement parameterizations, plus a fixed-matrix escape hatch.

### Movement type (`move_type`)

| Value | Description |
|---|---|
| 0 | Unstructured Markov. Transition probabilities estimated via multinomial logit (softmax), with region 1 as the implicit reference category. Supports blocking along population, age, year, season, and sex dimensions |
| 1 | CTMC. A continuous-time Markov chain builds an instantaneous rate matrix $Q = D + Z$ from diffusion ($D$, isotropic dispersal scaled by region area) and taxis ($Z$, directional preference). Transition probabilities are obtained by matrix exponentiation: $\mathbf{P} = \exp(Q)$ |

### Fixed movement (`use_fixed_movement = 1`)

Bypasses estimation entirely. Supply `Fixed_Movement` as an array `[p × r_\text{from} × r_\text{to} × y × \tau × a × s]`.

### Unstructured Markov block structure (`move_type = 0`)

Each dimension can be pooled (`'constant'`) or blocked:

| Argument | Description |
|---|---|
| `Movement_popblk_spec` | Population blocks |
| `Movement_ageblk_spec` | Age blocks (e.g., `list(1:5, 6:30)` for age-dependent movement) |
| `Movement_yearblk_spec` | Year blocks |
| `Movement_seasblk_spec` | Season blocks |
| `Movement_sexblk_spec` | Sex blocks |

### CTMC configuration (`move_type = 1`)

| Argument | Description |
|---|---|
| `ctmc_move_dat` | Data frame of covariates with required columns `pop`, `regions`, `years`, `seas`, `ages`, `sexes`, plus any variables referenced in the formulas. For projection years, covariate lookups are capped at the last historical year unless extended rows are supplied |
| `diffusion_formula` | R formula for diffusion (e.g., `~ 1`, `~ bs(years, df = 4)`) |
| `preference_formula` | R formula for taxis/preference |
| `adjacency_mat` | Square `[n_regions × n_regions]` binary connectivity matrix, with a **zero diagonal**: residency falls out of the generator and is not specified here. A fully connected matrix is `1 - diag(n_regions)`; note that `diag(1, n_regions)` is the identity, which connects nothing and is rejected |
| `area_r` | Numeric vector of region areas (scales diffusion rates) |
| `ctmc_diffusion_bounds` | 0/1: if `1`, shifts diffusion columns to guarantee all off-diagonal generator-matrix entries are non-negative |
| `ctmc_scale_by_seasdur` | 0/1 (default `1`): if `1`, treats the generator as an annual rate and exponentiates `Q * seasdur[s]` each season, so movement and mortality share time units. `0` results in `Q * 1` each season.
| `move_expm_nsub` | Integer (default `0`): how matrix exponentials of the generator are evaluated. `0` is exact (`Matrix::expm`); a power of two `n >= 1` substitutes `n` implicit backward Euler substeps. See below

### Movement and mortality sequencing (`move_timing`)

Controls how movement and mortality are ordered within a season. The three
options coincide exactly when total mortality is constant across regions, when
mortality is zero, and when movement is absent, so single-region models are
unaffected by the setting.

| Value | Sequencing |
|---|---|
| `0` (default) | Movement then mortality. Historical `SPoRC` behavior |
| `1` | Mortality then movement |
| `2` | Continuous: movement and mortality act simultaneously, via `expm(Q * seasdur - diag(Z))` |

`move_timing = 2` requires an estimated CTMC generator (`move_type = 1` with
`use_fixed_movement = 0`); a discrete movement matrix has no guaranteed real
generator, so it is rejected rather than approximated.

The setting is not confined to the projection step. Anything observed partway
through a season has to be evaluated consistently with it:

| Quantity | `0` | `1` | `2` |
|---|---|---|---|
| Spawning biomass | Post-movement location | Pre-movement location | Partially redistributed, `expm(A * t_spawn)` |
| Catch, discards | Baranov, post-movement | Baranov, pre-movement | Season-integrated abundance (spatial Baranov) |
| Fishery index | Post-movement `N` | Pre-movement `N` | Season-integrated abundance |
| Survey index | `N * exp(-t_srv * Z)` | `N * exp(-t_srv * Z)` | Snapshot, `expm(A * t_srv) N` |
| Tag recaptures | Baranov on tag cohort | Baranov on tag cohort | Season-integrated tag abundance |

where `A = t(Q) * seasdur - diag(Z)`. Equilibrium initialization (including the
matrix plus-group solution), per-recruit reference points, and projections all
use the same seasonal operator, so they inherit the setting as well. Note the
survey/fishery contrast: a survey is a snapshot at an instant within the season
and so uses a partial propagation, whereas catch and the fishery index
accumulate over the season and so use the integral of that propagation. Full
equations are in the *Description of Model Equations* vignette.

### Evaluating the matrix exponential (`move_expm_nsub`)

Both places `SPoRC` turns a generator into fractions go through one routine:
`Get_Movement`, converting `Q` into the `Movement` array, and every
`move_timing = 2` seasonal operator. `move_expm_nsub` selects how that
exponential is evaluated.

| Value | Scheme |
|---|---|
| `0` (default) | Exact, `Matrix::expm(A)` |
| `n >= 1`, a power of two | Implicit (backward Euler), `(I - A/n)^-n` |

`n = 1` is plain `solve(I - A)`. Larger `n` applies the same step over `n`
subintervals, reached by squaring the substep operator: each squaring doubles the
exponent, so `n` costs one linear solve plus `log2(n)` matrix products. That is why
`n` is restricted to powers of two, which squaring reaches exactly. The scheme is
first order either way, so the finer ladder a general integer would allow buys
nothing.

The reason to want this is the adjoint. Under reverse-mode AD the derivative of a
linear solve is another solve, whereas the derivative of a matrix exponential is
much more expensive than its forward evaluation. Benchmarking the exponential on
its own over 600 six-region strata, the gradient runs about 18x faster at `n = 1`,
4x at `n = 8` and 2.5x at `n = 64`, against a tape that is slightly more expensive
to build.

Inside a real fit the gain is far smaller, because the exponential is only a slice
of the gradient and binary powering's `log2(n)` matrix products claw back the rest
as `n` grows. On a three-region, one-season, 25-year estimation model the measured
whole-model speedups were 2.7x at `n = 1`, 1.3x at `n = 64` and 1.1x at `n = 512`
under `move_timing = 2`, and about 1.2x at any `n` under `move_timing = 0`. Since
the accurate settings are the slow ones, the option is worth adopting only where
the exponential is a large share of the work: cost grows with the cube of
`n_regions` and with the number of (pop, year, season, age, sex) strata, so measure
on the model you intend to fit rather than assuming the isolated figure.

What it costs is accuracy, and not by a small margin. The scheme is first order in
`1/n`, so it is a different discretization of the movement process rather than a
cheap route to the same numbers. On a six-region chain generator with `Z` between
0.25 and 0.42:

| `n` | max abs error in the operator | error in survival |
|---|---|---|
| 1 | 1.3e-1 | 4.6e-2 |
| 8 | 2.1e-2 | 7.0e-3 |
| 64 | 2.8e-3 | 9.0e-4 |
| 1024 | 1.8e-4 | 5.6e-5 |

At `n = 1` survival is `1 / (1 + Z)` where the exponential gives `exp(-Z)`. Since
substeps are nearly free, `n` in the hundreds is the sensible setting if the intent
is to approximate the exponential; small `n` is only defensible if backward Euler
is being adopted as the movement process itself.

Two structural properties survive at any `n`, which is what makes the option safe
to use rather than merely fast. `I - A/n` is a non-singular M-matrix, so its inverse
is non-negative and movement fractions cannot go negative; with no mortality the
result is exactly column-stochastic, so fractions still sum to one without
renormalization. And because the seasonal operator and the catch integral come from
the same Van Loan block, `1'T = 1' - z'Integral` holds algebraically, so catch plus
survivors equals the starting abundance to machine precision even where the operator
itself is a coarse approximation.

### Continuous movement deviations (`cont_vary_movement`)

Origin-destination deviations applied multiplicatively to off-diagonal rates (CTMC) or additively on the logit scale (unstructured).

| String | Deviation structure |
|---|---|
| `"none"` | No deviations (default) |
| `"iid_y"` | Year only |
| `"iid_a"` | Age only |
| `"iid_y_a"` | Year × age |
| `"iid_y_a_s"` | Year × age × sex |
| `"iid_y_seas_a_s"` | Year × season × age × sex |
| `"iid_p_y"` | Population × year |
| `"iid_p_a"` | Population × age |
| `"iid_p_y_a"` | Population × year × age |
| `"iid_p_y_a_s"` | Population × year × age × sex |
| `"iid_p_y_seas_a_s"` | Population × year × season × age × sex |

### Additional movement controls

| Argument | Description |
|---|---|
| `do_recruits_move` | 0 = age-1 fish do not move; 1 = recruits follow the movement matrix |
| `Use_Movement_Prior` | 0/1 toggle for Dirichlet movement priors (unstructured) |
| `Movement_prior` | Data frame with `pop`, `region_from`, `year`, `seas`, `age`, `sex`, `alpha` (Dirichlet concentration vector of length `n_regions`) |
| `Movement_cont_pe_pars_spec` | Estimation structure for process-error hyperparameters: `"none"`, `"fix"`, `"est_all"`, `"est_shared"` |

---

# Catch, Fishing Mortality, and Discards

Configured via `Setup_Mod_Catch_and_F()`.

### Catch conditioning

| Argument | Description |
|---|---|
| `ObsCatch` | Observed aggregate catch `[r × y × τ × f]` |
| `ObsCatch_pop` | Population-specific catch `[p × r × y × τ × f]` |
| `catch_units` | Per-fleet: 0 = abundance, 1 = biomass |
| `Use_F_pen` | 0/1 toggle for the fishing mortality deviation penalty (`Fmort_nLL`); see [Which deviations are penalized] |
| `sigmaC_spec` / `sigmaC_pop_spec` | Catch observation-error estimation |

### Fishing mortality process error

`ln_F_devs` (annual log-scale deviations about `ln_F_mean`) can follow one
of three process-error structures, set via `Fdev_model`:

| String | Description |
|---|---|
| `"iid"` | Independent annual deviations (default) |
| `"rw"` | Random walk |
| `"ar1"` | First-order autoregressive |

`Fdev_pen_center` (`"fixed"`, the default, or `"own_mean"`) sets where the iid deviation penalty is centered; see the deviation penalty centering discussion in the Recruitment section for the shared rationale. Under the mean-plus-deviations parameterization the level of $F$ is already carried by `ln_F_mean`, so `"own_mean"` penalizes only the deviations' spread and avoids constraining the level twice, matching assessments whose deviation vectors sum to zero. It leaves `ln_F_mean` and the deviations' level mutually unidentified unless one is fixed, which `ln_F_mean_spec = "fix"` does; setup warns when it is combined with an estimated mean in configurations where nothing else reads the mean. Note also that the default iid penalty with `"fixed"` centering acts as a hidden "biomass tracks catch" prior when catch trends strongly; `Fdev_model = "rw"` is usually the better remedy for that than re-centering.

`ln_F_mean_spec` (`"est"`, the default, or `"fix"`) chooses the fishing mortality parameterization. `"est"` is the mean-plus-deviations form. `"fix"` maps `ln_F_mean` off at its starting value (zero unless supplied), so the deviations are free annual log-F outright: $F = \exp(\epsilon)$. It must be paired with `Fdev_pen_center = "own_mean"`, `Fdev_model = "rw"`, or `Use_F_pen = 0` — an `"iid"` or `"ar1"` penalty centered on the fixed zero mean would shrink the deviations toward $F = 1$, and setup rejects that combination.

`sigmaF_spec` controls sharing/fixing of the process-error standard
deviation (`ln_sigmaF`) across region × season × fleet, using the same
`"est_all"` / `"est_shared_<dims>"` / `"fix"` convention described in
[Parameter sharing conventions]. `Fdev_rho_spec` analogously controls the
AR1 correlation parameter (`Fdev_rho`) and is only active when
`Fdev_model = "ar1"`; for `"iid"` or `"rw"`, `Fdev_rho` is unused and
mapped to `NA` regardless of what is supplied.

Catch-active years (`UseCatch == 1` or any `UseCatch_pop == 1`) do not
need to be contiguous under `"rw"` or `"ar1"`, a fishery may close for
several years and reopen later. The transition between two active years
separated by a gap of $d$ closed years is taken directly over the
elapsed gap (the same marginal transition obtained by estimating
deviations for the closed years and integrating them out, without
actually estimating them). See [Fishing Mortality Deviations] in
`vignette("c_model_equations")` for the exact equations, and
[Which deviations are penalized] for how the active sequence is determined.

### Discard and retention framework

SPoRC decomposes total fishing mortality at age into retained and dead-discard components:

$$F_{p,r,\tau,a,s} = \sum_f F_{r,\tau,f} \left[\underbrace{\text{sel}_{a,f} \cdot \text{ret}_{a,f}}_{\text{retained}} + \underbrace{\text{sel}_{a,f} \cdot (1 - \text{ret}_{a,f}) \cdot \text{dmr}_f}_{\text{dead discards}}\right]$$

| Argument | Description |
|---|---|
| `ret_sel_model` | Retention selectivity functional form (same options as `fish_sel_model`) |
| `ret_sel_blocks` | Retention selectivity time blocks |
| `cont_tv_ret_sel` | Continuous time-varying retention selectivity |
| `dmr_mean_spec` | Estimation structure for dead discard mortality rate means |
| `dmr_dev_spec` | Estimation structure for DMR deviations: `"fix"` (default) or `"est_all"` |
| `Use_dmr_pen` | 0/1 toggle for the DMR deviation penalty (`dmr_nLL`); must be `1` when `dmr_dev_spec = "est_all"` |
| `discard_units` | Per-fleet: 0 = abundance, 1 = biomass, 2 = abundance fraction, 3 = biomass fraction |
| `ObsDiscard` / `ObsDiscard_pop` | Observed aggregate and population-specific discards |
| `sigmaD_spec` / `sigmaD_pop_spec` | Discard observation-error estimation |

### Estimating discard mortality rates

`dmr_dev_spec = "est_all"` estimates an annual deviation in every fished
region × year × season × fleet cell, not only in cells carrying a discard
observation. This is deliberate. From the decomposition above, `dmr` scales
the dead-discard component of total mortality, so it enters $Z$ and therefore
propagates into predicted retained catch, indices, and compositions:

$$C_{a} = \frac{F^{\text{ret}}_{a}}{Z_{a}} N_{a} \left(1 - e^{-Z_{a}}\right)$$

A cell is therefore informative about `dmr` whenever it is fished and
retention is less than one, with or without discard data. Discard
observations are not the dividing line, `dmr` in fact cancels out of the
predicted discard itself, which divides the dead discards back through by it,
leaving only the same $Z$ dependence.

The cells excluded are true closures, matching the condition under which the
objective pins `dmr` to zero: no aggregate or population-specific catch is
fit, and the aggregate catch observation is a recorded zero rather than
missing (`NA`). A missing observation is treated as a fished year, on the
assumption that fishing continued and we simply lack a value to fit.

Note that a fleet with full retention (`ret_sel = 1`, the package default)
contributes no dead discards at all, so `dmr` drops out of the objective
entirely and its deviations have a flat gradient. `dmr_dev_spec` and
`dmr_mean_spec` are only meaningful once retention is actually modeled, see
`vignette("s_discard_retention")`.

### Which deviations are penalized

The `ln_F_devs`, `logit_dmr_devs` and `ln_RecDevs` penalties are evaluated on
exactly the deviations that are estimated, read off the map mirrors
`map_ln_F_devs`, `map_logit_dmr_devs` and `map_ln_RecDevs` in the data list
rather than recomputed from the catch indicators or the deviation index
ranges. `fit_model()` refreshes those mirrors from the map immediately
before building the model, so a deviation mapped off by hand after setup is
neither estimated nor penalized:

```{r, eval = FALSE}
# drop dmr deviations for a stretch of years, penalty included
mp <- array(input_list$map$logit_dmr_devs, dim = dim(input_list$par$logit_dmr_devs))
mp[, 10:20, , ] <- NA
input_list$map$logit_dmr_devs <- factor(mp)
```

Two consequences worth knowing. Mapping a deviation off pins it at whatever
sits in `$par`, which is `0` by default, so `dmr` falls back on
`plogis(logit_dmr_mean)`, but a starting value supplied through `...` is
kept, not reset to zero. And under `Fdev_model = "rw"` or `"ar1"`, a pinned
deviation is dropped from the active sequence, widening the gap $d$ between
the deviations either side of it rather than being treated as an actual year.

For `ln_RecDevs` this matters most when `ln_sigmaR` is estimated. A deviation
fixed at zero still sits at the penalty's mean, so were it penalized it would
add to the sum of squares' denominator without adding any spread, and
$\sigma_R$ would be pulled low by roughly $\sqrt{n_{est} / n_{total}}$. Holding
the first 80 of 255 quarterly deviations, for instance, biases $\sigma_R$ down
by about 17%.

With `ln_sigmaR` fixed the excluded terms are constants: the objective shifts
by a fixed amount and its gradient is unchanged everywhere, so the likelihood
surface and the location of any optimum are untouched, and `Rec_nLL` becomes
comparable across runs that hold different numbers of deviations. Note that
this is not the same as the *fitted values* being unchanged. `nlminb` tests
convergence on relative changes in the objective's value, so on a model that
stops short of convergence (a flat ridge, or a `"singular convergence"` or
`"false convergence"` message), shifting the objective by a constant can move
where the optimizer halts. If excluding these terms visibly changes your
estimates, the identifiability of the model is the thing to look at, not the
penalty.

---

# Biological Inputs

Supplied via `Setup_Mod_Biologicals()` (EM) or `Setup_Sim_Biologicals()` (OM).

| Input | Dimensions | Description |
|---|---|---|
| `WAA` | `[p × r × y × τ × a × s]` | Spawning weight-at-age (used for SSB) |
| `WAA_fish` | `[p × r × y × τ × a × s × f]` | Fishery-specific weight-at-age. Defaults to `WAA` if `NULL` |
| `WAA_srv` | `[p × r × y × τ × a × s × f]` | Survey-specific weight-at-age. Defaults to `WAA` if `NULL` |
| `MatAA` | `[p × r × y × τ × a × s]` | Maturity-at-age (proportions in [0, 1]) |
| `SizeAgeTrans` | `[p × r × y × τ × l × a × s]` | Size-age transition matrix, shared by every fleet unless `SizeAgeTrans_fish`/`SizeAgeTrans_srv` overrides it. Required when `fit_lengths = 1` |
| `SizeAgeTrans_fish`, `SizeAgeTrans_srv` | `NULL` or `[p × r × y × τ × l × a × s × f]` | Optional per-fleet size-age keys, `SizeAgeTrans` with a trailing fleet dimension appended. `NULL` (default) reads every fleet from the shared `SizeAgeTrans`. Only valid under `growth_model = "none"`; a growth model already derives one key per fleet at that fleet's own timing and rejects these to avoid two sources for the same quantity. The fixed-data counterpart of `WAA_fish`/`WAA_srv` overriding the shared `WAA`, and of `Setup_Sim_Biologicals`'s `SizeAgeTrans_fish_input`/`SizeAgeTrans_srv_input` on the operating-model side |
| `fit_lengths` | 0/1 | Toggle for fitting length compositions |
| `do_caal` | 0/1 | Toggle for reporting the joint arrays at length and age (`Fish_caal`, `Fish_caal_discard`, `Srv_caal`), dimensioned `[p × r × y × τ × l × a × s × f]` like `CAA` and `CAL`. Requires `fit_lengths = 1`, and is switched on by itself when conditional age-at-length data are supplied, since their expected rows are read from these arrays. Each array carries a length and an age dimension at once, so they are the largest objects the model builds and every cell lands on the AD tape; leave off unless they are wanted |
| `growth_model` | `"none"` / `"vb_schnute"` / `"richards"` | `"vb_schnute"` builds each fleet's size-age transition inside the model from von Bertalanffy parameters in Schnute's form (`L1` at `growth_A1`, `L2` at `growth_A2`, `K`, `CV1`, `CV2`): linear growth from `growth_L0` at age zero up to `L1`, CV interpolated on length between the references, the plus-group mean size adjusted for fish older than the accumulator age, and a binned normal key with the tails in the end bins, read at each fleet's own timing (`t_fish`, `t_srv`). `"richards"` is the same curve applied to lengths raised to a sixth estimated parameter `rho`, which nests the von Bertalanffy form at `rho = 1`. Requires `fit_lengths = 1` |
| `ln_growth_pars` (via `...`) | `[p × r × s × n]` | Starting values for `L1, L2, K, CV1, CV2` (then `rho` under `"richards"`), log scale, passed through `starting_values` as every other parameter's are. Defaults to the ends of the length bins with `K = 0.15` and CVs of `0.1`, so supply your own for any real model |
| `growth_spec` | `"est_all"` / `"est_shared_r"` / `"est_shared_s"` / `"est_shared_r_s"` / `"fix"` | How the growth parameters are shared across regions and sexes. `growth_fix` holds individual parameters at their starting values |
| `growth_A1`, `growth_A2` | scalars | Reference ages for `L1` and `L2`; `growth_A2 = "Linf"` makes `L2` the asymptote itself, with no second reference age to solve it from |
| `growth_len_lower` | `[l]` | Lower edges of the length bins; `lens` in `Setup_Mod_Dim` are midpoints |
| `growth_cv_type`, `growth_sd_type`, `growth_dist` | `"len"` / `"age"`; `"cv"` / `"sd"`; `"normal"` / `"lognormal"` | Whether the CV interpolates on mean length or on age between the reference ages; whether `CV1`, `CV2` scale the mean or are standard deviations; the distribution of length at age behind the key |
| `growth_plus_group` | `"mixture"` / `"curve"` | `"mixture"` (default) takes the plus group's mean length as the survivorship-weighted mixture of the ages it holds, numbers declining at an assumed 0.2 per year and length rising from the curve at the accumulator age to the asymptote; `"curve"` reads the curve at the accumulator age |
| `growth_tv_model` | `NULL` or per parameter | Time variation of the growth parameters. Either a vector of `"none"`/`"iid"`/`"rw"` in parameter order, or named by parameter (`L1`, `L2`, `K`, `CV1`, `CV2`, `rho`) with the rest constant. A varying parameter gains a deviation series `ln_growth_devs[p, r, y, k, s]` and a log sigma in the first stream of `growth_pe_pars`; its realized value by year is reported as `growth_pars_y` |
| `growth_tv_years` | `NULL`, a vector, or a list | Calendar years each varying parameter's deviations are active in. `NULL` uses every year, a vector applies to all of them, a list names them per parameter. Deviations outside the range are held at zero |
| `growth_tv_link` | `"log"` / `"logit"` | The scale a deviation enters on. `"log"` multiplies the parameter by `exp(dev)`; `"logit"` keeps it strictly inside `growth_par_bounds` however large the deviation, which every growth parameter is |
| `growth_par_bounds` | `[n × 2]` | Lower and upper bounds per growth parameter, natural scale; required under the logit link |
| `growth_tv_sigma_spec` | `"fix"` / `"est"` | Whether the process error standard deviations of the deviations are estimated. Their starting values are the first stream of `growth_pe_pars`, one slot per growth parameter, supplied through `starting_values` and defaulting to `log(0.1)`. `growth_rw_init_sigma` gives the first year of a random walk its own standard deviation |
| `growth_tv_spec` | `"est_all"` / `"est_shared_r"` / `"est_shared_s"` / `"est_shared_r_s"` | How the deviation series are shared across regions and sexes |
| `growth_tv_type` | `"curve"` / `"cohort"` | How the deviated parameters reach size at age. `"curve"` (default) reads each year's sizes off that year's own curve, so a fish carries no memory of the years it lived through. `"cohort"` carries size at age forward: each cohort grows by the increment the current year's parameters imply from the size it reached, ages still in the linear phase keep the length at `growth_A1` their birth year gave them, the first age past `growth_A1` sits on the current year's curve, and the plus group blends the cohort entering it with the fish already there by their numbers at age. The CV at age is then held at the first year's. Because the plus group reads abundance, growth under this option is evaluated inside the population dynamics year loop |
| `growth_semipar` | `"none"` / `"iid"` / `"rw"` / `"2dar1"` / `"3dmarg"` / `"3dcond"` | Semi-parametric growth: a year-by-age surface `ln_growth_semipar_devs[p, r, y, a, s]` multiplying the parametric mean length at age, so the curve stays the parametric part and the deviations hold departures from it. The structures are the same process errors the semi-parametric selectivity forms use; the correlated ones are what make the surface estimable, since the deviations are not identified year by year and age by age from length data alone. The spread follows the deviated mean, so under `growth_cv_type = "len"` a deviation also moves the fish along the CV ramp and under `"age"` the spread at age is untouched |
| `growth_semipar_spec` | `"fix"` / `"est"` | Whether the surface's process error parameters are estimated. Their starting values are the second stream of `growth_pe_pars`, in the same slots the selectivity forms use, supplied through `starting_values` and defaulting to a scale of `0.05` with correlations of `0.3` |
| `growth_semipar_ages`, `growth_semipar_years` | `NULL` or vectors | Ages and calendar years the surface is estimated over; outside them the deviations are held at zero, which is how a surface is restricted to the ages the length data actually inform |
| `LenBinMap` | `NULL` or `[l × l_\text{obs}]` | Maps the model's length bins onto the bins the compositions are recorded on, each row summing to one, for compositions on coarser bins than the population carries. Expected compositions are mapped through it inside the likelihood exactly as the ageing error matrix maps ages, and the observed arrays are then dimensioned by the observed bins |
| `AgeingError` | `NULL`, `[a × a_\text{obs}]` or `[y × a × a_\text{obs}]` | Maps the model's ages onto the ages the compositions are recorded on: genuine misclassification (a true age 7 read as 6, 7 or 8), a collapse onto coarser observed bins, or both. Each row sums to one, or to zero to drop a model age from the observations. The age-axis twin of `LenBinMap`, applied in the same position inside the likelihood and validated by the same `check_bin_map()` |
| `AgeingError_fish`, `AgeingError_srv` | `NULL`, `[a × a_\text{obs} × n_f]` or `[y × a × a_\text{obs} × n_f]` | Per-fleet ageing error, for when a fishery reading otoliths and a survey reading scales do not misclassify the same way. `NULL` (default) gives every fleet the shared `AgeingError`, which is what a model written before these existed does. Every fleet must land on the same observed age bins, since the observed composition arrays carry one age dimension shared across fleets. Wired through the operating model too, as `AgeingError_fish_input` / `AgeingError_srv_input` in `Setup_Sim_Biologicals` |
| `waa_model`, `wt_len_pars` | `"data"` / `"wt_len"`; 2 values or `[p × r × s × 2]` | `"data"` (default) reads weight at age from the `WAA*` arguments. `"wt_len"` derives `WAA` (at spawning time) and each fleet's `WAA_fish` and `WAA_srv` (at that fleet's `t_fish` or `t_srv`) from its key and $W = aL^b$ at the bin midpoints instead of reading them as data. `Get_Reference_Points` reads the derived `rep$WAA` when it is present, falling back to `data$WAA` otherwise; that fallback is a zero placeholder under `"wt_len"`, so pass `rep` from a fitted model rather than calling it on `data` alone |

Growth is not seasonal. The curve depends only on real elapsed time since `A1`, so splitting a year into more seasons changes nothing about size at age: a von Bertalanffy or Richards curve with a single `K` gives the same length whether the time between two points is one whole year or several sub-annual steps that sum to it. There is no option to give `K` (or any other growth parameter) a within-year seasonal pattern, e.g. a multiplier per season the way Stock Synthesis's `MGparm_seas_effects` does. Every fleet's timing (`t_fish`, `t_srv`, `t_spawn`) still reads its own point on that single annual curve, so within-year growth differences across fleets are entirely a function of *when* in the year they sample, not of the growth rate itself varying by season.

---

# Observation Model: Indices and Compositions

Indices and compositions are configured together for fishery fleets (`Setup_Mod_FishIdx_and_Comps()`) and survey fleets (`Setup_Mod_SrvIdx_and_Comps()`).

### Indices of abundance

| Argument | Description |
|---|---|
| `ObsFishIdx` / `ObsSrvIdx` | Observed index values |
| `ObsFishIdx_SE` / `ObsSrvIdx_SE` | Log-scale standard errors |
| `fish_idx_type` / `srv_idx_type` | `"abd"` abundance, `"biom"` biomass, `"none"`. Survey fleets also take `"recdev"`, an index of the recruitment deviations themselves |
| `ObsFishIdx_pop` / `ObsSrvIdx_pop` | Population-specific indices (separate likelihood contribution) |
| `ObsFishIdx_pop_SE` / `ObsSrvIdx_pop_SE` | Population-specific index SEs |

### Age-disaggregated observations

`SPoRC` fits catch as an aggregate with a composition, and an index the same way.
ICES age-structured assessments generally do neither: they fit catch at age and index at age
directly, every age its own observation with its own catchability and its own
variance.

The two are not interchangeable. Splitting an at-age observation into a total and
a composition is exact for Poisson and multinomial, where the total is a genuine
count sum. However, both families are lognormal, and the sum of lognormals is not
lognormal, so no composition likelihood recovers the at-age statement! 
A fleet uses one form or the other. Supplying both for the same fleet is an
error, since it is the same information stated twice.

| Argument | Stream |
|---|---|
| `ObsCatchAA` / `UseCatchAA` | Retained catch at age |
| `ObsDiscardAA` / `UseDiscardAA` | Discards at age |
| `ObsFishIdxAA` / `UseFishIdxAA` | Fishery index at age |
| `ObsSrvIdxAA` / `UseSrvIdxAA` | Survey index at age |

Each has a `_pop` counterpart with a leading population dimension. All eight
evaluate the same at-age likelihood through one function, differing only in which
array supplies the prediction.

#### How a stream is reported

Every at-age array is dimensioned `[n_regions, n_years, n_seas, n_ages, n_sexes,
n_fleets]`, the layout of the prediction arrays the likelihood reads, so the two
line up margin for margin. That is the storage, not the claim. A fleet says which
of regions and sexes it reports separately through a `Type`, using the same
vocabulary the compositions use:

| String | Regions | Sexes |
|---|---|---|
| `"agg"` | summed | summed |
| `"spltRaggS"` | separate | summed. Default |
| `"aggRspltS"` | summed | separate |
| `"spltRspltS"` | separate | separate |

`CatchAA_Type`, `DiscardAA_Type`, `FishIdxAA_Type` and `SrvIdxAA_Type` take one
setting for every fleet or one per fleet, and each has a `_pop` counterpart. A
margin the fleet sums over carries its observation in slot one of that margin,
and a use flag anywhere else on it is refused rather than quietly ignored, since
the prediction has already added those cells in.

The sex margin is required whatever the fleet reports. A stream summed over
sexes still carries its observation in the full array, in sex slot one, and an
array one dimension short is refused rather than promoted: the alternative is a
silent reinterpretation of what you passed.

#### Density and where the observation error comes from

An at-age stream has the same choices the aggregated index has, per fleet:

| Argument | Options |
|---|---|
| `*AA_LikeType` | `"lognormal"` (default) or `"normal"` |
| `*AA_sigma_form` | `"none"` (default), `"data"`, `"est_additive"`, `"est_quadrature"` |
| `Obs*AA_SE` | Reported standard errors, shaped like the observations |

`"none"` uses the estimated parameter alone, which is what a stream with no
reported errors means. `"data"` uses the reported standard errors alone and holds
the parameter fixed, since nothing reads it. The other two combine them, added or
in quadrature. Disaggregating a survey index by age therefore no longer costs you
its survey-design errors.

#### Key matrices

Parameters are coupled through integer key arrays `[n_ages, n_sexes, n_fleets]`,
in which equal entries share a parameter and `NA` excludes one. This is the key
matrix convention ICES assessments use for coupling variances and catchability,
with `NA` marking an excluded parameter, and the fleet last because that is where
`SPoRC` puts it. One structure covers every sharing pattern that would otherwise
need its own argument:

| Intent | Key column for a fleet and sex |
|---|---|
| One parameter per age | `1 2 3 4 5` |
| By age group | `1 1 2 2 2` |
| One for the fleet | `1 1 1 1 1` |
| Age not observed | `NA` |

The sex margin is required here too; a key coupling the sexes says so by
repeating its entries across them. An age and sex a fleet never observes carries
no parameter whatever the key says, which is what holds the unused sexes of a
sex-summed stream out.

`sigmaCAA_key`, `sigmaDAA_key`, `sigmaFishIdxAA_key` and `sigmaSrvIdxAA_key`
carry the observation error. Catchability at age is not keyed here: an index fit
age by age carries its age shape in selectivity, through the `"nonparfree"` form
(see Selectivity above), since a free catchability per age and a selectivity
estimated at age are the same statement twice. A parameter informed by fewer than two observations is refused, and
fewer than five warns: an observation error standard deviation with a single
observation drives the likelihood to negative infinity rather than failing
outright, and the optimizer reports convergence either way.

A season needing its own observation error is a separate fleet, the same way
anything else needing its own selectivity or catchability is.

#### Correlation across ages

Each stream is set where its data are set, following how the package separates
fishery from survey elsewhere, and the population-specific form carries its own
setting rather than borrowing the aggregated one:

| Argument | Set in |
|---|---|
| `AgeObsCorr_catch`, `AgeObsCorr_discard` (and `_pop`) | `Setup_Mod_Catch_and_F()` |
| `AgeObsCorr_fish_idx` (and `_pop`) | `Setup_Mod_FishIdx_and_Comps()` |
| `AgeObsCorr_srv_idx` (and `_pop`) | `Setup_Mod_SrvIdx_and_Comps()` |

| String | Meaning | Parameters per fleet |
|---|---|---|
| `"iid"` | Ages independent within a cell. Default | none |
| `"1dar1"` | AR(1) in age distance | 1 |
| `"us"` | Unstructured across ages | $n_a(n_a-1)/2$ |
| `"2dar1"` | Separable AR(1) over ages and years | 2 |

Each takes one setting for every fleet or one per fleet, so a model can leave one
fleet independent and correlate another. The three ICES `obsCorStruct` options are
`"iid"`, `"1dar1"` and `"us"`; `"2dar1"` adds the year dimension.

`"1dar1"` is a statement about age distance, not about position in the observed
vector. A fleet observing ages 2, 3, 5 and 6 has a gap, and lag one across that
gap is not lag one, so the covariance is built from the ages themselves. Where
the observed ages are consecutive the autoregressive recursion is used instead,
which is the same density more cheaply.

`"us"` places no shape on how ages covary, at the cost of a parameter per pair.
The parameters are unconstrained: they fill the strict lower triangle of a matrix
whose rows are then normalized to unit length, which makes it a Cholesky factor,
so the result is a valid correlation matrix for any values and any principal
submatrix is one too. A fleet asking for more correlations than it has observed
cells is refused, and fewer than three cells per parameter warns.

`"2dar1"` correlates over ages and years jointly, with the covariance the
Kronecker product of an AR(1) over each. It is defined on a complete grid, so a
fleet's observed ages and years must form one; a gap is refused at setup rather
than filled in. Use `"1dar1"` or `"us"` for a ragged stream. The year correlation
is `trans_rho_<stream>_year`.

How the correlations are shared is a spec string, the same one the rest of the
package uses for `sigmaF_spec` and `Fdev_rho_spec`, rather than a structure of
its own:

The correlations sit over region, sex and fleet, with a leading population
margin for the population-specific streams, so the abbreviations are `r`, `s`,
`f` and `p`:

| `rho_*_spec` | Meaning |
|---|---|
| `NULL` | `"est_shared_r_s"`, or `"est_shared_p_r_s"` for a `_pop` stream. Default |
| `"est_shared_r_s"` | One per fleet |
| `"est_shared_s"` | One per region and fleet |
| `"est_shared_r"` | One per sex and fleet |
| `"est_shared_r_s_f"` | A single correlation |
| `"est_all"` | One per region, sex and fleet |
| `"fix"` | Held at their starting values |

One spec governs the stream's across-age correlation, its across-year
correlation and its unstructured matrix together, so two fleets that share under
`"us"` share a whole matrix rather than one entry of it. A region, sex or
population a fleet never observes carries no correlation parameter whatever the
spec says, which is what holds the unused slots of a summed margin out: a stream
summed over regions gets one correlation, in region one, however free the spec.

A cell with a single observed age carries no correlation to describe and falls
back to independent. Be careful estimating an AR(1) on a fleet observing only two
or three ages: the correlation is barely identified and will sit on its bound.

A correlated cell contributes its whole density to the first age present, leaving
the remaining ages at zero, so the reported likelihood array still sums to the
cell's contribution but is no longer readable age by age.

#### Units

`catch_units` and `discard_units` apply to the at-age streams as they do to the
aggregated ones: `"abd"` reads the numbers at age, `"biom"` multiplies them by
the fleet's weight at age.

Worth knowing that for a lognormal stream with a weight at age supplied as data,
the two are the same statement. The residual is
$\log O_{a} - \log(\hat{C}_{a} W_{a}) = \log(O_{a}/W_{a}) - \log \hat{C}_{a}$,
and $W_{a}$ is a constant, so it cancels and the likelihood is identical. Catch
at age arrives in numbers, so `"abd"` is the natural choice.

`"biom"` becomes a different statement in two places. Under
`*AA_LikeType = "normal"` the residual is $O_{a} - \hat{C}_{a} W_{a}$ on the
arithmetic scale, so the weight rescales it against a fixed $\sigma_{a}$. And
where weight at age is derived from estimated growth rather than supplied,
$W_{a}$ carries parameters, so a biomass observation at age informs growth and a
numbers one does not.

#### Discards at age

Discards at age are the total discarded, not the dead fraction the population
dynamics track, which is the same statement the aggregated discard stream makes:
the prediction is $D_{a}/\text{dmr}$. They are reported in numbers or weight, so a
fleet fitting discards at age needs `discard_units` of `"abd"` or `"biom"`; the
fraction units are a property of the catch as a whole rather than of one age and
are refused.

#### Identifiability

Two things are worth knowing before fitting these.

An estimated observation error needs the thing predicting the mean to be more
constrained than the data. Fishing mortality with free annual deviations can
reproduce every catch at age exactly, at which point the residuals vanish and the
estimated standard deviation collapses towards zero, which the likelihood rewards
without limit. Effort-driven fishing mortality, or a blocked or penalized
structure, is what makes an estimated catch standard deviation meaningful.

Fishing mortality must be constrained in the same way for the at-age catch
likelihood itself to be identifiable. Four ages by forty years by two seasons of
free fishing mortality is roughly the number of observations, and even under a
separable structure a ridge opens in which each age's fishing mortality absorbs
its own abundance.

### Estimated index observation error (`sigmaFishIdx_spec` / `sigmaSrvIdx_spec`)

An index carries a standard error from its own survey design, and an assessment
may additionally estimate a component covering everything that design does not.
this is often called an extra standard deviation. By default `SPoRC` uses
the reported standard errors as they stand, which is `"fix"`.

One parameter is estimated per fleet, `ln_sigmaFishIdx` or `ln_sigmaSrvIdx`.
The aggregated and population-specific streams are separate observations with
separate error structures, so they carry separate parameters, in the same way
`ln_sigmaC` and `ln_sigmaC_pop` do. The population streams are configured
through `sigmaFishIdx_pop_spec` and `sigmaSrvIdx_pop_spec`.

| String | Total standard deviation |
|---|---|
| `"fix"` | The reported standard error, unchanged. Default |
| `"est_additive"` | Reported standard error plus the estimated component. The additive convention, and what the bridges here reproduce |
| `"est_quadrature"` | Reported standard error and estimated component added in quadrature, treating them as independent variances |
| `"est_replace"` | The estimated standard deviation alone, discarding the reported errors. What several ICES assessments do |

`sigmaFishIdx_map` and `sigmaSrvIdx_map`, with `_pop` variants, give the
estimation groups as an integer vector of length `n_fish_fleets` or
`n_srv_fleets`. Fleets sharing a value share a parameter and `NA` holds a fleet
at its starting value, which is how a fleet a reference assessment pinned at a
bound is carried without pretending it was estimated.

Two things are worth knowing before turning this on. An estimated component is
confounded with a likelihood weight, because a weight on a normal likelihood is
the same statement as dividing the variance by that weight, so the estimate will
absorb the weight; `Setup_Mod_Weighting()` warns when both are used. A fleet
with a multivariate normal index likelihood takes its scale from the supplied
covariance and ignores a standard deviation entirely, so requesting one there is
an error rather than a silently unidentified parameter.

The simulator draws index observations from whatever total the fitted model
used, so `simulation_self_test()` exercises the estimated component without
further configuration.

### Index error structure (`FishIdx_LikeType` / `SrvIdx_LikeType`)

Per fleet. A fleet's population-specific index stream follows the same choice
for `"lognormal"` and `"normal"`, but stays lognormal under `"mvn"`, whose
covariance describes the region-aggregated series only. The simulator
(`Setup_Sim_Fishing` / `Setup_Sim_Survey`, and `simulation_self_test`
automatically) draws index observations under the same structures, with an
`"mvn"` fleet drawn through a one-factor decomposition of its covariance:

| String | Description |
|---|---|
| `"lognormal"` | Default. SEs are log-scale standard deviations |
| `"normal"` | Normal on the arithmetic scale; SEs are arithmetic standard deviations |
| `"mvn"` | Multivariate normal on the arithmetic scale with a fixed covariance supplied via `FishIdx_Cov` / `SrvIdx_Cov` (one matrix per `"mvn"` fleet, square with one row per fitted observation, ordered as observations appear scanning the fleet's use flags in array order). Validated at setup for symmetry and positive definiteness |

Recommendations. `"lognormal"` is right for almost all abundance indices (positive, multiplicative errors). Use `"mvn"` when the index provider supplies a covariance across years (e.g., a model-based index from VAST or sdmTMB whose inter-annual correlations are real information a diagonal likelihood would double-count). Use `"normal"` only for series that can legitimately go near zero or negative, or when bridging an assessment that fits on the arithmetic scale. Two caveats: OSA residuals are only available for `"lognormal"` fleets, and the MVN density includes the $-\tfrac{n}{2}\log(2\pi)$ constant that some other implementations omit, so absolute likelihood values are not directly comparable across implementations even when fits match.

### Index timing and age restriction

| Argument | Description |
|---|---|
| `t_fish` | `[r × τ × f]`, fraction of the season elapsed when the fishery index is observed; numbers decay by $e^{-t Z}$ before the index is formed, mirroring `t_srv`. Default `0` (start of season, the historical behavior). Set `0.5` for a mid-season CPUE snapshot |
| `fish_idx_ages` / `srv_idx_ages` | Which ages contribute to each fleet's index total: a list (one vector of ages per fleet, `NULL` = all) or an `[n_ages × n_fleets]` 0/1 array. The restriction applies to the index *sum* only; selectivity, catch, and the fleet's compositions are untouched. Restricting a survey to one age turns it into an index of that age alone (e.g., an age-1 acoustic recruitment index), without needing a knife-edge selectivity that would corrupt the compositions |

### Selecting fish at length rather than at age

Fish of one age are spread over a range of lengths, and a length-based gear
takes the long ones more often than the short ones. Selecting at age first
averages that away: it gives the age one selectivity value and then treats every
fish of the age as equally catchable. The options below decide whether that
averaging happens, once for the compositions and once for the catch weight.

All of them require length-based selectivity, and the two weight options also
require `waa_model = "wt_len"` so the model knows what a fish of each length
weighs. `srv_waa_selected` only applies to an index in weight
(`srv_idx_type = "biom"`), since an index in numbers has no weight at age in it.

If selectivity is flat across the lengths an age covers, every option below is
the same as the default. The question to ask is how steep the selectivity curve
is over one age's length range.

| Argument | Description |
|---|---|
| `FishLenComps_sel` / `SrvLenComps_sel` | `"age"` (default) or `"length"`, per fleet. `"age"` selects the catch or index at age and spreads it over lengths afterwards, so the length composition within an age is the size-age key's own. `"length"` spreads the fish at each age over the key first and selects them length by length, $C_{l} = s(l)\sum_{a}P(l \mid a)\,N_{a}\!\left( 1 - e^{-Z_{a}} \right)F/Z_{a}$, so the long fish of an age are taken more often. The key is the fleet's own, at `t_fish` or `t_srv`. Use `"length"` when selectivity is length based and the length compositions are what inform it. The two are different expected compositions, not two roundings of the same one: in an EBS Pacific cod model the fishery length likelihood is 119.36 under one and 198.55 under the other |
| `fish_waa_selected` / `srv_waa_selected` | 0/1 per fishery or survey fleet. `1` makes the fleet's catch biomass, or a survey's index in weight, use the mean weight of the fish it takes at each age, $\sum_{l}P(l \mid a)s(l)w(l)/\sum_{l}P(l \mid a)s(l)$, instead of the population mean weight at that age. Use `1` when the gear selects strongly within an age, so that a caught fish is larger or smaller than an average fish of its age. In an EBS Pacific cod model the caught two-year-olds weigh 40 percent more than the average two-year-old, which moves the catch by 0.44 percent. With flat or age-based selectivity the two are the same |

### Composition bin restriction (`*_bins`)

Which *observed* bins a fleet's compositions are fitted over. Observed and expected compositions are both subset to the named bins and renormalized within them, so bins outside are left out of the likelihood rather than forced to be explained. Use for gears that never resolve part of the range (a fishery that never catches the youngest ages, whose structural zeros would otherwise carry information); leave alone when the zeros are informative sampling zeros.

Every composition stream takes one, and they all share the same format: a list with one element per fleet, each a vector of bin indices or `NULL` for all bins, or an `[n_obs_bins x n_fleets]` array of 0/1 weights. Indices are into *observed* bins, that is after any `AgeingError` or `LenBinMap`. Default `NULL` fits every bin.

| Stream | Fishery | Survey |
|---|---|---|
| Marginal age comps | `FishAgeComps_bins` | `SrvAgeComps_bins` |
| Marginal length comps | `FishLenComps_bins` | `SrvLenComps_bins` |
| Conditional age-at-length | `Fish_caal_bins` | `Srv_caal_bins` |
| Population-specific age comps | `FishAgeComps_pop_bins` | `SrvAgeComps_pop_bins` |
| Population-specific length comps | `FishLenComps_pop_bins` | `SrvLenComps_pop_bins` |
| Discard age comps | `FishAgeComps_discard_bins` | -- |
| Discard length comps | `FishLenComps_discard_bins` | -- |
| Population-specific discard age comps | `FishAgeComps_discard_pop_bins` | -- |
| Population-specific discard length comps | `FishLenComps_discard_pop_bins` | -- |

```r
# the acoustic survey resolves ages 2-15 only; every other fleet fits all ages
SrvAgeComps_bins = list(NULL, 2:15, NULL, NULL)
```

Every fleet must keep at least two bins. A composition fitted over one bin says nothing, since the proportion in a lone bin is one whatever the model predicts, and the one-step-ahead machinery degenerates around it: the logistic-normal families spend a bin as the additive log-ratio reference and have none left, and the discrete families mark their only bin as the multinomial's determined cell, leaving `get_osa()` nothing to keep. Setup refuses it and names the fleet.

The restriction applies whatever the composition type. For sex-joint comps (`spltRjntS`) the named bins are dropped from each sex's block of the `[bin x sex]` stack, so the sex ratio the joint comps carry becomes the ratio *within* the fitted bins, which is usually what you want when the excluded bins are unsampled rather than genuinely absent. Logistic-normal covariances are built across all observed bins and then cut down to the fitted ones, so a gap in the named bins still counts towards the AR1 lag between the bins either side of it. One-step-ahead residuals are packed and evaluated over the same restricted bins, so `get_osa()` and the likelihood always agree on what was fitted.

### Restricting bins versus remapping them

`*_bins` and the bin maps (`AgeingError`, `LenBinMap`) are easy to confuse and do different things.

| | What it does | When you need it |
|---|---|---|
| `AgeingError` / `LenBinMap` | **Remaps**. An `[n_model_bins x n_obs_bins]` matrix the expected composition is multiplied through, so it lands on the bins the data were recorded on. Mass is redistributed and, wherever a row sums to one, preserved | The data are on different bins from the model: 1 cm model bins fit to 5 cm length compositions, or true ages smeared across observed ages by reading error |
| `*_bins` | **Drops**. Names a subset of the observed bins; the rest are excluded from the likelihood and the remainder renormalized. Bin definitions do not change, mass in the dropped bins is discarded | The data are on the model's bins but part of the range is not informative: a gear that never ages its youngest fish, or length bins nothing is ever sampled from |

They compose, in that order: remap model bins onto observed bins, then restrict to the observed bins being fitted. Neither substitutes for the other. No subset of fifty 1 cm bins is ever ten 5 cm bins, so a restriction cannot do a map's job; and a map that zeroed a bin out would leave an observed bin with a structural-zero expectation the composition likelihood cannot fit, so a map cannot do a restriction's job. `check_bin_map()` enforces both halves of that: map rows must sum to one (or to zero, to drop a model bin from the observations entirely), and no observed bin may be left with nothing mapped into it.

Worked on six model bins carrying expected proportions `0.05, 0.10, 0.25, 0.30, 0.20, 0.10`:

| | Result | Bins | What happened to the first two bins' mass |
|---|---|---|---|
| `LenBinMap` collapsing pairs | `0.15, 0.55, 0.30` | 6 → 3 | Summed into the first observed bin. Still there |
| `*_bins = 3:6` | `0.29, 0.35, 0.24, 0.12` | 6 → 4 | Discarded. The remaining four renormalize to one |

Both shrink the vector, which is why they get confused, but only the map carries the first two bins' fish somewhere rather than throwing them away. If bins 1 and 2 are 10 cm fish your survey does catch but records in a coarser bin, you want the first row. If they are 10 cm fish your survey never sees, so their observed proportion is a sampling artifact rather than a measurement, you want the second. Choosing the map when you meant the restriction quietly asks the model to explain a bin that carries no information; choosing the restriction when you meant the map throws away fish that were really there.

A restriction leaves a visible trace at setup, so you can confirm it took:

```
SrvAgeComps_bins for fleet 2 is fitted over observed bins: 2, 3, 4, ..., 15
```

### Composition likelihood families

Age and length compositions each accept one of five likelihood families:

| Value | Likelihood |
|---|---|
| 0 | Multinomial |
| 1 | Dirichlet-multinomial (overdispersion parameter $\theta$ estimated per fleet) |
| 2 | Logistic-normal, independent bins |
| 3 | Logistic-normal with AR(1) correlation across bins |
| 4 | Logistic-normal with AR(1) bin correlation and constant cross-sex correlation |

These are set per data stream via `comp_fishage_like`, `comp_fishlen_like`, `comp_srvage_like`, `comp_srvlen_like`, and their `_pop` and `_discard` variants.

### Composition structure types

Each composition data stream has a "type" controlling how data are aggregated:

| Value | Structure |
|---|---|
| 0 | Aggregated across sexes and regions |
| 1 | Split by sex and region (no implicit sex-ratio information) |
| 2 | Joint across sexes, split by region (preserves sex-ratio information) |
| 999 | No data for this fleet/year |

### Data streams

SPoRC supports a full matrix of composition data streams, each independently configurable:

| Category | Aggregate | Population-specific |
|---|---|---|
| Fishery age | `comp_fishage_like` | `comp_fishage_pop_like` |
| Fishery length | `comp_fishlen_like` | `comp_fishlen_pop_like` |
| Fishery discard age | `comp_fishage_discard_like` | `comp_fishage_discard_pop_like` |
| Fishery discard length | `comp_fishlen_discard_like` | `comp_fishlen_discard_pop_like` |
| Survey age | `comp_srvage_like` | `comp_srvage_pop_like` |
| Survey length | `comp_srvlen_like` | `comp_srvlen_pop_like` |

Each stream carries its own ISS arrays, $\theta$ parameters (log-scale overdispersion), and correlation parameters.

### Conditional age-at-length

Supplied on `Setup_Mod_FishIdx_and_Comps()` (`ObsFish_caal`, `UseFish_caal`, `ISS_Fish_caal`, `Fish_caal_LikeType`, `Fish_caal_Type`) and `Setup_Mod_SrvIdx_and_Comps()` (the `Srv_` twins), with the weights `Wt_Fish_caal` and `Wt_Srv_caal` on `Setup_Mod_Weighting()`. A CAAL observation is the age composition of the fish aged from one length bin, so the arrays carry a length dimension ahead of the age one (`[r × y × τ × l × a × s × f]`), the use flag and input sample size are per length bin (`[r × y × τ × l × f]`, `[r × y × τ × l × s × f]`), and the sample size is the number aged from the bin, not the number measured. The expected row is the joint catch at length and age normalized across ages, which `do_caal = 1` builds and the data switch on by themselves.

| Argument | Options |
|---|---|
| `Fish_caal_LikeType`, `Srv_caal_LikeType` | `"none"`, `"Multinomial"`, `"Dirichlet-Multinomial"`. The logistic-normal families are not offered: a bin's age sample is small and mostly zeros |
| `Fish_caal_Type`, `Srv_caal_Type` | The same `"CompType_Year_x-y_Fleet_z"` strings as the marginal compositions; `spltRspltS` when each row holds one sex's otoliths |
| `ln_Fish_caal_theta`, `ln_Srv_caal_theta` | Dirichlet-multinomial overdispersion, one per region and sex shared across length bins |

Fit CAAL with the marginal length compositions and without the marginal age compositions of the same fleet, which would count the same fish twice. Use `comp_const_obs = 0` with the Dirichlet-multinomial, since every structurally empty bin otherwise contributes to $	heta$. One-step-ahead residuals come from `get_osa(comp_source = "Fish_caal")` or `"Srv_caal"`, with a `len` column for the bin each row was conditioned on.

---

# Tagging

Configured via `Setup_Mod_Tagging()` (EM) or `Setup_Sim_Tagging()` (OM).

SPoRC implements a conventional mark-recapture framework (Brownie-type likelihood) where tagged individuals follow the full seasonal population dynamics, movement, mortality, and fishing, after release.

| Feature | Description |
|---|---|
| Release structure | Tags released by population, region, season, age, and sex |
| Recapture structure | Fleet-specific recaptures with release-cohort × recapture-year likelihood |
| Tag shedding | Chronic tag-loss rate (estimated or fixed) |
| Initial tag mortality | Immediate post-release mortality parameter |
| Reporting rates | Fleet-specific tag-reporting rates (estimated or fixed) |
| Population/age/sex attribution | Tags can carry population, age, and sex information at release |

---

# Reference Points

Computed post-estimation via `Get_Reference_Points()`.

The function accepts a `type` argument for spatial structure and a `what` argument for the reference-point method. All methods project a single recruit through the full age, season, and spatial structure to compute SBPR and YPR.

### Available methods

| `type` | `what` | Description | Multi-pop? |
|---|---|---|---|
| `"single_region"` | `"SPR"` | $F_{\text{SPR}_x}$, no movement |, |
| `"single_region"` | `"BH_MSY"` | Beverton-Holt $F_\text{MSY}$, no movement |, |
| `"multi_region"` | `"independent_SPR"` | Per-region $F_{\text{SPR}_x}$ ignoring movement | ✓ |
| `"multi_region"` | `"independent_BH_MSY"` | Per-region $F_\text{MSY}$ ignoring movement | ✓ |
| `"multi_region"` | `"global_SPR"` | Single $F_{\text{SPR}_x}$ applied uniformly across regions, with movement | ✓ |
| `"multi_region"` | `"global_BH_MSY"` | Single $F_\text{MSY}$ with movement (single-pop only) |, |
| `"multi_region"` | `"local_BH_MSY"` | Region-specific $F_\text{MSY}$ values jointly maximizing total yield under movement. Uses Newton-Raphson to solve equilibrium recruitment by origin | ✓ |

### Controls

| Argument | Description |
|---|---|
| `SPR_x` | Target SPR fraction (e.g., 0.4 for $B_{40\%}$) |
| `n_avg_yrs` | Terminal years to average demographic rates (selectivity, $M$, WAA, maturity, movement) |
| `calc_rec_st_yr` | First year for computing mean historical recruitment |
| `rec_age` | Recruitment lag used to exclude most-recent years from the mean |
| `is_discard_fleet` | Integer vector (per fleet) flagging discard-only fleets to exclude from landed yield in MSY calculations |
| `local_bh_msy_newton_steps` | Newton iterations for the local $F_\text{MSY}$ solve (default 6) |

---

# Simulation and Closed-Loop MSE

The operating-model side uses a parallel set of `Setup_Sim_*` functions that mirror the estimation-model pipeline but populate a `sim_list` rather than `input_list`. Key simulation-specific functions:

| Function | Purpose |
|---|---|
| `Simulate_Pop_Static()` | Forward-project the OM without feedback (open-loop) |
| `condition_closed_loop_simulations()` | Closed-loop MSE: periodically re-fits the EM, applies an HCR, and updates $F$ |
| `simulation_data_to_SPoRC()` | Converts OM output (with observation error) to `input_list` format for the EM |
| `simulation_self_test()` | Simulation-estimation test: fits the EM back to OM-generated data |
| `get_closed_loop_reference_points()` | Computes reference points inside the MSE feedback loop |

Key closed-loop arguments in `condition_closed_loop_simulations()`:

| Argument | Description |
|---|---|
| `closed_loop_yrs` | Year range over which feedback is active |
| `assessment_period` | Frequency of EM re-fitting (e.g., every 2 years) |
| `use_true_values` | If `TRUE`, the HCR uses OM-truth instead of EM estimates (perfect-information benchmark) |

---

# Estimation and Optimization

`fit_model()` constructs the RTMB automatic-differentiation function, optimizes via `nlminb`, and refines with Newton steps.

| Argument | Default | Description |
|---|---|---|
| `data` |, | `input_list$data` |
| `parameters` |, | `input_list$par` |
| `mapping` |, | `input_list$map` |
| `random` | `NULL` | Parameter names to marginalize as random effects (Laplace approximation) |
| `newton_loops` | `3` | Post-convergence Newton steps ($\Delta\theta = -H^{-1}g$) to reduce residual gradients |
| `do_optim` | `TRUE` | `FALSE` returns the un-optimized `MakeADFun` object for debugging |
| `nlminb_control` | `list(iter.max = 1e5, eval.max = 1e5, rel.tol = 1e-15)` | Passed to `stats::nlminb` |

---

# Diagnostics

| Function | What it does |
|---|---|
| `do_retrospective()` | Sequentially peels terminal years and re-fits; reports Mohn's $\rho$ per quantity |
| `do_jitter()` | Refits from perturbed starting values to test convergence stability |
| `do_likelihood_profile()` | Profiles the likelihood surface over user-specified parameters |
| `do_francis_reweighting()` | Computes Francis TA1.8 weights for composition data |
| `run_francis()` | Iterative Francis reweighting loop (re-fits after each adjustment) |
| `get_osa()` | One-step-ahead residuals, compositions, conventional tagging, and Catch/Discard/FishIdx/SrvIdx indices (external post-hoc, or internal model-based via `do_internal_comp_osa`/`do_internal_conv_tag_osa`). See `vignette("u_osa_residuals")` |
| `do_runs_test()` | Runs test for serial correlation in residuals |
| `get_model_rep_from_mcmc()` | Extracts model report quantities across MCMC posterior draws (compatible with `adnuts` / `tmbstan`) |
| `marg_AIC()` | Marginal AIC for models with random effects |

---

# Plotting

| Function | Output |
|---|---|
| `plot_all_basic()` | Multi-panel diagnostic overview |
| `get_ts_plot()` | Time series: SSB, recruitment, $F$, catch, depletion |
| `get_idx_fits_plot()` | Index fits (observed vs. predicted) |
| `get_catch_fits_plot()` | Catch fits |
| `get_comp_prop()` | Composition fits (bubble plots / proportion plots) |
| `get_selex_plot()` | Selectivity-at-age or selectivity-at-length |
| `get_biological_plot()` | WAA, maturity, natural mortality |
| `get_nLL_plot()` | Likelihood component breakdown |
| `get_retrospective_plot()` | Retrospective trajectories |
| `get_data_fitted_plot()` | Data-availability timeline |
| `plot_resids()` | Residual diagnostics |
| `get_retrospective_relative_difference()` | Relative difference plots for retrospective analyses |

All plotting functions accept either a single fitted model or a list of models for side-by-side comparison.
