---
title: "Structuring Your Data for SPoRC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structuring Your Data for SPoRC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`vignette("a_model_dimensions")` already lists, field by field, every element of
`input_list$data`, including its name, its dimensions, and what it means. What it doesn't show
is the step most people actually get stuck on: turning raw
observations into an array with those exact dimensions, in that exact order. This
vignette walks through that step with a worked example. Consider `a_model_dimensions` as
the reference and this vignette as the "how to build" companion.

## The core convention

Every array that goes into `Setup_Mod_` is built from the same pool of axes, always
in the same relative order, with only the axes relevant to that particular quantity
included:

```
population < region < year < season < age or length < sex < fleet
```

For example, a fishery abundance/biomass index is `[region, year, season, fleet]`,
with no population, age, or sex axis. A fishery age composition is
`[region, year, season, age, sex, fleet]`. A population-specific weight-at-age is
`[population, region, year, season, age, sex]`. The exact axes for a given argument
are documented in that `Setup_Mod_` function's `?help` page and in
`vignette("a_model_dimensions")` but the relative pattern does not really change. Some potential issues that might arise include:

1. Array positions are positional, not literal. Index `i` along the age axis
   means `ages[i]` from your `Setup_Mod_Dim()` call, not "age `i`". For instance, if
   `ages <- 2:20`, the first slice along that axis is age 2. The same applies to
   `years` and `lens`. Indexing with a raw age/year value instead of its position
   in the corresponding vector is the single most common way to build an incompatible array.
2. Sex order is fixed. When `n_sexes == 2`, index 1 is always female and index 2
   is always male.
3. Every `Obs` array has a paired `Use` indicator array. `Use` is 0/1-valued
   and marks which region/year/season/fleet cells actually have data and should
   enter the likelihood. Getting `Obs` shaped correctly is necessary but not
   sufficient. In particular, a cell with `Obs == 0` because there's genuinely no data there
   still needs `Use == 0`, or the model will interpret it as an observed zero.

## Worked example: fishery catch and age composition

Suppose you have two raw data files: a time series of annual catch, and a table of
fishery age composition counts. First, define the model dimensions you're targeting
(normally the output of `Setup_Mod_Dim()`, shown here as plain vectors for clarity):

```{r, eval = F}
years <- 2016:2020
ages  <- 2:6
n_regions <- 1
n_seas <- 1
n_sexes <- 1
n_fish_fleets <- 1
```

### Catch

```{r, eval = F}
raw_catch <- data.frame(
  year = 2016:2020,
  fleet = 1,
  catch_mt = c(120, 135, 98, 150, 142)
)
raw_catch
```

`ObsCatch` needs dimensions `[n_regions, n_years, n_seas, n_fish_fleets]`. Build an
empty array of that shape, then place each raw observation using `match()` against
your dimension vectors rather than the raw year value itself:

```{r, eval = F}
ObsCatch <- array(0, dim = c(n_regions, length(years), n_seas, n_fish_fleets))

for (i in seq_len(nrow(raw_catch))) {
  yr_idx <- match(raw_catch$year[i], years)
  ObsCatch[1, yr_idx, 1, raw_catch$fleet[i]] <- raw_catch$catch_mt[i]
}

ObsCatch[1, , 1, 1]
```

`UseCatch` shares `ObsCatch`'s dimensions and flags which cells are real:

```{r, eval = F}
UseCatch <- array(0, dim = dim(ObsCatch))
UseCatch[ObsCatch > 0] <- 1
```

### Age composition

Composition data adds an age axis, so the target shape becomes
`[n_regions, n_years, n_seas, n_ages, n_sexes, n_fish_fleets]`:

```{r, eval = F}
raw_agecomp <- expand.grid(year = years, age = ages)
set.seed(1)
raw_agecomp$count <- rpois(nrow(raw_agecomp), lambda = 20)
head(raw_agecomp)
```

```{r, eval = F}
ObsFishAgeComps <- array(
  0,
  dim = c(n_regions, length(years), n_seas, length(ages), n_sexes, n_fish_fleets)
)

for (i in seq_len(nrow(raw_agecomp))) {
  yr_idx  <- match(raw_agecomp$year[i], years)
  age_idx <- match(raw_agecomp$age[i], ages)
  ObsFishAgeComps[1, yr_idx, 1, age_idx, 1, 1] <- raw_agecomp$count[i]
}

ObsFishAgeComps[1, 1, 1, , 1, 1] # full age vector for the first modeled year
```

Note that `UseFishAgeComps` only needs the non-age axes, `[n_regions, n_years, n_seas, n_fish_fleets]`, since it flags whole region/year/season/fleet cells, not individual ages:

```{r, eval = F}
UseFishAgeComps <- array(0, dim = c(n_regions, length(years), n_seas, n_fish_fleets))
UseFishAgeComps[1, , 1, 1] <- 1
```

## Structuring tagging data

Conventional tagging data is the most structurally involved input SPoRC accepts.
Releases are organized into cohorts, recaptures are tracked by time-at-liberty
rather than by calendar year, and whether a release's population, age, or sex is
even known depends on how the tagging program was run. The same reshape approach
from above still applies; it's just spread across more axes.

### Release cohorts

Every release event ("cohort") is one row of `conv_tag_release_indicator`, an
integer matrix with columns region, year, and season. As with the arrays above,
region and year are positions, not literal values: year `1` means `years[1]`, not
the calendar year 2016.

```{r, eval = F}
conv_tag_release_indicator <- matrix(
  c(1, 1, 1,   # region 1, year index 1, season 1
    2, 2, 1),  # region 2, year index 2, season 1
  ncol = 3, byrow = TRUE,
  dimnames = list(NULL, c("region", "year", "season"))
)
n_conv_tag_cohorts <- nrow(conv_tag_release_indicator)
```

`conv_tagged_fish`, dimensioned `[n_conv_tag_cohorts, n_pop, n_ages, n_sexes]`,
records how many fish were released per cohort. Which of the population, age, and
sex axes actually get used depends on `conv_fish_tag_attr`, which tells SPoRC which
of those dimensions were actually resolved at the point of tagging. A tagging
program that doesn't record age or sex at release (`conv_fish_tag_attr = "none"`)
pools every released fish for a cohort into index 1 of the population, age, and sex
axes, leaving the rest of those axes at zero:

```{r, eval = F}
conv_tagged_fish <- array(0, dim = c(n_conv_tag_cohorts, n_pop, length(ages), n_sexes))
conv_tagged_fish[1, 1, 1, 1] <- 50 # 50 fish released in cohort 1
conv_tagged_fish[2, 1, 1, 1] <- 30 # 30 fish released in cohort 2
```

If your tagging program does record age and sex at release, set `conv_fish_tag_attr`
to include `"a"`/`"s"` and fill each cohort's actual age/sex slot directly instead of
pooling into index 1. Region and fleet are always retained regardless of
`conv_fish_tag_attr`; only population, age, and sex are optional.

### Recaptures

`obs_conv_tag_fish_recap` is where the complexity concentrates:
`[conv_tag_max_liberty, n_seas, n_conv_tag_cohorts, n_pop, n_regions, n_ages, n_sexes, n_fish_fleets]`.
The first axis is time-at-liberty (periods since release): a fish recaptured one
period after release goes in position 1, two periods later in position 2, and so
on, up to `conv_tag_max_liberty`. Recaptures beyond that horizon simply have nowhere
to go and are ignored / not included in the array.

Given raw recapture records (which cohort, how many periods at liberty, which
region it was recaptured in, and how many fish):

```{r, eval = F}
conv_tag_max_liberty <- 3

raw_recaps <- data.frame(
  cohort       = c(1, 1, 1, 2, 2),
  liberty      = c(1, 2, 3, 1, 2),
  recap_region = c(1, 1, 2, 2, 2),
  count        = c(5, 3, 1, 4, 2)
)
```

the same reshape pattern builds the array, indexing directly by cohort and liberty
(both already positions) and by recapture region, with age, sex, and fleet pooled
into index 1 since this example's `conv_fish_tag_attr = "none"` and it only has one
fishery fleet:

```{r, eval = F}
obs_conv_tag_fish_recap <- array(
  0,
  dim = c(conv_tag_max_liberty, n_seas, n_conv_tag_cohorts, n_pop, n_regions, length(ages), n_sexes, n_fish_fleets)
)

for (i in seq_len(nrow(raw_recaps))) {
  obs_conv_tag_fish_recap[
    raw_recaps$liberty[i], 1, raw_recaps$cohort[i], 1,
    raw_recaps$recap_region[i], 1, 1, 1
  ] <- raw_recaps$count[i]
}
```

Note that a cohort released in one region can be recaptured in a different region;
region moves freely on the recapture side even when it's fixed on the release side,
and that movement between release and recapture region is exactly the signal
tagging data provides for estimating movement. If tagging data is primarily there to
inform movement, the recapture region axis is the one most worth double-checking.

## Common pitfalls

- Wrong axis order. Swapping, e.g., the year and age axes doesn't throw an error
  on its own. R will happily assign into a differently-shaped array as long as the
  total element count is compatible via recycling, silently producing garbage. Check
  `dim()` against the function's documented order immediately after building each
  array.
- Indexing by value instead of position. `ObsFishAgeComps[1, 2020, ...]` will
  either error (if `2020` is out of range for that axis) or silently write to the
  wrong slot. We would recommend using `match(2020, years)` instead.
- `Use` left out of sync with `Obs`. A zero in `Obs` is ambiguous between "we
  observed zero" and "we have no data here"; only `Use` disambiguates it. Building
  `Use` from `Obs > 0` (as above) is a reasonable default only when a true
  zero-count observation genuinely can't occur in your data; otherwise set it
  explicitly from which cells you actually sampled.
- Sex order reversed. Because `n_sexes == 2` always means female-then-male, a
  reversed sex axis won't error. It'll just fit a model where sex-specific
  parameters are swapped.
- `ages`/`lens`/`years` not sorted or not matching `Setup_Mod_Dim()` exactly.
  Every axis position is defined relative to these vectors, so they need to be the
  same vectors (order and values) you pass to `Setup_Mod_Dim()`.
- Recaptures placed past `conv_tag_max_liberty`. Anything recaptured further from
  release than `conv_tag_max_liberty` periods has nowhere to go in the array and is
  silently dropped from the likelihood. Set it generously relative to your tag
  return data rather than trimming it to a typical case.
- `conv_fish_tag_attr` inconsistent with how the array is actually filled. If
  `conv_fish_tag_attr = "none"` for a given axis but counts are still spread across
  more than index 1 of that axis, SPoRC has no way to know those extra indices are
  meaningful. Pool fully into index 1 for whichever axes aren't attended.

## Validating before you fit

Every `Setup_Mod_` function checks the dimensions of what you hand it internally and
fails fast with a specific message (e.g. "ObsFishAgeComps is not the correct
dimension. Should be n_regions, n_years, n_seas, ... ") rather than letting a
misshapen array propagate into a cryptic RTMB error. s

## See also

- `vignette("a_model_dimensions")`: the full field-by-field reference for every
  `input_list$data` element.
- `vignette("o_get_started")`: an end-to-end example that puts already-structured
  data through the full `Setup_Mod_` pipeline.
- `vignette("j_starting_mapping")`: the `"fix"`/`"est_shared"`/`"est_all"` spec
  pattern used for tagging mortality, shedding, and reporting rate parameters,
  among others.
- `vignette("architecture")`: how the `Setup_Mod_` pipeline itself is organized,
  for anyone extending it to a new data source.
